---
title: "ASESII 24A02 Project-Based Quiz 1"
subtitle: "Ridge, Lasso, and Regularization for Neural Features"
author: "Group 06: [names and student IDs]"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
params:
  group_number: 6
fontsize: 11pt
geometry: margin=1in
editor_options:
  markdown:
    wrap: 72
---

<<<<<<< Updated upstream
```{=html}
<!--
Copy this template into your group submission folder, rename it
GroupXX_ProjectQuiz1.Rmd, place references.bib beside it, and set the group
number parameter before rendering. Remove instructional comments from the
final report.
-->
```

# Authorship and contribution statement {-}

We confirm that every group member can explain the submitted analysis.
AI use is reported in `GroupXX_AI_Log.csv`, and the group checked every
AI-assisted item used in this report.

| Student               | ID       | Main contributions | Verification performed |
| --------------------- | -------- | ------------------ | ---------------------- |
| Le Dat Phuc An        | 24125052 | [Work]             | [Check]                |
| Pham Nguyen Minh Quan | 24125041 | [Work]             | [Check]                |
| Tran Canh Anh Tuan    | 24125048 | [Work]             | [Check]                |
| Vong Chi Van          | 24125049 | [Work]             | [Check]                |

# Abstract {-}

*Maximum 150 words: prediction problem, methods, main evidence, and
recommendation.*

# Reproducible setup {.unnumbered}

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("tidyverse", "glmnet", "broom", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)

group_number <- as.integer(params$group_number)
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number
)
```


## Shared helper functions {.unnumbered}

```{r helpers}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction)))
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  eigenvalues <- eigen(
    crossprod(x) / nrow(x),
    symmetric = TRUE,
    only.values = TRUE
  )$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
plot_model_result <- function(model, model_name) {
   model_result <- augment(model)
  ggplot(model_result, aes(x = .fitted, y = .resid)) +
    geom_point(color = "steelblue", size = 2, alpha = 0.8) +

    geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +

    theme_minimal() +

    labs(
      title = paste("Residual Plot Model:", model_name),
      subtitle = "Checking for non-linear patterns and constant variance",
      x = "Predicted lpsa (.fitted)",
      y = "Residuals (.resid)"
    )
}
```


# Problem 1. Prediction design and leakage control

## 1A. Target and predictors

```{r data-configuration, eval=FALSE}
config <- if (group_number %% 2L == 1L) {
  list(
    file = "fat.csv",
    response = "brozek",
    excluded = c("brozek", "siri", "density", "free")
  )
} else {
  list(
    file = "prostate.csv",
    response = "lpsa",
    excluded = "lpsa"
  )
}

data_raw <- read.csv(find_exam_data(config$file), check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}
```


*Define the prediction task and justify every exclusion.*

### 1. Response = $lpsa$

$lpsa$ measures the natural Logarithm of Prostate-Specific Antigen,
which is a continuous number

### 2. Candidate Predictors

|                      |                                  |                     |
|---------------------|-------------------------------|--------------------|
| Candidate Predictors | Full Name                        | Type                |
| lcavol               | Log Cancer Volume                | Continuous          |
| lweight              | Log Prostate Weight              | Continuous          |
| age                  | Age                              | Continuous          |
| lbph                 | Log Benign Prostatic Hyperplasia | Continuous          |
| svi                  | Seminal Vesicle Invasion         | Binary (0 / 1)      |
| lcp                  | Log Capsular Penetration         | Continuous          |
| gleason              | Gleason Score                    | Ordered Categorical |
| pgg45                | Percentage Gleason Grade 4/5     | Continuous (0–100)  |

### 3. Unit of Observation

Each row of the dataset contains the information of a radical
prostactomy treated patient which comprises of the target and the
predictors values.

### 4. Study Population

The dataset contains 97 patients aged 41 to 79 with clinically localized
adenocarcinoma who went through radical prostactomy at Stanford
University Center by Dr. Thomas A. Stamey and his surgical team during
the late 1980s.

### 5. Prediction Objective

To build a regression model that predicts the log prostate-specific
antigen level (lpsa) using the clinical variables (lcavol, lweight, age,
lbph, svi, lcp, gleason, and pgg45).

### 6. Dataset Source

<https://sci-hub.sg/storage/2024/6429/4cbb75646440af614abced23a70ba059/stamey1989.pdf>

### 7. Required Exclusions

The response variable, `lpsa`, was excluded from the predictor matrix because it is the outcome to be predicted. The remaining eight clinical variables were retained as predictors. Since the dataset contains no variables derived from `lpsa`, no additional exclusions were required. This prevents target leakage and ensures a fair comparison among the regression models.

## 1B. Split and train-only preprocessing

```{r split-and-scale, eval=FALSE}
split <- split_rows(nrow(analysis_data), seed = seeds$split)
train_id <- split$train
test_id <- split$test

x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_prep <- fit_scaler(x_raw[train_id, , drop = FALSE])
x_train <- apply_scaler(x_raw[train_id, , drop = FALSE], x_prep)
x_test <- apply_scaler(x_raw[test_id, , drop = FALSE], x_prep)
y_train <- y_raw[train_id]
y_test <- y_raw[test_id]

foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
```


*Report seeds, dimensions, scaling, folds, and leakage safeguards.*

### 1. Split seed

-   Seed `split` for 80/20 split: 240206
-   Seed `cv` for K-fold split: 240306

### 2. Train/Test Split

The dataset of 97 observations was partitioned into:

-   **Training Set:** 77 observations
-   **Test Set:** 20 observations

### 3. Feature Scaling Rule

Continuous predictor variables were centered and scaled using the helper functions `fit_scaler()` and `apply_scaler()`. The scaling parameters were estimated **only from the training data** in `fit_scaler()` and then applied unchanged to both the training and test sets by `apply_scaler()`.

Specifically, each predictor was transformed as

$$
x_{\text{scaled}}
=
\frac{x-\bar{x}_{\text{train}}}
{\sqrt{\frac{1}{n_{\text{train}}}\sum_{i=1}^{n_{\text{train}}}(x_i-\bar{x}_{\text{train}})^2}},
$$

where \(\bar{x}_{\text{train}}\) is the mean of the predictor in the training set, and the denominator is the population standard deviation computed from the training observations. Using the same scaling parameters for the test set prevents information leakage from the holdout data.

### 4. Five-Fold Assignment Rule

A single, immutable five-fold cross-validation assignment vector
(`foldid`) was generated on the 77 training observations using the
designated `cv` random seed. To ensure fair and controlled comparisons,
this exact same fold assignment was reused across all models evaluated:

-   Ridge Regression
-   Lasso Regression
-   Elastic Net
-   All subsequent hyperparameter grid evaluations

### 5. Missing-Value Plan

There are no missing values in the dataset

## 1C. Training-data exploration

```{r training-eda, eval=FALSE}
# Use analysis_data[train_id, ] only. Add focused plots and summaries.

## Response distribution

summary(y_train)
sd(y_train)

hist(y_train)
qqnorm(y_train)
qqline(y_train)

## Relationship exploration

cor_matrix <- cor(analysis_data[train_id, ])

heatmap_df <- expand.grid(
  Var1 = rownames(cor_matrix),
  Var2 = colnames(cor_matrix)
)

heatmap_df$value <- c(cor_matrix)

ggplot(heatmap_df, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 3) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  coord_fixed() +
  theme_minimal()

## One possible anomaly or influential observation

boxplot(analysis_data[train_id, ], 
        las = 2, 
        main = "Predictor Distributions (All Labels Visible)",
        col = "lightgray", 
        outchar = TRUE)

## One pair or group of predictors for which multicollinearity may matter

# (See correlation heatmap.)
```


*Interpret multicollinearity, one anomaly, and the question
regularization will address.*

The exploratory analysis in this section uses only the training
observations to avoid information leakage from the holdout set.

### Response distribution

The response variable (`lpsa`) is approximately symmetric, with the mean
close to the median. Most observations are concentrated between 1.5 and
3.5, while only a few observations appear in the tails. The histogram
and Q–Q plot indicate that the Gaussian assumption is broadly
reasonable, although slight departures from the reference line are
present in the upper tail.

### One correlation table or focused relationship plot

The correlation heatmap shows that **lcavol** has the strongest positive
correlation with the response (`r = 0.75`), indicating that tumor volume
is the most informative individual predictor of PSA level. Moderate
positive correlations are also observed for **lcp** (`r = 0.59`) and
**svi** (`r = 0.58`), whereas `age`, `lbph`, and `lweight` exhibit
relatively weak marginal relationships with the response.

### One possible anomaly or influential observation

The boxplots identify several potential outliers, particularly for `age`
and `pgg45`, with `pgg45` exhibiting the greatest variability among the
predictors. These observations may have a greater influence on the
fitted regression model than typical observations. However, because
there is no evidence that they arise from data entry errors, they are
retained and their influence will be assessed using formal regression
diagnostics in the OLS analysis.

### One pair or group of predictors for which multicollinearity may matter

The strongest correlations among the predictors are observed between
`gleason` and `pgg45` (`r = 0.76`), `lcavol` and `lcp` (`r = 0.73`),
`lcp` and `svi` (`r = 0.67`), and `lcavol` and `svi` (`r = 0.59`). These
predictor pairs appear to contain overlapping information, suggesting
that multicollinearity may be present. Consequently, ordinary least
squares coefficient estimates may become unstable and exhibit inflated
variance, motivating the use of ridge and lasso regression to improve
coefficient stability and variable selection.

### Question for regularized models

Based on the exploratory analysis, the following question will guide the
remaining analysis:

> **Can ridge regression reduce the instability caused by
> multicollinearity, and can lasso identify a smaller subset of
> predictors while maintaining comparable predictive performance?**

# Problem 2. OLS, ridge, and lasso

## 2A. OLS baseline

```{r ols, eval=FALSE}
x_train_df <- as_tibble(x_train)
x_train_df$lpsa <- y_train
m1 <- lm(lpsa  ~ .,data = x_train_df)
model_coefs <- tidy(m1, conf.int = TRUE )
print(model_coefs)

train_preds <- predict(m1, newdata =  x_train_df)
train_mse <- mean((x_train_df$lpsa - train_preds)^2)
train_rmse <-sqrt(train_mse)
cat("\n--- Training Error ---\n")
cat("Training MSE :", round(train_mse, 5), "\n")
cat("Training RMSE:", round(train_rmse, 5), "\n")
cv_mses <- numeric(5)
for(i in 1:5)
{
  fold_train <- x_train_df[foldid != i, ]
  fold_val <- x_train_df[foldid == i, ]
  model <-lm(lpsa  ~ .,data = fold_train)
  fold_preds <- predict(model, newdata =  fold_val)
  cv_mses[i] <-  mean((fold_val$lpsa - fold_preds)^2)
  log_message <- sprintf("Fold %d completed | Validation MSE: %.4f\n", i,cv_mses[i])
  cat(log_message)
  print(tidy(model, conf.int = TRUE ))
  print(plot_model_result(model,paste('fold',i)))
}
cv_mse_mean <- mean(cv_mses)
cv_rmse_mean <- mean(sqrt(cv_mses))
cat("\n--- CV Error ---\n")
cat("CV MSE :", round(cv_mse_mean, 5), "\n")
cat("CV RMSE:", round(cv_rmse_mean, 5), "\n")
print(plot_model_result(m1,'model'))

# Fit OLS on x_train and y_train. Do not access y_test here.
```


## 2B. Ridge

```{r ridge, eval=FALSE}
model_ridge <- fit_cv_glmnet(x_train,y_train,alpha= 0, foldid = foldid)
plot(model_ridge)
cat(sprintf("lambda.min: %.5f\n", model_ridge$lambda.min))
cat(sprintf("lambda.1se: %.5f\n\n", model_ridge$lambda.1se))

min_id <- which(model_ridge$lambda == model_ridge$lambda.min)
se1_id <- which(model_ridge$lambda == model_ridge$lambda.1se)
cat(sprintf("cvm.min: %.5f\n", model_ridge$cvm[min_id]))
cat(sprintf("cvm.1se: %.5f\n\n", model_ridge$cvm[se1_id]))

plot(model_ridge$glmnet.fit, xvar ='lambda', label = TRUE )#note the plot automatically turn into the plot.glmnet function
title("Ridge Coefficient Paths", line = 2.5)
# Use fit_cv_glmnet(..., alpha = 0, foldid = foldid).
# Report lambda.min, lambda.1se, paths, coefficients, and CV evidence.
```


## 2C. Lasso

```{r lasso, eval=FALSE}
model_lasso <- fit_cv_glmnet(x_train,y_train,alpha= 1, foldid = foldid)
plot(model_ridge)
cat(sprintf("lambda.min: %.5f\n", model_lasso$lambda.min))
cat(sprintf("lambda.1se: %.5f\n\n", model_lasso$lambda.1se))

cat("--- Nonzero Predictors (lambda.min) ---\n")
coef_min <- coef(model_ridge, s ='lambda.min')
print(c_min[coef_min[,1]!= 0])# mask which is 0 on the coefficient column
cat("\n--- Nonzero Predictors (lambda.1se) ---\n")
coef_1se <- coef(model_ridge, s ='lambda.1se')
print(c_1se[coef_1se[,1]!=0])

min_id <- which(model_lasso$lambda == model_lasso$lambda.min)
se1_id <- which(model_lasso$lambda == model_lasso$lambda.1se)
cat(sprintf("cvm.min: %.5f\n", model_lasso$cvm[min_id]))
cat(sprintf("cvm.1se: %.5f\n\n", model_lasso$cvm[se1_id]))

plot(model_lasso$glmnet.fit, xvar ='lambda', label = TRUE )
title("Lasso Coefficient Paths", line = 2.5)
# Use fit_cv_glmnet(..., alpha = 1, foldid = foldid).
# Report lambda.min, lambda.1se, paths, and nonzero predictors.
```


## 2D. Training-based comparison and locked core model

```{r core-comparison, eval=FALSE}
# Build a comparison from training/CV quantities only.
```


**Core model nominated before holdout evaluation:** [method and reason]

## 2E. Perturbation or stability check

**Prediction before running the check:** [prediction]

```{r perturbation, eval=FALSE}
# Implement one allowed training-only stability check.
```


**Observed result and explanation:** [result]

# Problem 3. Mathematical mechanisms

## 3A. Ridge and conditioning

*Write the ridge derivation. Define dimensions and explain uniqueness.*

Let
$X \in \mathbb{R}^{n \times p}, y \in \mathbb{R}^n, b \in \mathbb{R}^p$
and define $$
G = \frac{1}{n} X^\top X, \quad
z = \frac{1}{n} X^\top y
$$ consequently, $G \in \mathbb{R}^{p \times p}$ and
$z \in \mathbb{R}^p$.

From the objective function
$$Q_\lambda(b) = \frac{1}{2n}\|y - Xb\|_2^2 + \frac{\lambda}{2}\|b\|_2^2$$

We expand Ridge objective function $$
\begin{align}
  Q_\lambda (b) &= \frac{1}{2n} \|y - X b\|_2^2 + \frac{\lambda}{2} \|b\|_2^2, quad (\lambda > 0) \\
  &= \frac{1}{2n} (y - X b)^\top (y - X b) + \frac{\lambda}{2} b^\top b \\
  &= \frac{1}{2n} (y^\top y - y^\top X b - b^\top X^\top y + b^\top X^\top X b) + \frac{\lambda}{2} b^\top b \\
  &= \frac{1}{2n} (y^\top y - 2 y^\top X b + b^\top X^\top X b) + \frac{\lambda}{2} b^\top b \\
\end{align}
$$

We take the gradient with respect to $b$ annd substitute $$
\begin{align}
  \nabla_b Q_\lambda (b) &= 0 - \frac{1}{n} y^\top X + \frac{1}{n} X^\top X b + \lambda b \\
  &= - z + G b + \lambda b
\end{align}
$$

Set $\nabla_b Q_\lambda (b) = 0 \Leftrightarrow (G + \lambda I_p) b = z$
which is the \textbf{normal equation}.

Because $G = \frac{1}{n}X^\top X$ is a positive semi-definite Gram
matrix, all its eigenvalues are $\ge 0$. Adding a penalty $\lambda > 0$
along the diagonal (i.e., $\lambda I_p$) shifts all eigenvalues up by
$\lambda$. Thus, the matrix $(G + \lambda I_p)$ is strictly positive
definite (all eigenvalues $\ge \lambda > 0$), making it full rank and
invertible. This guarantees a unique minimizer

$$\hat{b}_\lambda^{\text{ridge}} = (G + \lambda I_p)^{-1} z$$

```{r ridge-conditioning, eval=FALSE}
# Use x_train and the fitted ridge lambda.min.
# safe_condition_numbers(x_train, lambda = ...)

# Assume x_train, y_train, and lambda.min from a previous cv.glmnet are loaded
n <- nrow(x_train)

# Compute G and z
G <- (1/n) * t(x_train) %*% x_train

# Compute the eigenvalues of G
evals_G <- eigen(G)$values
mu_max <- max(evals_G)
mu_min <- min(evals_G)

# Compute kappa_2(G)
kappa_G <- ifelse(zapsmall(mu_min) == 0, Inf, mu_max / mu_min)

# Compute kappa_2(G + lambda*I)
lambda_min <- ... # Insert your extracted lambda.min
kappa_ridge <- (mu_max + lambda_min) / (mu_min + lambda_min)

cat("Condition number of G (kappa_G):", kappa_G, "\n")
cat("Condition number of Ridge (kappa_ridge):", kappa_ridge, "\n")
```
*Explanation.*
- *Numerically:* Condition number measures how sensitive the solution is to small changes or errors in the data. An infinite or very large condition number means $G$ is singular or ill-conditioned, making matrix inversion unstable or impossible. By adding a positive constant $\lambda$ to the diagonal elements, we strictly bound the smallest eigenvalue away from zero. This reduces the condition number, stabilizing the matrix inversion and computational precision.
Numerically:Without Ridge: If $X$ has highly correlated columns, $\mu_{\min}$ is very close to $0$, causing $\kappa_2(G)$ to blow up to infinity. This makes the matrix inversion highly unstable and sensitive to floating-point errors.
With Ridge: Adding $\lambda > 0$ to the diagonal ensures the smallest eigenvalue is strictly bounded away from zero ($\mu_{\min} + \lambda \ge \lambda$). This dramatically decreases the condition number, stabilizing the matrix inversion and making the numerical computation robust.

- *Geometrically:* The matrix $G$ represents the curvature of the unpenalized least-squares loss surface. If $G$ has a zero (or near-zero) eigenvalue, the loss surface forms a flat "valley" in that principal direction, meaning there is no unique bottom (minimizer). Adding $\lambda I_p$ adds isotropic (uniform spherical) curvature to the loss function. Geometrically, this bends those flat valleys upward into a strictly convex, bowl-shaped surface, ensuring a single, globally unique minimum exists.
Geometrically:
Without Ridge (OLS): The contours of the least squares loss function can form long, narrow ellipses (or flat valleys if perfectly collinear), indicating high uncertainty in certain directions.
With Ridge: Adding the $L_2$ penalty $\frac{\lambda}{2}\Vert{}b\Vert{}_2^2$ is equivalent to shrinking the solution towards the origin. Geometrically, the contours of $(G + \lambda I_p)$ become more circular (spherical) compared to $G$. The feature space is restricted, reducing variance along the directions of smallest variance (the smallest eigenvalues).

## 3B. Lasso optimality and soft thresholding

*Derive the zero/nonzero optimality conditions and the
orthonormal-design formula.*

Let
$X \in \mathbb{R}^{n \times p}, y \in \mathbb{R}^n, b \in \mathbb{R}^p$
and define $$
G = \frac{1}{n} X^\top X = I_p, \quad
z = \frac{1}{n} X^\top y
$$ consequently, $G \in \mathbb{R}^{p \times p}$ and
$z \in \mathbb{R}^p$.

The objective function $$
R_\lambda (b) = \frac{1}{2n} \|y - X b\|_2^2 + \lambda \|b\|_1, \quad \|b\|_1 = \sum_{j=1}^p |b_j|, \lambda > 0
$$

We expand Lasso objective function $$
\begin{aligned}
R_\lambda (b) &= \frac{1}{2n} \|y - X b\|_2^2 + \lambda \|b\|_1 \\
              &= \frac{1}{2n} (y - X b)^\top (y - X b) + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} (y^\top y - y^\top X b - b^\top X^\top y + b^\top X^\top X b) + \lambda \sum_{j=1}^p |b_j|
\end{aligned}
$$

The term $y^\top X b$ is is a scalar, therefore
$(y^\top X b)^\top = b^\top X^\top (y^\top)^\top = b^\top X^\top y$.

Then the objective becomes $$
\begin{aligned}
R_\lambda (b) &= \frac{1}{2n} y^\top y - \frac{1}{n} b^\top X^\top y + \frac{1}{2n} b^\top X^\top X b + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} y^\top y - b^\top z + \frac{1}{2} b^\top b + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} y^\top y + \sum_{j=1}^p \left( - b_j z_j + \frac{1}{2} b_j^2 + \lambda |b_j| \right)
\end{aligned}
$$

The term $\frac{1}{2n} y^\top y$ is a constant with respect to the
variable $b$. We can drop it, leaving a sum of $p$ independent
1-dimensional optimization problems.

Consider a fixed $j$, and let
$l_j (t) = \frac{1}{2} t^2 - t z_j + \lambda |t|$, for
$t \in \mathbb{R}$. We analyze this in three cases because $|t|$ is
*non-differentiable* at $t=0$.

-   **Case 1:** If $t > 0$, then $|t| = t$ and
    $l_j (t) = \frac{1}{2} t^2 - t z_j + \lambda t$. Thus,
    $l_j' (t) = t - z_j + \lambda$.
    $$ l_j' (t) = 0 \iff t = z_j - \lambda \quad (1) $$ Moreover, since
    we assumed $t > 0$, equation $(1)$ is a valid minimum if and only if
    $z_j > \lambda$.

-   **Case 2:** If $t < 0$, then $|t| = - t$ and
    $l_j (t) = \frac{1}{2} t^2 - t z_j - \lambda t$. Thus,
    $l_j' (t) = t - z_j - \lambda$.
    $$ l_j' (t) = 0 \iff t = z_j + \lambda \quad (2) $$ Moreover, since
    we assumed $t < 0$, equation $(2)$ is a valid minimum if and only if
    $z_j < -\lambda$.

-   **Case 3:** If $|z_j| \le \lambda$, there is no valid solution where
    the derivative equals zero (neither Case 1 nor Case 2). Because the
    function is strictly convex, the global minimum must exist, which
    means it is trapped exactly at the non-differentiable corner:
    $t = 0$. We verify that $t = 0$ minimizes $l_j (t)$ (3).

For any $t \in \mathbb{R}$ $$
\begin{aligned}
l_j (t) - l_j (0) &= \frac{1}{2} t^2 - t z_j + \lambda |t| \\
                  &\ge \frac{1}{2} t^2 - |t| |z_j| + \lambda |t| \quad (-t z_j \ge -|t| |z_j|) \\
                  &\ge \frac{1}{2} t^2 - |t| (|z_j| - \lambda) \\
                  &\ge 0 \quad (|z_j| - \lambda \le 0)
\end{aligned}
$$ Therefore, $l_j (t) \ge l_j (0)$ for all $t \in \mathbb{R}$, proving
$t=0$ is the global minimum for this range.

Combining these cases yields the soft-thresholding estimator for
$\hat{b}_j$: $$
\hat{b}_j = \begin{cases}
  z_j - \lambda & \text{if } z_j > \lambda, \\
  0             & \text{if } |z_j| \le \lambda, \\
  z_j + \lambda & \text{if } z_j < -\lambda
\end{cases}
$$

**Soft-thresholding operator under orthonormal design**
Assume that $\frac{1}{n} X^T X = I_p$. Let $z = \frac{1}{n} X^T y$. Then the Lasso minimizer of $R_\lambda (b)$ is _unique_ and has coordinates
$$
\hat{b}_j = S_\lambda (z_j), \quad \text{where the soft-thresholding function is } S_\lambda (z) = \text{sgn}(z) (|z| - \lambda)_+
$$ where $(u)_+ = \max(u, 0)$.

**Uniqueness.** The uniqueness is guaranteed because the objective
function is strictly convex. Under the orthogonal design
($\frac{1}{n} X^T X = I_p$), the variables are perfectly independent,
forming a strictly convex "bowl" shape. Even with the sharp point added
by the $L_1$ penalty, the strict convexity mathematically ensures there
is exactly one absolute global minimum.

**Intuition.**
Under $\frac{1}{n} X^T X = I_p$, we can compare the closed-form estimators
- **Ridge:** $\hat{b}_j^{\text{Ridge}} = (1 + 2 \lambda)^{-1} z_j$ (proportional shrinkage). It divides the coefficient by a constant. The coefficients get extremely small, but they never become zero. The optimality condition means $b_j$ only equals exactly zero if $z_j$ is exactly zero (which is highly improbable with real data).
- **Lasso:** $\hat{b}_j^{\text{Lasso}} = \text{sgn}(z_j) (|z_j| - \lambda)_+$ (absolute shrinkage). It subtracts a constant from the magnitude. If a coefficient is originally small, the penalty forces it to exactly zero. Case 3 shows that Lasso maps a whole range of values to exactly zero, creating sparsity. This gives Lasso its *feature selection* capability.

## 3C. Connect algebra to evidence

*Connect one specific numerical result to Problem 3A or 3B.*

# Problem 4. Elastic net and a neural-feature bridge

## 4A. Research question and source map

**Research direction:** [weight decay / sparse weights / pruning /
dropout]

| Claim | Exact primary source and locator | What it establishes | What it does not establish |
|------------------|------------------|------------------|------------------|
| [Claim] | [Citation, section/page/equation] | [Support] | [Boundary] |

## 4B. Elastic net on original predictors

```{r original-elastic-net, eval=FALSE}
alpha_grid <- c(0.10, 0.25, 0.50, 0.75, 0.90)

# Fit one cv.glmnet object per alpha with the same foldid.
# Select alpha and lambda using training CV only.
```


## 4C. Fixed ReLU features

```{r random-relu-features, eval=FALSE}
relu <- function(z) pmax(z, 0)
M <- 200L

# TODO:
# 1. Generate A and bias once from seeds$features using the stated distributions.
# 2. Apply the same A and bias to x_train and x_test.
# 3. Fit a scaler to H_train only and apply it to H_test.
# 4. Fit ridge, lasso, and elastic net on H_train with the shared foldid.
```


*State which weights are fixed, which are learned, and how active
features are counted.*

## 4D. Locked declaration and final holdout evaluation

> **Locked before viewing `y_test`:** [preprocessing, candidate
> specifications, selected tuning values, feature seed, nominated
> deployment model, metrics]

```{r final-holdout-evaluation, eval=FALSE}
# This must be the first analysis chunk that uses y_test for evaluation.
# Generate all fixed candidate predictions and one RMSE/MAE table here.
```


*State whether the holdout evidence supports the predeclared choice. Do
not retune.*

## 4E. Deep-learning connection and limitations

*Distinguish penalty, weight decay, output-weight sparsity, pruning,
dropout, and a trained deep network. Cite the sources from 4A and
discuss at least two limitations.*

# Problem 5. Report quality and reproducibility

## 5A. Reproduction record

``` text
1. Open a clean R session.
2. [Exact commands and folder layout.]
3. Render GroupXX_ProjectQuiz1.Rmd.
```

## 5B. Recommendation and limitations

*Give the final recommendation, distinguish prediction from
interpretation, and state limitations.*

## 5C. AI verification record

| AI-assisted item | How it was checked | Evidence | Modification or rejection |
|------------------|------------------|------------------|------------------|
| [Item] | [Method] | [Test/source/output] | [Decision] |

# Preflight and session information {.unnumbered}

```{r preflight, eval=FALSE}
# Add every object required to reproduce the final table and figures.
required_objects <- c(
  "train_id", "test_id", "x_train", "x_test", "y_train", "y_test", "foldid"
)
missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects)) {
  stop(
    "Missing required objects: ",
    paste(missing_objects, collapse = ", ")
  )
}
stopifnot(length(intersect(train_id, test_id)) == 0L)
```

```{r session-info}
sessionInfo()
```