---
title: "ASESII 24A02 Project-Based Quiz 1"
subtitle: "Ridge, Lasso, and Regularization for Neural Features"
author: "Group 06: [names and student IDs]"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
bibliography: references.bib
link-citations: true
params:
  group_number: 6
fontsize: 11pt
geometry: margin=1in
editor_options: 
  markdown: 
    wrap: 72
---

# Authorship and contribution statement {-}

We confirm that every group member can explain the submitted analysis.
AI use is reported in `GroupXX_AI_Log.csv`, and the group checked every
AI-assisted item used in this report.

| Student               | ID       | Main contributions | Verification performed |
|-------------------|-----------------|-----------------|--------------------|
| Le Dat Phuc An        | 24125052 | [Work]             | [Check]                |
| Pham Nguyen Minh Quan | 24125041 | [Work]             | [Check]                |
| Tran Canh Anh Tuan    | 24125048 | [Work]             | [Check]                |
| Vong Chi Van          | 24125049 | [Work]             | [Check]                |

# Abstract {.unnumbered}

*Maximum 150 words: prediction problem, methods, main evidence, and
recommendation.*

# Reproducible setup {.unnumbered}

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("tidyverse", "glmnet", "broom", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)

group_number <- as.integer(params$group_number)
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number
)
```

## Shared helper functions {.unnumbered}

```{r helpers}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction)))
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  G <- crossprod(x) / nrow(x)
  eigenvalues <- eigen(G, symmetric = TRUE, only.values = TRUE)$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
plot_model_result <- function(model, model_name) { 
   model_result <- augment(model)
  ggplot(model_result, aes(x = .fitted, y = .resid)) +
    geom_point(color = "steelblue", size = 2, alpha = 0.8) +
    
    geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
    
    theme_minimal() +

    labs(
      title = paste("Residual Plot Model:", model_name),
      subtitle = "Checking for non-linear patterns and constant variance",
      x = "Predicted lpsa (.fitted)",
      y = "Residuals (.resid)"
    ) 
}

vif_manual <- function(model) {
  X <- model.matrix(model)[, -1, drop = FALSE]

  sapply(seq_len(ncol(X)), function(i) {
    fit <- lm(X[, i] ~ X[, -i])
    1 / (1 - summary(fit)$r.squared)
  }) |> setNames(colnames(X))
}


```

# Problem 1. Prediction design and leakage control

## 1A. Target and predictors

```{r data-configuration, eval=FALSE}
config <- if (group_number %% 2L == 1L) {
  list(
    file = "fat.csv",
    response = "brozek",
    excluded = c("brozek", "siri", "density", "free")
  )
} else {
  list(
    file = "prostate.csv",
    response = "lpsa",
    excluded = "lpsa"
  )
}

data_raw <- read.csv(find_exam_data(config$file), check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}
```

*Define the prediction task and justify every exclusion.*

### 1. Response = $lpsa$

$lpsa$ measures the natural Logarithm of Prostate-Specific Antigen,
which is a continuous number

### 2. Candidate Predictors

|                      |                                  |                     |
|---------------------|-------------------------------|--------------------|
| Candidate Predictors | Full Name                        | Type                |
| lcavol               | Log Cancer Volume                | Continuous          |
| lweight              | Log Prostate Weight              | Continuous          |
| age                  | Age                              | Continuous          |
| lbph                 | Log Benign Prostatic Hyperplasia | Continuous          |
| svi                  | Seminal Vesicle Invasion         | Binary (0 / 1)      |
| lcp                  | Log Capsular Penetration         | Continuous          |
| gleason              | Gleason Score                    | Ordered Categorical |
| pgg45                | Percentage Gleason Grade 4/5     | Continuous (0–100)  |

### 3. Unit of Observation

Each row of the dataset contains the information of a radical
prostactomy treated patient which comprises of the target and the
predictors values.

### 4. Study Population

The dataset contains 97 patients aged 41 to 79 with clinically localized
adenocarcinoma who went through radical prostactomy at Stanford
University Center by Dr. Thomas A. Stamey and his surgical team during
the late 1980s.

### 5. Prediction Objective

To build a regression model that predicts the log prostate-specific
antigen level (lpsa) using the clinical variables (lcavol, lweight, age,
lbph, svi, lcp, gleason, and pgg45).

### 6. Dataset Source

<https://sci-hub.sg/storage/2024/6429/4cbb75646440af614abced23a70ba059/stamey1989.pdf>

### 7. Required Exclusions

The response variable, `lpsa`, was excluded from the predictor matrix
because it is the outcome to be predicted. The remaining eight clinical
variables were retained as predictors. Since the dataset contains no
variables derived from `lpsa`, no additional exclusions were required.
This prevents target leakage and ensures a fair comparison among the
regression models.

## 1B. Split and train-only preprocessing

```{r split-and-scale, eval=FALSE}
split <- split_rows(nrow(analysis_data), seed = seeds$split)
train_id <- split$train
test_id <- split$test

x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_prep <- fit_scaler(x_raw[train_id, , drop = FALSE])
x_train <- apply_scaler(x_raw[train_id, , drop = FALSE], x_prep)
x_test <- apply_scaler(x_raw[test_id, , drop = FALSE], x_prep)
y_train <- y_raw[train_id]
y_test <- y_raw[test_id]

foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
```

*Report seeds, dimensions, scaling, folds, and leakage safeguards.*

### 1. Split seed

- Seed `split` for 80/20 split: 240206
- Seed `cv` for K-fold split: 240306

### 2. Train/Test Split

The dataset of 97 observations was partitioned into:

- **Training Set:** 77 observations
- **Test Set:** 20 observations

### 3. Feature Scaling Rule

Continuous predictor variables were centered and scaled using the helper
functions `fit_scaler()` and `apply_scaler()`. The scaling parameters
were estimated **only from the training data** in `fit_scaler()` and
then applied unchanged to both the training and test sets by
`apply_scaler()`.

Specifically, each predictor was transformed as
$$
x_{\text{scaled}} =
\frac{x - \bar{x}_{\text{train}}}
{\sqrt{\frac{1}{n_{\text{train}}} \sum_{i=1}^{n_{\text{train}}} (x_i - \bar{x}_{\text{train}})^2}},
$$
where \(\bar{x}_{\text{train}}\) is the mean of the predictor in the training
set, and the denominator is the population standard deviation computed from
the training observations. Using the same scaling parameters for the test set
prevents information leakage from the holdout data.

### 4. Five-Fold Assignment Rule

A single, immutable five-fold cross-validation assignment vector
(`foldid`) was generated on the 77 training observations using the
designated `cv` random seed. To ensure fair and controlled comparisons,
this exact same fold assignment was reused across all models evaluated:

- Ridge Regression
- Lasso Regression
- Elastic Net
- All subsequent hyperparameter grid evaluations

### 5. Missing-Value Plan

No missing values were detected. If any had been present, they would
have been handled using a training-only imputation procedure before
model fitting.

## 1C. Training-data exploration

The exploratory analysis in this section uses only the training
observations to avoid information leakage from the holdout test set.

### Response distribution

```{r, eval=FALSE}
summary(y_train)
sd(y_train)

hist(
  y_train,
  main = "Histogram of y_train",
  xlab = "y_train"
)

qqnorm(y_train)
qqline(y_train)
```

The response variable (`lpsa`) has a mean of **2.54** and a median of
**2.59**, indicating an approximately symmetric distribution. Most
observations lie between **1.73** (first quartile) and **3.28** (third
quartile), with a standard deviation of **1.15**.

The histogram suggests that most observations are concentrated around
the center, while only a few observations appear in the lower and upper
tails. The normal Q–Q plot shows that most observations follow the
reference line reasonably well, with slight departures in the tails.
Overall, the response distribution is approximately normal, making it
suitable for linear regression.

```{r, eval=FALSE}
cor_matrix <- cor(analysis_data[train_id, ])

heatmap_df <- expand.grid(
  Var1 = rownames(cor_matrix),
  Var2 = colnames(cor_matrix)
)

heatmap_df$value <- c(cor_matrix)

ggplot(heatmap_df, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 3) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  coord_fixed() +
  theme_minimal()
```

The correlation heatmap shows that **lcavol** has the strongest positive
correlation with the response ($r = 0.75$), indicating that tumor volume
is the most informative individual predictor of PSA level. Moderate
positive correlations are also observed for **lcp** ($r = 0.59$) and
**svi** ($r = 0.58$), whereas `age`, `lbph`, and `lweight` exhibit
relatively weak marginal relationships with the response.

### One possible anomaly or influential observation

```{r, eval=FALSE}
boxplot(analysis_data[train_id, ])
```

The boxplots identify several potential outliers, particularly in
**age** and **pgg45**. Among the predictors, `pgg45` exhibits the
largest variability, while `age` contains several observations outside
the whiskers. These observations are retained because there is no
evidence that they result from data entry errors.

### One pair or group of predictors for which multicollinearity may matter

```{r, eval=FALSE}
```

The strongest correlations among the predictors are observed between
`gleason` and `pgg45` ($r = 0.76$), `lcavol` and `lcp` ($r = 0.73$),
`lcp` and `svi` ($r = 0.67$), and `lcavol` and `svi` ($r = 0.59$). These
predictor pairs appear to contain overlapping information, suggesting
that multicollinearity may be present. Consequently, ordinary least
squares coefficient estimates may become unstable and exhibit inflated
variance, motivating the use of ridge and lasso regression to improve
coefficient stability and variable selection.

### Question for regularized models

Based on the exploratory analysis, the following question will guide the
remaining analysis:

> Can regularization improve predictive performance and coefficient
> stability when predictors contain overlapping information?

*Interpret multicollinearity, one anomaly, and the question
regularization will address.*

# Problem 2. OLS, ridge, and lasso

## 2A. OLS baseline

The ordinary least squares (OLS) model was fitted using only the
standardized training observations. Throughout this section, the holdout
test set remained completely untouched to ensure a fair comparison with
the regularized regression models.

### Fit OLS

```{r ols-fit, eval=FALSE}
x_train_df <- as.data.frame(x_train)
x_train_df$lpsa <- y_train

ols_model <- lm(lpsa ~ ., data = x_train_df)
```

The fitted OLS model serves as the baseline model for all subsequent
comparisons with ridge and lasso regression.

------------------------------------------------------------------------

### Coefficient estimates

```{r ols-summary, eval=FALSE}
summary(ols_model)
```

The fitted OLS model is statistically significant overall (**F =
15.78**, p \< 0.001), indicating that at least one predictor contributes
to explaining the variation in the response variable.

The model achieves a **Multiple** $R^2$ of 0.6500 and an **Adjusted**
$R^2$ of 0.6088, suggesting that approximately **65%** of the
variability in `lpsa` is explained by the eight clinical predictors.

Among the predictors,

- **lcavol** ($\beta=0.674$, p \< 0.001),
- **lweight** ($\beta=0.202$, p = 0.0438),
- **age** ($\beta=-0.194$, p = 0.0435), and
- **svi** ($\beta=0.282$, p = 0.0206)

are statistically significant at the 5% significance level.

The remaining predictors (`lbph`, `lcp`, `gleason`, and `pgg45`) are not
individually significant after adjusting for the other predictors in the
model.

The estimated coefficient for **age** is negative ($\beta=-0.194$). This
result suggests that the estimated effect of `age` changes after
accounting for the remaining predictors in the model. The possibility of
correlation among predictors is therefore investigated using variance
inflation factors in the following subsection.

### Training error

```{r ols-training-error, eval=FALSE}
train_pred <- predict(m1)

train_mse <- mean((y_train - train_pred)^2)
train_rmse <- sqrt(train_mse)

cat("Training MSE :", round(train_mse, 4), "\n")
cat("Training RMSE:", round(train_rmse, 4), "\n")
```

The fitted OLS model achieved

- **Training MSE = 0.4544**
- **Training RMSE = 0.6741**

These quantities measure the goodness-of-fit on the training
observations used to estimate the regression coefficients. Since the
model is evaluated on the same data used for fitting, the training error
is expected to underestimate the prediction error on unseen
observations.

------------------------------------------------------------------------

### Five-fold cross-validation

```{r ols-cv, eval=FALSE}
cv_mse <- numeric(5)
coefs_list <-list()
for(i in 1:5){

  train_fold <- x_train_df[foldid != i, ]
  valid_fold <- x_train_df[foldid == i, ]

  model <- lm(lpsa ~ ., data = train_fold)

  pred <- predict(model, newdata = valid_fold)
  fold_coef <- as.matrix(coef(model, s="lambda.min"))
  coefs_list[[i]] <- as.data.frame(t(fold_coef))
  cv_mse[i] <- mean((valid_fold$lpsa - pred)^2)
}
coef_ols_df <- bind_rows(coefs_list)
cv_table <- data.frame(
  Fold = 1:5,
  Validation_MSE = round(cv_mse, 4)
)

knitr::kable(cv_table)

cat("Mean CV MSE :", round(mean(cv_mse), 4), "\n")
cat("Mean CV RMSE:", round(sqrt(mean(cv_mse)), 4), "\n")
```

The validation errors obtained from the five folds are

| Fold | Validation MSE |
|-----:|---------------:|
|    1 |         0.8259 |
|    2 |         0.8568 |
|    3 |         0.4011 |
|    4 |         0.4893 |
|    5 |         0.9002 |

The average cross-validation error is

- **Mean CV MSE = 0.6947**
- **Mean CV RMSE = 0.8335**

As expected, the cross-validation error is larger than the training
error because each validation fold consists of observations that were
not used to estimate the regression coefficients. Although some
variation exists among the five folds, the validation errors remain
within a similar range, suggesting that the predictive performance of
the OLS model is reasonably stable across different training subsets.

------------------------------------------------------------------------

### Residual diagnostics

#### Residual versus fitted values

```{r ols-residual, eval=FALSE}
plot(
  fitted(m1),
  resid(m1),
  xlab = "Fitted values",
  ylab = "Residuals",
  main = "Residuals vs Fitted"
)

abline(h = 0, col = "red")
```

The residual-versus-fitted plot shows that the residuals are randomly
scattered around zero without noticeable curvature or funnel-shaped
patterns. This suggests that the assumptions of

- linearity, and
- constant error variance (homoscedasticity)

are reasonably satisfied. No obvious systematic pattern indicating model
misspecification is observed.

#### Normal Q-Q plot

```{r ols-qqplot, eval=FALSE}
qqnorm(resid(m1))
qqline(resid(m1), col = "red")
```

The normal Q-Q plot indicates that most residuals lie close to the
reference line, with only slight departures in the two tails. Therefore,
the residual distribution appears approximately normal.

The normal Q–Q plot indicates that the residuals are approximately
normally distributed, with only slight deviations in the tails. This
behavior is consistent with the mildly skewed response distribution
observed during the exploratory analysis. Since the residuals satisfy

$$\hat{\mathbf e}=(I-H)\mathbf y,$$

where $\hat{\mathbf e}$ is the residual vector, $\mathbf y$ is the
response vector, $I$ is the identity matrix, and
$H=X(X^\top X)^{-1}X^\top$ is the hat matrix, their distribution
naturally reflects the response after removing the fitted linear
component. Overall, there is no strong evidence against the normality
assumption.

#### Influence plot

```{r ols-influence, eval=FALSE}
plot(ols_model, which = 5)
```

Most observations have low leverage and moderate residuals. Although a
few observations (e.g., 23, 37, and 54) stand out slightly, none appears
to have an unusually large influence on the fitted model.

#### Cook's Distance

```{r ols-cook-distance, eval=FALSE}
cook <- cooks.distance(ols_model)

plot(
  cook,
  type = "h",
  lwd = 3,
  col = "red",
  xlab = "Observation",
  ylab = "Cook's Distance",
  main = "Cook's Distance"
)

abline(
  h = 4 / nrow(x_train_df),
  col = "blue",
  lty = 2
)
```

Several observations exceed the reference line $4/n$, but all Cook's
distance values are well below 1. Therefore, there is no strong evidence
of influential observations.

Taken together, the residual diagnostics do not reveal any serious
violation of the standard assumptions underlying ordinary least squares
regression.

------------------------------------------------------------------------

### Ill-conditioned design

```{r ols-vif, eval=FALSE}
vif_manual(ols_model)
```

The variance inflation factors (VIFs) were computed to assess
multicollinearity among the predictors.

The VIF values range from **1.33** to **3.24**, with the largest values
observed for

- `lcp` (3.24),
- `pgg45` (2.97),
- `gleason` (2.68), and
- `lcavol` (2.42).

None of the predictors exceeds the commonly used threshold of **5**,
indicating that severe multicollinearity is not present in the training
design matrix.

This finding is consistent with the correlation analysis presented in
Section 1C. Although several predictor pairs, such as `gleason`–`pgg45`
and `lcavol`–`lcp`, exhibit moderate pairwise correlations, pairwise
correlation alone does not necessarily imply problematic
multicollinearity. The VIF measures how well each predictor can be
explained by **all remaining predictors simultaneously**, making it a
more appropriate diagnostic for coefficient instability.

Overall, the VIF results suggest that the correlations among predictors
are not sufficiently strong to produce substantial inflation of the
coefficient variances. Therefore, no clear evidence of an
ill-conditioned design was found, and the OLS estimates provide an
appropriate baseline for comparison with ridge and lasso regression in
the following sections.

------------------------------------------------------------------------

## 2B. Ridge

### Fitting Ridge

The ridge model is fitted using glmnet by setting alpha = 0 with the
training data.

```{r ridge, eval=FALSE}
# Use fit_cv_glmnet(..., alpha = 0, foldid = foldid).
# Report lambda.min, lambda.1se, paths, coefficients, and CV evidence.
ridge_model <- fit_cv_glmnet(x_train, y_train, alpha = 0, foldid = foldid)

plot(ridge_model)
abline(v = -log(ridge_model$lambda.min), col = "green", lty = 2, lwd = 1.5)
abline(v = -log(ridge_model$lambda.1se), col = "blue", lty = 2, lwd = 1.5)
y_pos <- max(ridge_model$cvup) 
text(x = -log(ridge_model$lambda.min), y = y_pos, labels = "lambda.min", pos = 2, col = "green", cex = 0.9)
text(x = -log(ridge_model$lambda.1se), y = y_pos, labels = "lambda.1se", pos = 2, col = "blue", cex = 0.9)
```

The cross-validation curve demonstrates that the mean squared error
(MSE) reaches its minimum before increasing as the penalty becomes too
severe, the variance of the MSE also decrease when $\lambda$ decrease.
Because Ridge regression lacks a coordinate-wise thresholding mechanism,
it does not perform variable selection, exactly 8 non-zero predictors
remain active across the entire penalty path.

### Investigation on $\lambda_{min}$ and $\lambda_{1se}$

We can extract the specific penalties and their corresponding training
CV errors:

```{r}
cat(sprintf("lambda.min: %.5f\n", ridge_model$lambda.min))
cat(sprintf("lambda.1se: %.5f\n\n", ridge_model$lambda.1se))
```

Evaluating the coefficients at $\lambda_{min}$ and $\lambda_{1se}$
reveals how the penalty shrinks the estimates. At $\lambda_{min}$, the
largest coefficient is lcavol (0.4587) and the smallest is age
(-0.1123).

```{r}
min_id <- which(ridge_model$lambda == ridge_model$lambda.min)
cat(sprintf("cvm.min: %.5f\n", ridge_model$cvm[min_id]))
coef(ridge_model, s ="lambda.min")
```

As expected, the coefficients at the stricter $\lambda_{1se}$ are
shrunken even closer to zero.

```{r}
se1_id <- which(ridge_model$lambda == ridge_model$lambda.1se)
cat(sprintf("cvm.1se: %.5f\n\n", ridge_model$cvm[se1_id]))
coef(ridge_model, s ="lambda.1se")
```

### Coefficient paths

Plotting the coefficient paths visualizes the shrinkage. As the penalty
decreases (moving from left to right), the coefficients expand toward
their Ordinary Least Squares (OLS) values. On the far left, the model is
"blind" due to the massive penalty, so all weights are near $0$. As
moving to the right, the dominant predictors rise first because they
provide the biggest drop in error. Once the dominant signals are
established, the correlated predictors begin to rise or fall to fill in
the remaining details. If a variable's weight decreases, it is because
that variable is being used to counteract the dominant variables, or
because it is refining the model's prediction at the expense of a
simpler, more biased solution.

```{r}

# the plot automatically turn into the plot.glmnet function
# 1. Generate the base plot
plot(ridge_model$glmnet.fit, xvar = "lambda", label = TRUE)
title("Ridge Coefficient Paths", line = 2.5)

abline(v = -log(ridge_model$lambda.min), col = "green", lty = 2, lwd = 1.5)
abline(v = -log(ridge_model$lambda.1se), col = "blue", lty = 2, lwd = 1.5)

legend("topleft", 
       legend = c("lambda.min", "lambda.1se"), 
       col = c("green", "blue"), 
       lty = 2, 
       lwd = 1.5, 
       cex = 0.8, 
       bg = "white")

predictor_names <- rownames(ridge_model$glmnet.fit$beta)

legend("topleft", 
       inset = c(0, 0.21), # Pushes the legend down by 15% of the plot height
       legend = paste(1:length(predictor_names), predictor_names, sep = " - "), 
       col = 1:length(predictor_names), 
       lty = 1,                         
       lwd = 1.5, 
       cex = 0.8, 
       bg = "white",
       ncol = 2)
# Use fit_cv_glmnet(..., alpha = 0, foldid = foldid).
```
**TODO:** explanation

We can see the shrinkage corresponding to the OLS solution, at
$\lambda_{min}$ the length of the Ridge weight vector is around 0.71
with respect to the OLS weight. This is desirable since $\lambda\_{min}
= 0.28$, which is not a too strict penalty.

```{r}
coef_matrix <- as.matrix(coef(ridge_model$glmnet.fit)[-1, ])
ridge_l2_norms <- sqrt(colSums(coef_matrix^2))
ols_coefs <- coef(ols_model)[-1]
ols_l2_norm <- sqrt(sum(ols_coefs^2))

shrinkage_ratio <- ridge_l2_norms / ols_l2_norm

x_min <- -log(ridge_model$lambda.min)
x_1se <- -log(ridge_model$lambda.1se)

coef_min <- as.matrix(coef(ridge_model, s = "lambda.min"))[-1, ]
ratio_min <- sqrt(sum(coef_min^2)) / ols_l2_norm

coef_1se <- as.matrix(coef(ridge_model, s = "lambda.1se"))[-1, ]
ratio_1se <- sqrt(sum(coef_1se^2)) / ols_l2_norm

plot(-log(ridge_model$lambda), shrinkage_ratio, 
     type = "l", col = "purple", lwd = 2,
     xlab = expression(-log(lambda)), 
     ylab = expression(paste("||", hat(beta)[lambda]^R, "||"[2], " / ||", hat(beta)^{OLS}, "||"[2])),
     main = "Percentage Shrinkage by Lambda",
     ylim = c(0, 1))

abline(v = x_min, col = "green", lty = 2, lwd = 1.5)
abline(v = x_1se, col = "blue", lty = 2, lwd = 1.5)

points(x_min, ratio_min, pch = 19, col = "green", cex = 1.2)
points(x_1se, ratio_1se, pch = 19, col = "blue", cex = 1.2)


text(x_min, ratio_min, labels = sprintf("(%.2f, %.2f)", x_min, ratio_min), pos = 4, col = "green", cex = 0.8)
text(x_1se, ratio_1se, labels = sprintf("(%.2f, %.2f)", x_1se, ratio_1se), pos = 2, col = "blue", cex = 0.8)

legend("topleft", 
       legend = c("Shrinkage Ratio", "lambda.min", "lambda.1se"), 
       col = c("purple", "green", "blue"), 
       lty = c(1, 2, 2), lwd = 1.5, cex = 0.8, bg = "white")
```

### General Behavior of Correlated Predictors Under Ridge Regression

When predictors are highly correlated, OLS models struggle to isolate
their individual effects, often arbitrarily inflating one coefficient
while suppressing or reversing its partner. This causes extreme
instability across different data splits.

Ridge regression fixes this by heavily punishing excessively large
individual weights. Instead of letting correlated variables compete,
Ridge forces them to distribute the predictive weight evenly. This
shrinks the inflated coefficients, pulls the suppressed coefficients
back to its true direction, and drastically tightens their variance to
create a stable model.

### Examples from the Data

**1. The `lcavol` and `lcp` Pair (Correlation: 0.73)**

- **OLS:** Without the penalty, these variables compete. `lcavol`
  absorbs a massive positive weight (0.674) while `lcp` is buried in the
  negatives (-0.092). The OLS boxplots show wide, unstable variance for
  both.

- **Ridge:** Ridge forces weight sharing. `lcavol` is shrunk down to
  0.459, which allows `lcp` to flip to a biologically consistent
  positive value (0.082). The blue and green Ridge boxplots tighten
  significantly, proving the variables are now acting as a stable,
  unified group.

**2. The `pgg45` and `gleason` Pair (Correlation: 0.76)**

- **OLS:** Despite moving together in reality, they receive conflicting
  signs (`pgg45` at 0.165, `gleason` at -0.013). The OLS boxplot for
  `gleason` swings erratically across the zero-line, showing the model
  cannot confidently determine its direction.

- **Ridge:** Ridge punishes the inflated `pgg45` weight, shrinking it to
  0.110. This relieves the mathematical tension, pulling `gleason` out
  of the negatives up to 0.024. Their Ridge boxplots narrow drastically,
  locking their shared predictive power into a cooperative, stable
  state.

```{r}
library(tidyverse)
library(glmnet)

coef_ridge_min_list <- list()
coef_ridge_1se_list <- list()

for(i in 1:5) {
  fold_x <- x_train[foldid != i, , drop = FALSE]
  fold_y <- y_train[foldid != i]
  
  fold_ridge <- glmnet(fold_x, fold_y, alpha = 0)
  
  fold_coef_min <- as.matrix(coef(fold_ridge, s = ridge_model$lambda.min))
  fold_coef_1se <- as.matrix(coef(fold_ridge, s = ridge_model$lambda.1se))
  
  coef_ridge_min_list[[i]] <- as.data.frame(t(fold_coef_min))
  coef_ridge_1se_list[[i]] <- as.data.frame(t(fold_coef_1se))
}

coef_ridge_min_df <- bind_rows(coef_ridge_min_list)
coef_ridge_1se_df <- bind_rows(coef_ridge_1se_list)

long_ols <- coef_ols_df %>% 
  mutate(Model = "1. OLS") %>% 
  pivot_longer(cols = -Model, names_to = "Predictor", values_to = "Coefficient")

long_ridge_min <- coef_ridge_min_df %>% 
  mutate(Model = "2. Ridge (min)") %>% 
  pivot_longer(cols = -Model, names_to = "Predictor", values_to = "Coefficient")

long_ridge_1se <- coef_ridge_1se_df %>% 
  mutate(Model = "3. Ridge (1se)") %>% 
  pivot_longer(cols = -Model, names_to = "Predictor", values_to = "Coefficient")

plot_data <- bind_rows(long_ols, long_ridge_min, long_ridge_1se) %>%
  filter(Predictor != "(Intercept)")
variance_table <- plot_data %>%
  group_by(Predictor, Model) %>%
  summarize(Variance = var(Coefficient), .groups = 'drop') %>%
  pivot_wider(names_from = Model, values_from = Variance)

cat("--- Variance of Coefficients Across Folds ---\n")
print(variance_table)

ggplot(plot_data, aes(x = Predictor, y = Coefficient, fill = Model)) +
  geom_boxplot(alpha = 0.8, outlier.shape = 21, color = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 0.8) +
  theme_minimal() +
  coord_flip() +
  labs(
    title = "Coefficient Stability Across CV Folds",
    subtitle = "Comparing Variance: OLS vs. Ridge (lambda.min) vs. Ridge (lambda.1se)",
    x = "Predictors",
    y = "Coefficient Estimate"
  ) +
  scale_fill_manual(values = c("1. OLS" = "lightgray", 
                               "2. Ridge (min)" = "steelblue", 
                               "3. Ridge (1se)" = "seagreen")) +
  theme(legend.position = "bottom", legend.title = element_blank())
```

### SVD Analysis

```{r}
# 1. Standardize your training data (glmnet does this automatically by default)
x_scaled <- scale(x_train)

# 2. Extract the Singular Values (d) using SVD
d <- svd(x_scaled)$d

# 3. Create a sequence of lambda values to match your previous plots
# (glmnet scales the penalty by N, so we extract N as well)
N <- nrow(x_train)
lambdas <- ridge_model$glmnet.fit$lambda
log_lambdas <- log(lambdas)

# 4. Calculate the shrinkage factor for each of the 8 singular values
# Formula: d^2 / (d^2 + N * lambda)
shrinkage_factors <- matrix(0, nrow = length(lambdas), ncol = length(d))
for(j in 1:length(d)) {
  shrinkage_factors[, j] <- (d[j]^2) / (d[j]^2 + N * lambdas)
}

# 5. Plot the Shrinkage Profiles
matplot(-log_lambdas, shrinkage_factors, type = "l", lty = 1, lwd = 2,
        col = hcl.colors(length(d), palette = "Zissou 1"), # A clean color palette
        xlab = expression(log(lambda)),
        ylab = expression(frac(d[j]^2, d[j]^2 + N * lambda)),
        main = "SVD Shrinkage Profiles by Singular Value",
        ylim = c(0, 1))

# 6. Add reference lines for your optimal lambdas
abline(v = -log(ridge_model$lambda.min), col = "green", lty = 2, lwd = 1.5)
abline(v = -log(ridge_model$lambda.1se), col = "blue", lty = 2, lwd = 1.5)

# 7. Add a legend detailing the actual d values
legend_labels <- sprintf("d%d = %.2f", 1:length(d), d)
legend("topleft", legend = legend_labels, 
       col = hcl.colors(length(d), palette = "Zissou 1"), 
       lty = 1, lwd = 2, cex = 0.8, bg = "white")
library(tidyverse)

# 1. Convert the V_matrix into a format ggplot can read
# (Assuming V_matrix was created in the previous step)
v_data <- as.data.frame(V_matrix) %>%
  rownames_to_column(var = "Predictor") %>%
  pivot_longer(cols = -Predictor, 
               names_to = "Singular_Vector", 
               values_to = "Loading")

# 2. Lock the column order so d1 stays on the left and d8 on the right
v_data$Singular_Vector <- factor(v_data$Singular_Vector, levels = colnames(V_matrix))

# 3. Plot the Heatmap
ggplot(v_data, aes(x = Singular_Vector, y = Predictor, fill = Loading)) +
  geom_tile(color = "gray90", size = 0.5) +
  # Use a diverging color scale: Blue for positive, Red for negative, White for zero
  scale_fill_gradient2(low = "firebrick", mid = "white", high = "steelblue", midpoint = 0) +
  theme_minimal() +
  labs(
    title = "SVD Predictor Loadings (V Matrix)",
    subtitle = "Visualizing the composition of each Principal Direction",
    x = "Singular Vector (Ranked by Variance Explained)",
    y = "Original Predictor",
    fill = "Weight"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, face = "bold"),
    axis.text.y = element_text(face = "bold"),
    panel.grid = element_blank() # Removes messy gridlines behind the tiles
  )
```

```{r lasso, eval=FALSE}
model_lasso <- fit_cv_glmnet(x_train,y_train,alpha= 1, foldid = foldid)
plot(model_ridge)
cat(sprintf("lambda.min: %.5f\n", model_lasso$lambda.min))
cat(sprintf("lambda.1se: %.5f\n\n", model_lasso$lambda.1se))

cat("--- Nonzero Predictors (lambda.min) ---\n")
coef_min <- coef(model_ridge, s ='lambda.min')
print(c_min[coef_min[,1]!= 0])# mask which is 0 on the coefficient column
cat("\n--- Nonzero Predictors (lambda.1se) ---\n")
coef_1se <- coef(model_ridge, s ='lambda.1se')
print(c_1se[coef_1se[,1]!=0])

min_id <- which(model_lasso$lambda == model_lasso$lambda.min)
se1_id <- which(model_lasso$lambda == model_lasso$lambda.1se)
cat(sprintf("cvm.min: %.5f\n", model_lasso$cvm[min_id]))
cat(sprintf("cvm.1se: %.5f\n\n", model_lasso$cvm[se1_id]))

plot(model_lasso$glmnet.fit, xvar ='lambda', label = TRUE )
title("Lasso Coefficient Paths", line = 2.5)
# Use fit_cv_glmnet(..., alpha = 1, foldid = foldid).
# Report lambda.min, lambda.1se, paths, and nonzero predictors.
lasso_model <- fit_cv_glmnet(x_train, y_train, alpha = 1, foldid = foldid)

plot(lasso_model)

cat(sprintf("lambda.min: %.5f\n", lasso_model$lambda.min))
cat(sprintf("lambda.1se: %.5f\n", lasso_model$lambda.1se))

cat("\n--- Nonzero Predictors (lambda.min) ---\n")
coef_lasso_min <- coef(lasso_model, s = "lambda.min")
print(coef_lasso_min[coef_lasso_min[, 1] != 0]) # mask 0 on the coefficient column

cat("\n--- Nonzero Predictors (lambda.1se) ---\n")
coef_lasso_1se <- coef(lasso_model, s = "lambda.1se")
print(coef_lasso_1se[coef_lasso_1se[, 1] != 0])

cv_lasso_min <- lasso_model$cvm[which(lasso_model$lambda == lasso_model$lambda.min)]
cv_lasso_1se <- lasso_model$cvm[which(lasso_model$lambda == lasso_model$lambda.1se)]
cat(sprintf("cvm.min: %.5f\n", cv_lasso_min))
cat(sprintf("cvm.1se: %.5f\n\n", cv_lasso_1se))

plot(lasso_model$glmnet.fit, xvar = "lambda", label = TRUE)
title("Lasso Coefficient Paths", line = 2.5)
```
**TODO:** explanation

## 2D. Training-based comparison and locked core model

```{r core-comparison, eval=FALSE}
# Build a comparison from training/CV quantities only.

# calculate L2 norm (excluding intercept)
get_norm <- function(coefs) {
  coef_mat <- as.matrix(coefs)
  sqrt(sum(coef_mat[-1, ]^2, na.rm = TRUE))
}

# count non-zero slopes (excluding intercept)
get_nonzero <- function(coefs) {
  coef_mat <- as.matrix(coefs)
  sum(coef_mat[-1, ] != 0, na.rm = TRUE)
}

# extraction for OLS
coef_ols <- coef(ols_model)
norm_ols <- get_norm(coef_ols)
nz_ols <- get_nonzero(coef_ols)
cv_ols <- cv_mse_mean

# extraction for Ridge
cv_ridge_min 
coef_ridge_min <- coef(ridge_model, s = "lambda.min")
norm_ridge_min <- get_norm(coef_ridge_min)
nz_ridge_min <- get_nonzero(coef_ridge_min)

cv_ridge_1se
coef_ridge_1se <- coef(ridge_model, s = "lambda.1se")
norm_ridge_1se <- get_norm(coef_ridge_1se)
nz_ridge_1se <- get_nonzero(coef_ridge_1se)

# extraction for Lasso
cv_lasso_min
coef_lasso_min
norm_lasso_min <- get_norm(coef_lasso_min)
nz_lasso_min <- get_nonzero(coef_lasso_min)

cv_lasso_1se
coef_lasso_1se
norm_lasso_1se <- get_norm(coef_lasso_1se)
nz_lasso_1se <- get_nonzero(coef_lasso_1se)

comparison_table <- data.frame(
  Model = c("OLS", "Ridge", "Ridge", "Lasso", "Lasso"),
  tuning_rule = c("none", "min", "1se", "min", "1se"),
  cv_error = c(cv_ols, cv_ridge_min, cv_ridge_1se, cv_lasso_min, cv_lasso_1se),
  coef_norm = c(norm_ols, norm_ridge_min, norm_ridge_1se, norm_lasso_min, norm_lasso_1se),
  nonzero_slopes = c(nz_ols, nz_ridge_min, nz_ridge_1se, nz_lasso_min, nz_lasso_1se)
)

knitr::kable(comparison_table, digits = 4)
```

**Core model nominated before holdout evaluation:** [method and reason]

**Why lambda.min and lambda.1se answer different decision questions. Distinguish predictive performance from variable selection and interpretability.**

lambda.min (predictive performance)

- "What model gives the absolute lowest cross-validation error?" 
- It prioritizes raw predictive accuracy. 
- Because the CV curve is often flat near the minimum, it tends to retain more variables, some of which might be noise.

lambda.1se (variable selection and interpretability)

- "What is the simplest model whose performance is within an acceptable margin of error of the absolute best?" 
- It selects the most heavily penalized model whose CV error is within one standard error of the minimum. 
- It prioritizes fewer variables, making the model easier to interpret and less prone to overfitting.
- It suffers from a marginal drop in theoretical predictive performance.

## 2E. Perturbation or stability check

- Investigation
  - repeat five-fold cross-validation with three prespecified fold seeds -> randomness of cross-validation
  - duplicate a highly correlated predictor with small random noise -> how different regularizations handle multicollinearity
  - bootstrap the training observations and record lasso selection frequencies -> variable selection stability

**Prediction before running the check:** [prediction]

```{r perturbation, eval=FALSE}
# Implement one allowed training-only stability check.
set.seed(42)

# Select a known strong predictor
target_idx <- 1 
target_name <- colnames(x_train)[target_idx]

# Create a duplicate with tiny random noise
noise <- rnorm(nrow(x_train), mean = 0, sd = 0.01)
duplicate_var <- x_train[, target_idx] + noise

x_train_pert <- cbind(x_train, duplicate_var = duplicate_var)

# Fit Ridge and Lasso on the perturbed data
ridge_pert <- fit_cv_glmnet(x_train_pert, y_train, alpha = 0, foldid = foldid)
lasso_pert <- fit_cv_glmnet(x_train_pert, y_train, alpha = 1, foldid = foldid)

coef_ridge_pert <- coef(ridge_pert, s = "lambda.1se")
coef_lasso_pert <- coef(lasso_pert, s = "lambda.1se")

cat("--- Ridge Coefficients (Original vs Duplicate) ---\n")
print(coef_ridge_pert[c(target_name, "duplicate_var"), , drop = FALSE])

cat("\n--- Lasso Coefficients (Original vs Duplicate) ---\n")
print(coef_lasso_pert[c(target_name, "duplicate_var"), , drop = FALSE])
```

**Observed result and explanation:** [result]

# Problem 3. Mathematical mechanisms

## 3A. Ridge and conditioning

*Write the ridge derivation. Define dimensions and explain uniqueness.*

Let
$X \in \mathbb{R}^{n \times p}, y \in \mathbb{R}^n, b \in \mathbb{R}^p$
and define $$G = \frac{1}{n} X^\top X, \quad
z = \frac{1}{n} X^\top y$$ consequently, $G \in \mathbb{R}^{p \times p}$
and $z \in \mathbb{R}^p$.

From the objective function
$$Q_\lambda(b) = \frac{1}{2n}\|y - Xb\|_2^2 + \frac{\lambda}{2}\|b\|_2^2$$

We expand Ridge objective function $$\begin{align}
  Q_\lambda (b) &= \frac{1}{2n} \|y - X b\|_2^2 + \frac{\lambda}{2} \|b\|_2^2, quad (\lambda > 0) \\
  &= \frac{1}{2n} (y - X b)^\top (y - X b) + \frac{\lambda}{2} b^\top b \\
  &= \frac{1}{2n} (y^\top y - y^\top X b - b^\top X^\top y + b^\top X^\top X b) + \frac{\lambda}{2} b^\top b \\
  &= \frac{1}{2n} (y^\top y - 2 y^\top X b + b^\top X^\top X b) + \frac{\lambda}{2} b^\top b \\
\end{align}$$

We take the gradient with respect to $b$ and substitute $$\begin{align}
  \nabla_b Q_\lambda (b) &= 0 - \frac{1}{n} y^\top X + \frac{1}{n} X^\top X b + \lambda b \\
  &= - z + G b + \lambda b
\end{align}$$

Set $\nabla_b Q_\lambda (b) = 0 \Leftrightarrow (G + \lambda I_p) b = z$
which is the \textbf{normal equation}.

Because $G = \frac{1}{n}X^\top X$ is a positive semi-definite Gram
matrix, all its eigenvalues are $\ge 0$. Adding a penalty $\lambda > 0$
along the diagonal (i.e., $\lambda I_p$) shifts all eigenvalues up by
$\lambda$. Thus, the matrix $(G + \lambda I_p)$ is strictly positive
definite (all eigenvalues $\ge \lambda > 0$), making it full rank and
invertible. This guarantees a unique minimizer

$$\hat{b}_\lambda^{\text{ridge}} = (G + \lambda I_p)^{-1} z$$

```{r ridge-conditioning, eval=FALSE}
# Use x_train and the fitted ridge lambda.min.
# safe_condition_numbers(x_train, lambda = ...)

condition_numbers <- safe_condition_numbers(x_train, ridge_model$lambda.min)
kappa_G <- condition_numbers["gram"]
kappa_ridge <- condition_numbers["ridge"]

cat("Condition number of G (kappa_G):", kappa_G, "\n")
cat("Condition number of Ridge (kappa_ridge):", kappa_ridge, "\n")
```

**Explanation.**

- *Numerically:*
  - Condition number $\kappa_2(G)$ measures how sensitive the solution is to
small changes or errors in the data.
  - If $X$ has highly correlated columns, $\mu_{\min}$ is very close to $0$,
causing $\kappa_2(G)$ to blow up to infinity. This makes the matrix inversion
highly unstable and sensitive to floating-point errors.
  - In Ridge regression, by adding $\lambda > 0$ to the diagonal elements, we strictly
bound the smallest eigenvalue away from zero.
  - This reduces the condition number, stabilizing the matrix inversion and computational precision.

- *Geometrically:*
  - The matrix $G$ represents the curvature of the unpenalized least-squares
loss surface. If $G$ has a zero (or near-zero) eigenvalue, the loss surface
forms a flat valley in that principal direction, meaning there is no unique
bottom (minimizer).
  - Adding $\lambda I_p$ adds uniform spherical curvature to
the loss function. This bends those flat valleys upward into a strictly convex,
bowl-shaped surface, ensuring a single, globally unique minimum exists.
  - From a top-down perspective, the contours of the least squares loss function
can form long and narrow ellipses, indicating high uncertainty in certain directions.
  - Adding a constraint pull the optimal coefficients toward the origin. It
transforms the elliptical contours of $G$ into a more circular shape.
  - This stabilizes the model by shrinking coefficients in the directions where
the training data has the least variance (the smallest eigenvalues).

## 3B. Lasso optimality and soft thresholding

*Derive the zero/nonzero optimality conditions and the orthonormal-design formula.*

Let
$X \in \mathbb{R}^{n \times p}, y \in \mathbb{R}^n, b \in \mathbb{R}^p$
and define $$G = \frac{1}{n} X^\top X = I_p, \quad
z = \frac{1}{n} X^\top y$$ consequently, $G \in \mathbb{R}^{p \times p}$
and $z \in \mathbb{R}^p$.

The objective function
$$R_\lambda (b) = \frac{1}{2n} \|y - X b\|_2^2 + \lambda \|b\|_1, \quad \|b\|_1 = \sum_{j=1}^p |b_j|, \lambda > 0$$

We expand Lasso objective function $$\begin{aligned}
R_\lambda (b) &= \frac{1}{2n} \|y - X b\|_2^2 + \lambda \|b\|_1 \\
              &= \frac{1}{2n} (y - X b)^\top (y - X b) + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} (y^\top y - y^\top X b - b^\top X^\top y + b^\top X^\top X b) + \lambda \sum_{j=1}^p |b_j|
\end{aligned}$$

The term $y^\top X b$ is is a scalar, therefore
$(y^\top X b)^\top = b^\top X^\top (y^\top)^\top = b^\top X^\top y$.

Then the objective becomes $$\begin{aligned}
R_\lambda (b) &= \frac{1}{2n} y^\top y - \frac{1}{n} b^\top X^\top y + \frac{1}{2n} b^\top X^\top X b + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} y^\top y - b^\top z + \frac{1}{2} b^\top b + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} y^\top y + \sum_{j=1}^p \left( - b_j z_j + \frac{1}{2} b_j^2 + \lambda |b_j| \right)
\end{aligned}$$

The term $\frac{1}{2n} y^\top y$ is a constant with respect to the
variable $b$. We can drop it, leaving a sum of $p$ independent
1-dimensional optimization problems.

Consider a fixed $j$, and let
$l_j (t) = \frac{1}{2} t^2 - t z_j + \lambda |t|$, for
$t \in \mathbb{R}$. We analyze this in three cases because $|t|$ is
*non-differentiable* at $t=0$.

- **Case 1:** If $t > 0$, then $|t| = t$ and
  $l_j (t) = \frac{1}{2} t^2 - t z_j + \lambda t$. Thus,
  $l_j' (t) = t - z_j + \lambda$.
  $$ l_j' (t) = 0 \iff t = z_j - \lambda \quad (1) $$ Moreover, since we
  assumed $t > 0$, equation $(1)$ is a valid minimum if and only if
  $z_j > \lambda$.

- **Case 2:** If $t < 0$, then $|t| = - t$ and
  $l_j (t) = \frac{1}{2} t^2 - t z_j - \lambda t$. Thus,
  $l_j' (t) = t - z_j - \lambda$.
  $$ l_j' (t) = 0 \iff t = z_j + \lambda \quad (2) $$ Moreover, since we
  assumed $t < 0$, equation $(2)$ is a valid minimum if and only if
  $z_j < -\lambda$.

- **Case 3:** If $|z_j| \le \lambda$, there is no valid solution where
  the derivative equals zero (neither Case 1 nor Case 2). Because the
  function is strictly convex, the global minimum must exist, which
  means it is trapped exactly at the non-differentiable corner: $t = 0$.
  We verify that $t = 0$ minimizes $l_j (t)$ (3).

For any $t \in \mathbb{R}$ $$\begin{aligned}
l_j (t) - l_j (0) &= \frac{1}{2} t^2 - t z_j + \lambda |t| \\
                  &\ge \frac{1}{2} t^2 - |t| |z_j| + \lambda |t| \quad (-t z_j \ge -|t| |z_j|) \\
                  &\ge \frac{1}{2} t^2 - |t| (|z_j| - \lambda) \\
                  &\ge 0 \quad (|z_j| - \lambda \le 0)
\end{aligned}$$ Therefore, $l_j (t) \ge l_j (0)$ for all
$t \in \mathbb{R}$, proving $t=0$ is the global minimum for this range.

Combining these cases yields the soft-thresholding estimator for
$\hat{b}_j$: $$\hat{b}_j = \begin{cases}
  z_j - \lambda & \text{if } z_j > \lambda, \\
  0             & \text{if } |z_j| \le \lambda, \\
  z_j + \lambda & \text{if } z_j < -\lambda
\end{cases}$$

The uniqueness is guaranteed because the objective function is strictly convex.
Under the orthogonal design, the variables are perfectly independent,
forming a strictly convex "bowl" shape. Even with the sharp point added
by the $L_1$ penalty, the strict convexity mathematically ensures there
is exactly one absolute global minimum.

**Soft-thresholding operator under orthonormal design**
Assume that
$\frac{1}{n} X^T X = I_p$. Let $z = \frac{1}{n} X^T y$. Then the Lasso
minimizer of $R_\lambda (b)$ is *unique* and has coordinates
$$\hat{b}_j = S_\lambda (z_j), \quad \text{where the soft-thresholding function is } S_\lambda (z) = \text{sgn}(z) (|z| - \lambda)_+$$
where $(u)_+ = \max(u, 0)$.


**Why lasso can produce exact zeros while ridge generally does not.**
Under $\frac{1}{n} X^T X = I_p$, we can compare the closed-form estimators

- **Ridge:**
$\hat{b}_j^{\text{Ridge}} = (1 + 2 \lambda)^{-1} z_j$ (proportional
shrinkage). It divides the coefficient by a constant. The coefficients
get extremely small, but they never become zero. The optimality
condition means $b_j$ only equals exactly zero if $z_j$ is exactly zero
(which is highly improbable with real data).
- **Lasso:**
$\hat{b}_j^{\text{Lasso}} = \text{sgn}(z_j) (|z_j| - \lambda)_+$
(absolute shrinkage). It subtracts a constant from the magnitude. If a
coefficient is originally small, the penalty forces it to exactly zero.
Case 3 shows that Lasso maps a whole range of values to exactly zero,
creating sparsity. This gives Lasso its **feature selection** capability.

## 3C. Connect algebra to evidence

*Connect one specific numerical result to Problem 3A or 3B.*

# Problem 4. Elastic net and a neural-feature bridge

## 4A. Research question and source map

**Research direction:** [weight decay / sparse weights / pruning /
dropout]

| Claim | Exact primary source and locator | What it establishes | What it does not establish |
|------------------|------------------|------------------|------------------|
| [Claim] | [Citation, section/page/equation] | [Support] | [Boundary] |

## 4B. Elastic net on original predictors

```{r original-elastic-net, eval=FALSE}
alpha_grid <- c(0.10, 0.25, 0.50, 0.75, 0.90)

# Fit one cv.glmnet object per alpha with the same foldid.
# Select alpha and lambda using training CV only.
```

## 4C. Fixed ReLU features

```{r random-relu-features, eval=FALSE}
relu <- function(z) pmax(z, 0)
M <- 200L

# TODO:
# 1. Generate A and bias once from seeds$features using the stated distributions.
# 2. Apply the same A and bias to x_train and x_test.
# 3. Fit a scaler to H_train only and apply it to H_test.
# 4. Fit ridge, lasso, and elastic net on H_train with the shared foldid.
```

*State which weights are fixed, which are learned, and how active
features are counted.*

## 4D. Locked declaration and final holdout evaluation

> **Locked before viewing `y_test`:** [preprocessing, candidate
> specifications, selected tuning values, feature seed, nominated
> deployment model, metrics]

```{r final-holdout-evaluation, eval=FALSE}
# This must be the first analysis chunk that uses y_test for evaluation.
# Generate all fixed candidate predictions and one RMSE/MAE table here.
```

*State whether the holdout evidence supports the predeclared choice. Do
not retune.*

## 4E. Deep-learning connection and limitations

*Distinguish penalty, weight decay, output-weight sparsity, pruning,
dropout, and a trained deep network. Cite the sources from 4A and
discuss at least two limitations.*

# Problem 5. Report quality and reproducibility

## 5A. Reproduction record

``` text
1. Open a clean R session.
2. [Exact commands and folder layout.]
3. Render GroupXX_ProjectQuiz1.Rmd.
```

## 5B. Recommendation and limitations

*Give the final recommendation, distinguish prediction from
interpretation, and state limitations.*

## 5C. AI verification record

| AI-assisted item | How it was checked | Evidence | Modification or rejection |
|------------------|------------------|------------------|------------------|
| [Item] | [Method] | [Test/source/output] | [Decision] |

# Preflight and session information {.unnumbered}

```{r preflight, eval=FALSE}
# Add every object required to reproduce the final table and figures.
required_objects <- c(
  "train_id", "test_id", "x_train", "x_test", "y_train", "y_test", "foldid"
)
missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects)) {
  stop(
    "Missing required objects: ",
    paste(missing_objects, collapse = ", ")
  )
}
stopifnot(length(intersect(train_id, test_id)) == 0L)
```

```{r session-info}
sessionInfo()
```
