---
title: "ASESII 24A02 Project-Based Quiz 1"
subtitle: "Ridge, Lasso, and Regularization for Neural Features"
author: "Group 06: [names and student IDs]"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
params:
  group_number: 0
fontsize: 11pt
geometry: margin=1in
---

<!--
Copy this template into your group submission folder, rename it
GroupXX_ProjectQuiz1.Rmd, place references.bib beside it, and set the group
number parameter before rendering. Remove instructional comments from the
final report.
-->

# Authorship and contribution statement {-}

We confirm that every group member can explain the submitted analysis. AI use is reported in `GroupXX_AI_Log.csv`, and the group checked every AI-assisted item used in this report.

| Student | ID | Main contributions | Verification performed |
|---|---|---|---|
| [Name] | [ID] | [Work] | [Check] |

# Abstract {-}

*Maximum 150 words: prediction problem, methods, main evidence, and recommendation.*

# Reproducible setup {-}

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("tidyverse", "glmnet", "broom", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)

group_number <- as.integer(params$group_number)
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number
)
```

## Shared helper functions {-}

```{r helpers}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction)))
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  eigenvalues <- eigen(
    crossprod(x) / nrow(x),
    symmetric = TRUE,
    only.values = TRUE
  )$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
```

# Problem 1. Prediction design and leakage control

## 1A. Target and predictors

```{r data-configuration, eval=FALSE}
config <- if (group_number %% 2L == 1L) {
  list(
    file = "fat.csv",
    response = "brozek",
    excluded = c("brozek", "siri", "density", "free")
  )
} else {
  list(
    file = "prostate.csv",
    response = "lpsa",
    excluded = "lpsa"
  )
}

data_raw <- read.csv(find_exam_data(config$file), check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}
```

*Define the prediction task and justify every exclusion.*

## 1B. Split and train-only preprocessing

```{r split-and-scale, eval=FALSE}
split <- split_rows(nrow(analysis_data), seed = seeds$split)
train_id <- split$train
test_id <- split$test

x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_prep <- fit_scaler(x_raw[train_id, , drop = FALSE])
x_train <- apply_scaler(x_raw[train_id, , drop = FALSE], x_prep)
x_test <- apply_scaler(x_raw[test_id, , drop = FALSE], x_prep)
y_train <- y_raw[train_id]
y_test <- y_raw[test_id]

foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
```

*Report seeds, dimensions, scaling, folds, and leakage safeguards.*

## 1C. Training-data exploration

```{r training-eda, eval=FALSE}
# Use analysis_data[train_id, ] only. Add focused plots and summaries.
```

*Interpret multicollinearity, one anomaly, and the question regularization will address.*

# Problem 2. OLS, ridge, and lasso

## 2A. OLS baseline

```{r ols, eval=FALSE}
# Fit OLS on x_train and y_train. Do not access y_test here.
```

## 2B. Ridge

```{r ridge, eval=FALSE}
# Use fit_cv_glmnet(..., alpha = 0, foldid = foldid).
# Report lambda.min, lambda.1se, paths, coefficients, and CV evidence.
```

## 2C. Lasso

```{r lasso, eval=FALSE}
# Use fit_cv_glmnet(..., alpha = 1, foldid = foldid).
# Report lambda.min, lambda.1se, paths, and nonzero predictors.
```

## 2D. Training-based comparison and locked core model

```{r core-comparison, eval=FALSE}
# Build a comparison from training/CV quantities only.
```

**Core model nominated before holdout evaluation:** [method and reason]

## 2E. Perturbation or stability check

**Prediction before running the check:** [prediction]

```{r perturbation, eval=FALSE}
# Implement one allowed training-only stability check.
```

**Observed result and explanation:** [result]

# Problem 3. Mathematical mechanisms

## 3A. Ridge and conditioning

*Write the ridge derivation. Define dimensions and explain uniqueness.*

Let $X \in \mathbb{R}^{n \times p}, y \in \mathbb{R}^n, b \in \mathbb{R}^p$ and define
$$
G = \frac{1}{n} X^\top X, \quad
z = \frac{1}{n} X^\top y
$$
consequently, $G \in \mathbb{R}^{p \times p}$ and $z \in \mathbb{R}^p$.

From the objective function
$$Q_\lambda(b) = \frac{1}{2n}\|y - Xb\|_2^2 + \frac{\lambda}{2}\|b\|_2^2$$

We expand Ridge objective function
$$
\begin{align}
  Q_\lambda (b) &= \frac{1}{2n} \|y - X b\|_2^2 + \frac{\lambda}{2} \|b\|_2^2, quad (\lambda > 0) \\
  &= \frac{1}{2n} (y - X b)^\top (y - X b) + \frac{\lambda}{2} b^\top b \\
  &= \frac{1}{2n} (y^\top y - y^\top X b - b^\top X^\top y + b^\top X^\top X b) + \frac{\lambda}{2} b^\top b \\
  &= \frac{1}{2n} (y^\top y - 2 y^\top X b + b^\top X^\top X b) + \frac{\lambda}{2} b^\top b \\
\end{align}
$$

We take the gradient with respect to $b$ annd substitute
$$
\begin{align}
  \nabla_b Q_\lambda (b) &= 0 - \frac{1}{n} y^\top X + \frac{1}{n} X^\top X b + \lambda b \\
  &= - z + G b + \lambda b
\end{align}
$$

Set $\nabla_b Q_\lambda (b) = 0 \Leftrightarrow (G + \lambda I_p) b = z$ which is the \textbf{normal equation}.

Because $G = \frac{1}{n}X^\top X$ is a positive semi-definite Gram matrix, all its eigenvalues are $\ge 0$. Adding a penalty $\lambda > 0$ along the diagonal (i.e., $\lambda I_p$) shifts all eigenvalues up by $\lambda$. Thus, the matrix $(G + \lambda I_p)$ is strictly positive definite (all eigenvalues $\ge \lambda > 0$), making it full rank and invertible. This guarantees a unique minimizer

$$\hat{b}_\lambda^{\text{ridge}} = (G + \lambda I_p)^{-1} z$$

```{r ridge-conditioning, eval=FALSE}
# Use x_train and the fitted ridge lambda.min.
# safe_condition_numbers(x_train, lambda = ...)
```

```{r ridge-conditioning}
# Assume x_train, y_train, and lambda.min from a previous cv.glmnet are loaded
# n <- nrow(x_train)

# Compute G and z
# G <- (1/n) * t(x_train) %*% x_train

# Compute the eigenvalues of G
# evals_G <- eigen(G)$values
# mu_max <- max(evals_G)
# mu_min <- min(evals_G)

# Compute kappa_2(G)
# kappa_G <- ifelse(zapsmall(mu_min) == 0, Inf, mu_max / mu_min)

# Compute kappa_2(G + lambda*I)
# lambda_min <- ... # Insert your extracted lambda.min
# kappa_ridge <- (mu_max + lambda_min) / (mu_min + lambda_min)

# cat("Condition number of G (kappa_G):", kappa_G, "\n")
# cat("Condition number of Ridge (kappa_ridge):", kappa_ridge, "\n")
```

## 3B. Lasso optimality and soft thresholding

*Derive the zero/nonzero optimality conditions and the orthonormal-design formula.*

Let $X \in \mathbb{R}^{n \times p}, y \in \mathbb{R}^n, b \in \mathbb{R}^p$ and define
$$
G = \frac{1}{n} X^\top X = I_p, \quad
z = \frac{1}{n} X^\top y
$$
consequently, $G \in \mathbb{R}^{p \times p}$ and $z \in \mathbb{R}^p$.

The objective function
$$
R_\lambda (b) = \frac{1}{2n} \|y - X b\|_2^2 + \lambda \|b\|_1, \quad \|b\|_1 = \sum_{j=1}^p |b_j|, \lambda > 0
$$

We expand Lasso objective function
$$
\begin{aligned}
L_\lambda (b) &= \frac{1}{2n} \|y - X b\|_2^2 + \lambda \|b\|_1 \\
              &= \frac{1}{2n} (y - X b)^\top (y - X b) + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} (y^\top y - y^\top X b - b^\top X^\top y + b^\top X^\top X b) + \lambda \sum_{j=1}^p |b_j|
\end{aligned}
$$

The term $y^\top X b$ is is a scalar, therefore $(y^\top X b)^\top = b^\top X^\top (y^\top)^\top = b^\top X^\top y$.

Then the objective becomes
$$
\begin{aligned}
L_\lambda (b) &= \frac{1}{2n} y^\top y - \frac{1}{n} b^\top X^\top y + \frac{1}{2n} b^\top X^\top X b + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} y^\top y - b^\top z + \frac{1}{2} b^\top b + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} y^\top y + \sum_{j=1}^p \left( - b_j z_j + \frac{1}{2n} b_j^2 + \lambda |b_j| \right)
\end{aligned}
$$

The term $\frac{1}{2n} y^\top y$ is a constant with respect to the variable $b$. We can drop it, leaving a sum of $p$ independent 1-dimensional optimization problems.

Consider a fixed $j$, and let $l_j (t) = \frac{1}{2} t^2 - t z_j + \lambda |t|$, for $t \in \mathbb{R}$.
We analyze this in three cases because $|t|$ is *non-differentiable* at $t=0$.

- **Case 1:** If $t > 0$, then $|t| = t$ and $l_j (t) = \frac{1}{2} t^2 - t z_j + \lambda t$. Thus, $l_j' (t) = t - z_j + \lambda$.
  $$ l_j' (t) = 0 \iff t = z_j - \lambda \quad (1) $$
  Moreover, since we assumed $t > 0$, equation $(1)$ is a valid minimum if and only if $z_j > \lambda$.

- **Case 2:** If $t < 0$, then $|t| = - t$ and $l_j (t) = \frac{1}{2} t^2 - t z_j - \lambda t$. Thus, $l_j' (t) = t - z_j - \lambda$.
  $$ l_j' (t) = 0 \iff t = z_j + \lambda \quad (2) $$
  Moreover, since we assumed $t < 0$, equation $(2)$ is a valid minimum if and only if $z_j < -\lambda$.

- **Case 3:** If $|z_j| \le \lambda$, there is no valid solution where the derivative equals zero (neither Case 1 nor Case 2). Because the function is strictly convex, the global minimum must exist, which means it is trapped exactly at the non-differentiable corner: $t = 0$. We verify that $t = 0$ minimizes $l_j (t)$ (3).

For any $t \in \mathbb{R}$
$$
\begin{aligned}
l_j (t) - l_j (0) &= \frac{1}{2} t^2 - t z_j + \lambda |t| \\
                  &\ge \frac{1}{2} t^2 - |t| |z_j| + \lambda |t| \quad (-t z_j \ge -|t| |z_j|) \\
                  &\ge \frac{1}{2} t^2 - |t| (|z_j| - \lambda) \\
                  &\ge 0 \quad (|z_j| - \lambda \le 0)
\end{aligned}
$$
Therefore, $l_j (t) \ge l_j (0)$ for all $t \in \mathbb{R}$, proving $t=0$ is the global minimum for this range.

Combining these cases yields the soft-thresholding estimator for $\hat{b}_j$:
$$
\hat{b}_j = \begin{cases}
  z_j - \lambda & \text{if } z_j > \lambda, \\
  0             & \text{if } |z_j| \le \lambda, \\
  z_j + \lambda & \text{if } z_j < -\lambda
\end{cases}
$$

**Soft-thresholding operator under orthonormal design**
Assume that $X^T X = I_p$. Let $z = X^T y$. Then the Lasso minimizer of $L_\lambda (b)$ is <u>unique</u> and has coordinates:
$$
\hat{b}_j = S_\lambda (z_j), \quad \text{where the soft-thresholding function is } S_\lambda (z) = \text{sgn}(z) (|z| - \lambda)_+
$$
where $(u)_+ = \max(u, 0)$.

**Uniqueness.**
The uniqueness is guaranteed because the objective function is strictly convex. Under the orthogonal design ($\frac{1}{n} X^T X = I_p$), the variables are perfectly independent, forming a strictly convex "bowl" shape. Even with the sharp point added by the $L_1$ penalty, the strict convexity mathematically ensures there is exactly one absolute global minimum.

**Intuition.**
Under $X^T X = I_p$, we can compare the closed-form estimators
- **Ridge:** $\hat{b}_j^{\text{Ridge}} = z_j / (1 + \lambda)$
  - *Proportional Shrinkage:* It divides the coefficient by a constant. The coefficients get extremely small, but they never become zero.
- **Lasso:** $\hat{b}_j^{\text{Lasso}} = \text{sgn}(z_j) (|z_j| - \lambda)_+$
  - *Absolute Shrinkage:* It subtracts a constant from the magnitude. If a coefficient is originally small, the penalty forces it to exactly zero. This gives Lasso its *feature selection* capability.

## 3C. Connect algebra to evidence

*Connect one specific numerical result to Problem 3A or 3B.*

# Problem 4. Elastic net and a neural-feature bridge

## 4A. Research question and source map

**Research direction:** [weight decay / sparse weights / pruning / dropout]

| Claim | Exact primary source and locator | What it establishes | What it does not establish |
|---|---|---|---|
| [Claim] | [Citation, section/page/equation] | [Support] | [Boundary] |

## 4B. Elastic net on original predictors

```{r original-elastic-net, eval=FALSE}
alpha_grid <- c(0.10, 0.25, 0.50, 0.75, 0.90)

# Fit one cv.glmnet object per alpha with the same foldid.
# Select alpha and lambda using training CV only.
```

## 4C. Fixed ReLU features

```{r random-relu-features, eval=FALSE}
relu <- function(z) pmax(z, 0)
M <- 200L

# TODO:
# 1. Generate A and bias once from seeds$features using the stated distributions.
# 2. Apply the same A and bias to x_train and x_test.
# 3. Fit a scaler to H_train only and apply it to H_test.
# 4. Fit ridge, lasso, and elastic net on H_train with the shared foldid.
```

*State which weights are fixed, which are learned, and how active features are counted.*

## 4D. Locked declaration and final holdout evaluation

> **Locked before viewing `y_test`:** [preprocessing, candidate specifications, selected tuning values, feature seed, nominated deployment model, metrics]

```{r final-holdout-evaluation, eval=FALSE}
# This must be the first analysis chunk that uses y_test for evaluation.
# Generate all fixed candidate predictions and one RMSE/MAE table here.
```

*State whether the holdout evidence supports the predeclared choice. Do not retune.*

## 4E. Deep-learning connection and limitations

*Distinguish penalty, weight decay, output-weight sparsity, pruning, dropout, and a trained deep network. Cite the sources from 4A and discuss at least two limitations.*

# Problem 5. Report quality and reproducibility

## 5A. Reproduction record

```text
1. Open a clean R session.
2. [Exact commands and folder layout.]
3. Render GroupXX_ProjectQuiz1.Rmd.
```

## 5B. Recommendation and limitations

*Give the final recommendation, distinguish prediction from interpretation, and state limitations.*

## 5C. AI verification record

| AI-assisted item | How it was checked | Evidence | Modification or rejection |
|---|---|---|---|
| [Item] | [Method] | [Test/source/output] | [Decision] |

# Preflight and session information {-}

```{r preflight, eval=FALSE}
# Add every object required to reproduce the final table and figures.
required_objects <- c(
  "train_id", "test_id", "x_train", "x_test", "y_train", "y_test", "foldid"
)
missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects)) {
  stop(
    "Missing required objects: ",
    paste(missing_objects, collapse = ", ")
  )
}
stopifnot(length(intersect(train_id, test_id)) == 0L)
```

```{r session-info}
sessionInfo()
```
