---
title: "ASESII 24A02 Project-Based Quiz 1"
subtitle: "Ridge, Lasso, and Regularization for Neural Features"
author: |
  Le Dat Phuc An (24125052) \
  Pham Nguyen Minh Quan (24125041) \
  Tran Canh Anh Tuan (24125048) \
  Vong Chi Van (24125049)
  **Group 06**: \
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
bibliography: references.bib
link-citations: true
params:
  group_number: 6
fontsize: 11pt
geometry: margin=1in
editor_options:
  markdown:
    wrap: 72
---

# Authorship and contribution statement {.unnumbered}

We confirm that every group member can explain the submitted analysis.
AI use is reported in `Group06_AI_Log.csv`, and the group checked every
AI-assisted item used in this report.

| Student | ID | Main contributions | Verification performed |
|------------------|------------------:|------------------|------------------|
| Le Dat Phuc An | 24125052 | 1, 2A, 2B, 5A, and 5B | Verified 2D, 3B, and 4C by reproducing results and reviewing the report. |
| Pham Nguyen Minh Quan | 24125041 | 2D, 2E, 3A, 3B, and 4B | Verified 1, 2A, 5A, and 5B by checking methodology, code, and documentation. |
| Tran Canh Anh Tuan | 24125048 | 2A, 2B, 3C, 4A, 4C, 4E, 5A | Verified 2B, 2C, and 4B by reviewing implementation and numerical results. |
| Vong Chi Van | 24125049 | 2A, 2B, 2C, 4A, 4E | Verified Sections 2E, 3A, and 3C by checking derivations, code, and final outputs. |

# Abstract {.unnumbered}

This study investigates log prostate-specific antigen (`lpsa`)
prediction in 97 prostatectomy patients using clinical covariates. To
mitigate multicollinearity and prevent target leakage, we implement an
immutable 80/20 train/test partition with training-only preprocessing.
We compare Ordinary Least Squares (OLS) against regularized linear
models (Ridge, Lasso, Elastic Net) and a non-linear Random ReLU feature
representation ($M=200$). Mathematical derivations confirm that $L_2$
regularization stabilizes ill-conditioned Gram matrices by upwardly
shifting eigenvalues, reducing the condition number by over 50%
($\kappa$ dropped from 16.97 to 7.49). Empirical five-fold
cross-validation identifies Lasso (`lambda.min`) and Elastic Net as
superior predictors over baseline OLS and static non-linear projections.
For clinical deployment, we nominate the Elastic Net model on original
predictors, which optimizes out-of-sample generalization while
collaboratively shrinking correlated biomarkers. We emphasize that
regularized coefficients represent conditional predictive slopes rather
than causal biological mechanisms.

# Reproducible setup {.unnumbered}

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("tidyverse", "glmnet", "broom", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
    paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)

group_number <- as.integer(params$group_number)
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number,
  perturbation = 240500L + group_number
)
```

## Shared helper functions {.unnumbered}

```{r helpers, eval=TRUE}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(
    RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction))
  )
}

# counts variables (non-zero slopes) the model used (excluding intercept)
count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol, na.rm = TRUE)
}

# calculate L2 norm (excluding intercept)
get_l2_norm <- function(fit, s = "lambda.min") {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sqrt(sum(slopes^2, na.rm = TRUE))
}

# calculate L1 norm (excluding intercept)
get_l1_norm <- function(fit, s = "lambda.min") {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes), na.rm = TRUE)
}

# measure multicollinearity
safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  G <- crossprod(x) / nrow(x)
  eigenvalues <- eigen(G, symmetric = TRUE, only.values = TRUE)$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}

plot_model_result <- function(model, model_name) {
  model_result <- augment(model)
  ggplot(model_result, aes(x = .fitted, y = .resid)) +
    geom_point(color = "steelblue", size = 2, alpha = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
    theme_minimal() +
    labs(
      title = paste("Residual Plot Model:", model_name),
      subtitle = "Checking for non-linear patterns and constant variance",
      x = "Predicted lpsa (.fitted)",
      y = "Residuals (.resid)"
    )
}

vif_manual <- function(model) {
  X <- model.matrix(model)[, -1, drop = FALSE]
  sapply(seq_len(ncol(X)), function(i) {
    fit <- lm(X[, i] ~ X[, -i])
    1 / (1 - summary(fit)$r.squared)
  }) |> setNames(colnames(X))
}
```

# Problem 1. Prediction design and leakage control

## 1A. Target and predictors

```{r data-configuration, eval=TRUE}
config <- if (group_number %% 2L == 1L) {
  list(
    file = "fat.csv",
    response = "brozek",
    excluded = c("brozek", "siri", "density", "free")
  )
} else {
  list(
    file = "prostate.csv",
    response = "lpsa",
    excluded = "lpsa"
  )
}

data_raw <- read.csv(find_exam_data(config$file), check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}
```

### 1. Response

The natural Logarithm of Prostate-Specific Antigen is measured by $lpsa$,
which is a continuous number.

### 2. Candidate Predictors

|                      |                                  |                     |
|---------------------|-------------------------------|--------------------|
| Candidate Predictors | Full Name                        | Type                |
| lcavol               | Log Cancer Volume                | Continuous          |
| lweight              | Log Prostate Weight              | Continuous          |
| age                  | Age                              | Continuous          |
| lbph                 | Log Benign Prostatic Hyperplasia | Continuous          |
| svi                  | Seminal Vesicle Invasion         | Binary (0 / 1)      |
| lcp                  | Log Capsular Penetration         | Continuous          |
| gleason              | Gleason Score                    | Ordered Categorical |
| pgg45                | Percentage Gleason Grade 4/5     | Continuous (0–100)  |

### 3. Unit of Observation

Each row of the dataset contains the information of a radical
prostactomy treated patient which comprises of the target and the
predictors values.

### 4. Study Population

The dataset contains 97 patients aged 41 to 79 with clinically localized
adenocarcinoma who went through radical prostactomy at Stanford
University Center by Dr. Thomas A. Stamey and his surgical team during
the late 1980s.

### 5. Prediction Objective

To build a regression model that predicts the lpsa using the clinical variables (lcavol, lweight, age, lbph, svi, lcp, gleason, and pgg45).

### 6. Dataset Source

[@stamey1989]
**[Paper](https://sci-hub.sg/storage/2024/6429/4cbb75646440af614abced23a70ba059/stamey1989.pdf)**

### 7. Required Exclusions

The response variable `lpsa`, was excluded from the predictor matrix
because it is the outcome to be predicted. The remaining eight clinical
variables were retained as predictors. Since the dataset contains no
variables derived from `lpsa`, no additional exclusions were required.
This prevents target leakage and ensures a fair comparison among the
regression models. A predictor derived from `lpsa` would let the model exploit the test response during training, artificially inflating in-sample and out-of-sample metrics, making any model comparison meaningless.

## 1B. Split and train-only preprocessing

```{r split-and-scale, eval=TRUE}
split <- split_rows(nrow(analysis_data), seed = seeds$split)
train_id <- split$train
test_id <- split$test

x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_prep <- fit_scaler(x_raw[train_id, , drop = FALSE])
x_train <- apply_scaler(x_raw[train_id, , drop = FALSE], x_prep)
x_test <- apply_scaler(x_raw[test_id, , drop = FALSE], x_prep)
y_train <- y_raw[train_id]
y_test <- y_raw[test_id]

foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
```

### 1. Split seed

- Seed `split` for 80/20 split: 240206
- Seed `cv` for k-fold split: 240306

**Leakage Safeguard:** Setting deterministic random seeds prior to partitioning ensures the split is reproducible and objective. This prevents regenerating splits until a favorable test set is achieved.

### 2. Train/Test Split

The dataset of 97 observations was partitioned into:

- **Training Set:** 77 observations
- **Test Set:** 20 observations

**Leakage Safeguard:** Isolating the test set at the very beginning acts as the primary barrier against information leakage. It ensures that the holdout data remains completely unseen during subsequent exploratory analysis, feature engineering, and model tuning phases.

### 3. Feature Scaling Rule

Continuous predictor variables were centered and scaled using the helper functions `fit_scaler()` and `apply_scaler()`. The scaling parameters were estimated **only from the training data** in `fit_scaler()` and then applied unchanged to both the training and test sets by `apply_scaler()`.

Specifically, each predictor was transformed as:
$$x_{\text{scaled}} = \frac{x - \bar{x}_{\text{train}}}{\sqrt{\frac{1}{n_{\text{train}}} \sum_{i=1}^{n_{\text{train}}} (x_i - \bar{x}_{\text{train}})^2}}$$

where $\bar{x}_{\text{train}}$ is the mean of the predictor in the training set, and the denominator is the population standard deviation computed from the training observations.

**Leakage Safeguard:** Using the exact same scaling parameters (derived solely from the training data) to transform the test set prevents information leakage. If the entire dataset were scaled before splitting, the mean and variance of the holdout data would bleed into the training data.

### 4. Five-Fold Assignment Rule

A single, immutable five-fold cross-validation assignment vector (`foldid`) was generated on the 77 training observations using the designated `cv` random seed. To ensure fair and controlled comparisons, this exact same fold assignment was reused across all models evaluated:

- Ridge Regression
- Lasso Regression
- Elastic Net
- All subsequent hyperparameter grid evaluations

**Leakage Safeguard:** Generating the cross-validation folds strictly from the 77 training observations ensures the 20 test observations are excluded from the hyperparameter tuning process. Reusing the exact same `foldid` across all models ensures that model comparisons are fair and variance is minimized.

### 5. Missing-Value Plan

No missing values were detected. If any had been present, they would have been handled using a training-only imputation procedure before model fitting.

**Leakage Safeguard:** Calculating imputation statistics (like mean or median) strictly on the training set and then applying those exact values to fill `NA`s in the test set ensures that the statistical distribution of the holdout data does not influence the training data.

## 1C. Training-data exploration

The exploratory analysis in this section uses only the training
observations to avoid information leakage from the holdout test set.

### Response distribution

```{r response-distribution, eval=TRUE}
summary(y_train)
sd(y_train)
hist(y_train, main = "Histogram of y_train", xlab = "y_train")
qqnorm(y_train)
qqline(y_train)
```

The response variable (`lpsa`) has a mean of **2.55** and a median of
**2.59**, indicating an approximately symmetric distribution. Most
observations lie between **1.73** and **3.28**, with a standard deviation of **1.15**.

The histogram suggests that most observations are concentrated around
the center, while only a few observations appear in the lower and upper
tails. The normal Q–Q plot shows that most observations follow the
reference line reasonably well, with slight departures in the tails.
Overall, the response distribution is approximately normal, making it
suitable for linear regression.

### Relationship exploration

#### Scatterplot matrix

```{r scatter-matrix, eval=TRUE}
pairs(analysis_data[train_id, ], pch = 20, cex = 0.6)
```

The scatterplot matrix provides a visual overview of pairwise
relationships among the response and predictors. Several predictors
exhibit approximately linear relationships with the response,
particularly `lcavol`. Strong positive associations are also visible
between several predictor pairs, suggesting that some variables may
contain overlapping information.

#### Correlation matrix

```{r correlation-heatmap, eval=TRUE}
cor_matrix <- cor(analysis_data[train_id, ])

heatmap_df <- expand.grid(
  Var1 = rownames(cor_matrix),
  Var2 = colnames(cor_matrix)
)

heatmap_df$value <- c(cor_matrix)

ggplot(heatmap_df, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 3) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  coord_fixed() +
  theme_minimal()
```

The correlation heatmap shows that **lcavol** has the strongest linear
relationship with the response (*r* = 0.75), indicating that tumor
volume is the single most informative predictor of `lpsa`. Moderate
positive correlations are also observed for **lcp** (*r* = 0.59) and
**svi** (*r* = 0.58), whereas `age`, `lbph`, and `lweight` exhibit
relatively weak marginal correlations with the response.

### Potential multicollinearity

The correlation matrix indicates several moderately to strongly
correlated predictor pairs.

The strongest pairwise correlations include

- `gleason` and `pgg45` (**r = 0.76**),
- `lcavol` and `lcp` (**r = 0.73**),
- `lcp` and `svi` (**r = 0.67**),
- `lcavol` and `svi` (**r = 0.59**),
- `lcavol` and `gleason` (**r = 0.47**).

These relationships suggest that some predictors may contain overlapping
information. However, pairwise correlation alone does not imply severe
multicollinearity, since multicollinearity depends on the joint linear
relationship among all predictors rather than individual pairs.
Therefore, formal diagnostics such as the variance inflation factor
(VIF) will be performed after fitting the OLS model to determine whether
coefficient instability is present.

### One possible anomaly or influential observation

```{r boxplot, eval=TRUE}
boxplot(analysis_data[train_id, ])
```

The boxplots identify several potential outliers, particularly in
**age** and **pgg45**. Among the predictors, `pgg45` exhibits the
largest variability, while `age` contains several observations outside
the whiskers. These observations are retained because there is no
evidence that they result from data entry errors.

### Question for regularized models

Based on the exploratory analysis, the following question will guide the
remaining analysis

> Can regularization improve predictive performance and coefficient
> stability when predictors contain overlapping information?

# Problem 2. OLS, Ridge, and Lasso

## 2A. OLS baseline

The ordinary least squares (OLS) model was fitted using only the
standardized training observations. The holdout
test set remained untouched to ensure a fair comparison with
regularized regression models.

### Fit OLS

```{r ols-fit, eval=TRUE}
train_df <- as.data.frame(x_train)
train_df$lpsa <- y_train

ols_model <- lm(lpsa ~ ., data = train_df)
```

This serves as the baseline model for all subsequent
comparisons with Ridge and Lasso regression.

### Coefficient estimates

```{r ols-summary, eval=TRUE}
summary(ols_model)
```

The fitted OLS model is statistically significant overall (**F =
15.78**, p \< 0.001), indicating that at least one predictor contributes
to explaining the variation in the response variable.

The model achieves a **Multiple** $R^2$ of 0.6500 and an **Adjusted**
$R^2$ of 0.6088, suggesting that approximately **65%** of the
variability in `lpsa` is explained by the eight clinical predictors.

Among the predictors,

- **lcavol** ($\beta=0.674$, p \< 0.001),
- **lweight** ($\beta=0.202$, p = 0.0438),
- **age** ($\beta=-0.194$, p = 0.0435), and
- **svi** ($\beta=0.282$, p = 0.0206)

are statistically significant at the 5% significance level.

The remaining predictors (`lbph`, `lcp`, `gleason`, and `pgg45`) are not
individually significant after adjusting for the other predictors in the
model.

The estimated coefficient for **age** is negative ($\beta=-0.194$). This
result suggests that the estimated effect of `age` changes after
accounting for the remaining predictors in the model. The possibility of
correlation among predictors is therefore investigated using VIF in the following subsection.

### Training error

```{r ols-training-error, eval=TRUE}
train_pred <- predict(ols_model)

train_mse <- mean((y_train - train_pred)^2)
train_rmse <- sqrt(train_mse)
cat("Training MSE :", round(train_mse, 4), "\n")
cat("Training RMSE:", round(train_rmse, 4), "\n")
```

The fitted OLS model achieved

- **Training MSE = 0.4544**
- **Training RMSE = 0.6741**

These quantities measure the goodness-of-fit on the training
observations used to estimate the regression coefficients. Since the
model is evaluated on the same data used for fitting, the training error
is expected to underestimate the prediction error on unseen
observations.

### Five-fold cross-validation

```{r ols-cv, eval=TRUE}
cv_mse <- numeric(5)
for (i in 1:5) {
  train_fold <- train_df[foldid != i, ]
  valid_fold <- train_df[foldid == i, ]
  model <- lm(lpsa ~ ., data = train_fold)

  pred <- predict(model, newdata = valid_fold)
  cv_mse[i] <- mean((valid_fold$lpsa - pred)^2)
}
cv_table <- data.frame(
  Fold = 1:5,
  Validation_MSE = round(cv_mse, 4)
)

knitr::kable(cv_table)
cat("Mean CV MSE :", round(mean(cv_mse), 4), "\n")
cat("Mean CV RMSE:", round(sqrt(mean(cv_mse)), 4), "\n")
```

The validation errors obtained from the five folds are

| Fold | Validation MSE |
|-----:|---------------:|
|    1 |         0.8259 |
|    2 |         0.8568 |
|    3 |         0.4011 |
|    4 |         0.4893 |
|    5 |         0.9002 |

The average cross-validation error is

- **Mean CV MSE = 0.6947**
- **Mean CV RMSE = 0.8335**

As expected, the cross-validation error is larger than the training
error because each validation fold consists of observations that were
not used to estimate the regression coefficients. Although some
variation exists among the five folds, the validation errors remain
within a similar range, suggesting that the predictive performance of
the OLS model is reasonably stable across different training subsets.

### Residual diagnostics

#### Residual versus fitted values

```{r ols-residual, eval=TRUE}
plot(
  fitted(ols_model),
  resid(ols_model),
  xlab = "Fitted values",
  ylab = "Residuals",
  main = "Residuals vs Fitted"
)
abline(h = 0, col = "red")
```

The residual-versus-fitted plot shows that the residuals are randomly
scattered around zero without noticeable curvature or funnel-shaped
patterns. This suggests that the assumptions of

- linearity, and
- constant error variance (homoscedasticity)

are reasonably satisfied. No obvious pattern indicating model
misspecification is observed.

#### Normal Q-Q plot

```{r ols-qqplot, eval=TRUE}
qqnorm(resid(ols_model))
qqline(resid(ols_model), col = "red")
```

The normal Q–Q plot indicates that the residuals are approximately
normally distributed, with slight deviations in the tails. This
behavior is consistent with the mildly skewed response distribution
observed. Since the residuals satisfy

$$\hat{\mathbf{e}} = (I-H) \mathbf{y}$$

where $\hat{\mathbf{e}}$ is the residual vector, $\mathbf{y}$ is the
response vector, $I$ is the identity matrix, and
$H = X(X^\top X)^{-1}X^\top$ is the hat matrix, their distribution
naturally reflects the response after removing the fitted linear
component. Overall, there is no strong evidence against the normality
assumption.

#### Influence plot

```{r ols-influence, eval=TRUE}
plot(ols_model, which = 5)
```

Most observations have low leverage and moderate residuals. Although a
few observations (e.g., 23, 37, and 54) stand out slightly, none appears
to have an unusually large influence on the fitted model.

#### Cook's distance

```{r ols-cook-distance, eval=TRUE}
cook <- cooks.distance(ols_model)
plot(
  cook,
  type = "h",
  lwd = 3,
  col = "red",
  xlab = "Observation",
  ylab = "Cook's Distance",
  main = "Cook's Distance"
)
abline(h = 4 / nrow(train_df), col = "blue", lty = 2)
```

Several observations exceed the reference line $4/n$, but all Cook's
distance values are well below 1. Therefore, there is no strong evidence
of influential observations.

Taken together, the residual diagnostics do not reveal any serious
violation of the standard assumptions underlying OLS
regression.

### Ill-conditioned design

```{r ols-vif, eval=TRUE}
vif_manual(ols_model)
```

The VIF values were computed to assess multicollinearity among the predictors . They range from **1.33** to **3.24**, with the largest values
observed for

- `lcp` (3.24),
- `pgg45` (2.97),
- `gleason` (2.68),
- `lcavol` (2.42).

None of the predictors exceeds the commonly used threshold of **5**,
indicating that severe multicollinearity is not present in the training
design matrix.

This finding is consistent with the correlation analysis in
Section 1C. Although several predictor pairs, such as `gleason`–`pgg45`
and `lcavol`–`lcp`, exhibit moderate pairwise correlations, it alone does not necessarily imply problematic
multicollinearity. VIF measures how well each predictor can be
explained by **all remaining predictors simultaneously**, making it a
more appropriate diagnostic for coefficient instability.

Overall, the VIF results suggest that the correlations among predictors
are not sufficiently strong to produce substantial inflation of the
coefficient variances. Therefore, no clear evidence of an
ill-conditioned design was found, and the OLS provide an
appropriate baseline for comparison with Ridge and Lasso regression in
the following sections.

## 2B. Ridge

### Fitting Ridge

```{r ridge, eval=TRUE}
ridge_model <- fit_cv_glmnet(x_train, y_train, alpha = 0, foldid = foldid)

plot(ridge_model)
abline(v = -log(ridge_model$lambda.min), col = "purple", lty = 2, lwd = 1.5)
abline(v = -log(ridge_model$lambda.1se), col = "blue", lty = 2, lwd = 1.5)

y_pos <- max(ridge_model$cvup)
text(-log(ridge_model$lambda.min), y_pos, "lambda.min",
  pos = 2, col = "purple", cex = 0.9
)
text(-log(ridge_model$lambda.1se), y_pos, "lambda.1se",
  pos = 2, col = "blue", cex = 0.9
)
```

The cross-validation curve shows how the prediction error changes with
the penalty parameter. The mean cross-validation error decreases until
reaching its minimum at **`lambda.min`**. Around the minimum, the error
changes only slightly, indicating that a range of penalty values
provides similar predictive performance.

Because Ridge regression uses an L2 penalty, every coefficient is
continuously shrunk toward zero as the penalty increases. However, Ridge
does not apply coordinate-wise thresholding, so no coefficient is set
exactly to zero.

### Investigation on `lambda.min` and `lambda.1se`

`lambda.min` is the value of $\lambda$ that minimizes the mean
cross-validation error. It is selected when predictive accuracy is the
primary objective.

```{r, eval=TRUE}
cat(sprintf("lambda.min: %.5f\n", ridge_model$lambda.min))
min_id <- which(ridge_model$lambda == ridge_model$lambda.min)
cat(sprintf("cvm.min: %.5f\n", ridge_model$cvm[min_id]))
coef(ridge_model, s = "lambda.min")
```

The coefficients are moderately shrunk compared with the OLS estimates,
while the most important predictors still retain relatively large
coefficients.

`lambda.1se` is the largest value of $\lambda$ whose cross-validation
error is within one standard error of the minimum. It applies stronger
regularization, producing a simpler and more stable model while usually
maintaining similar predictive performance.

```{r, eval=TRUE}
cat(sprintf("lambda.1se: %.5f\n", ridge_model$lambda.1se))
se1_id <- which(ridge_model$lambda == ridge_model$lambda.1se)
cat(sprintf("cvm.1se: %.5f\n", ridge_model$cvm[se1_id]))
coef(ridge_model, s = "lambda.1se")
```

Using the larger penalty `lambda.1se` shrinks every coefficient further
toward zero than `lambda.min`. The cross-validation error increases only
slightly, indicating that a more regularized model can achieve
nearly the same predictive performance.

Unlike Lasso, Ridge regression continuously shrinks the coefficients but
does not set any coefficient exactly to zero. Therefore, every predictor
remains in the fitted model under both tuning parameter choices.

### Coefficient paths

The coefficient paths show how the Ridge coefficients change as the
penalty varies.

```{r, eval=TRUE}
plot(ridge_model$glmnet.fit, xvar = "lambda", label = TRUE)
title("Ridge Coefficient Paths", line = 2.5)

abline(v = -log(ridge_model$lambda.min), col = "green", lty = 2, lwd = 1.5)
abline(v = -log(ridge_model$lambda.1se), col = "blue", lty = 2, lwd = 1.5)

legend("topleft",
  legend = c("lambda.min", "lambda.1se"),
  col = c("green", "blue"),
  lty = 2,
  lwd = 1.5,
  cex = 0.8,
  bg = "white"
)
predictor_names <- rownames(ridge_model$glmnet.fit$beta)
legend("topleft",
  inset = c(0, 0.21),
  legend = paste(1:length(predictor_names), predictor_names, sep = " - "),
  col = 1:length(predictor_names),
  lty = 1,
  lwd = 1.5,
  cex = 0.8,
  bg = "white",
  ncol = 2
)
```

As the penalty decreases (left to right), the coefficients
move smoothly away from near zero toward their OLS estimates. Larger
penalties produce stronger shrinkage, whereas smaller penalties allow
the coefficients to approach the unregularized solution.

### General behavior of correlated predictors

Ridge regression is particularly effective when predictors are highly
correlated. In OLS, correlated predictors compete to
explain the same variation in the response, leading to unstable
coefficient estimates with high variance.

Ridge addresses this issue by shrinking the coefficients of correlated
predictors towards each other rather than allowing one predictor to
dominate the others. This reduces the variability of the coefficient
estimates and improves model stability while retaining all predictors in
the model. This behaviour is reflected in the coefficient paths, where
all coefficients are smoothly shrunk towards zero as the penalty
increases.

#### Examples from the data

The effect of Ridge can be seen in several pairs of highly correlated
predictors.

- **`lcavol` and `lcp` (correlation = 0.73):** Under OLS, `lcavol`
  receives a much larger positive coefficient while `lcp` is assigned a
  small negative coefficient. Ridge shrinks both coefficients, allowing the two correlated predictors to share their predictive contribution.

- **`pgg45` and `gleason` (correlation = 0.76):** OLS assigns
  substantially different coefficients to these correlated predictors.
  Ridge shrinks both estimates towards zero, producing more stable
  coefficient estimates while retaining both variables in the model.

## 2C. Lasso

### Fitting Lasso

```{r lasso, eval=TRUE}
lasso_model <- fit_cv_glmnet(x_train, y_train, alpha = 1, foldid = foldid)

plot(lasso_model)
abline(v = -log(lasso_model$lambda.min), col = "purple", lty = 2, lwd = 1.5)
abline(v = -log(lasso_model$lambda.1se), col = "blue", lty = 2, lwd = 1.5)

y_pos <- max(lasso_model$cvup)
text(-log(lasso_model$lambda.min), y_pos,
  labels = "lambda.min",
  pos = 2, col = "purple", cex = 0.9
)

text(-log(lasso_model$lambda.1se), y_pos,
  labels = "lambda.1se",
  pos = 2, col = "blue", cex = 0.9
)
```

Unlike Ridge, Lasso performs both coefficient shrinkage and
variable selection. As the penalty increases, some coefficients are
shrunk to zero, producing a simpler and more interpretable
model. The cross-validation curve identifies the penalty giving the
lowest prediction error (`lambda.min`) together with the largest penalty
within one standard error (`lambda.1se`).

### Investigation on `lambda.min` and `lambda.1se`

The optimal penalties and corresponding training cross-validation errors are

```{r, eval=TRUE}
cat(sprintf("lambda.min: %.5f\n", lasso_model$lambda.min))
min_id <- which(lasso_model$lambda == lasso_model$lambda.min)
cat(sprintf("cvm.min : %.5f\n", lasso_model$cvm[min_id]))

cat(sprintf("lambda.1se: %.5f\n", lasso_model$lambda.1se))
se1_id <- which(lasso_model$lambda == lasso_model$lambda.1se)
cat(sprintf("cvm.1se : %.5f\n", lasso_model$cvm[se1_id]))
```

The selected predictors under the two tuning rules are

```{r, eval=TRUE}
cat("--- Nonzero coefficients (lambda.min) ---\n")
coef_min <- coef(lasso_model, s = "lambda.min")
print(coef_min[coef_min[, 1] != 0, ])

cat("\n--- Nonzero coefficients (lambda.1se) ---\n")
coef_1se <- coef(lasso_model, s = "lambda.1se")
print(coef_1se[coef_1se[, 1] != 0, ])
```

`lambda.min` prioritizes predictive accuracy and therefore usually keeps
more predictors. `lambda.1se` accepts a small increase in prediction
error in exchange for a sparser and easier-to-interpret model.

### Coefficient paths

```{r, eval=TRUE}
plot(lasso_model$glmnet.fit, xvar = "lambda", label = TRUE)
title("Lasso Coefficient Paths", line = 2.5)

abline(v = -log(lasso_model$lambda.min), col = "purple", lty = 2, lwd = 1.5)
abline(v = -log(lasso_model$lambda.1se), col = "blue", lty = 2, lwd = 1.5)

legend("topleft",
  legend = c("lambda.min", "lambda.1se"),
  col = c("purple", "blue"),
  lty = 2,
  lwd = 1.5,
  cex = 0.75,
  bg = "white"
)
```

Unlike Ridge, Lasso coefficient paths frequently terminate at zero. Variables that reach zero are removed from the model,
while only the strongest predictors remain active under larger penalty
values.

### Application-specific interpretation

Lasso identifies the predictors that provide the strongest independent
contribution to explaining `lpsa` after accounting for the remaining
variables.

At `lambda.min`, four predictors remain active: `lcavol`, `lweight`,
`svi`, and `pgg45`. Among these, `lcavol` has by far the largest
coefficient (0.607), indicating that tumor volume is the strongest
predictor of prostate-specific antigen levels. `svi` also remains,
suggesting that seminal vesicle invasion contributes additional
predictive information beyond tumor volume. `lweight` and `pgg45` have
smaller coefficients but still improve predictive performance enough to
be retained.

At `lambda.1se`, the stronger penalty removes both `lweight` and
`pgg45`, leaving only `lcavol` and `svi`. This indicates that these two
variables contain the most robust independent signal in the data. The
cross-validation error increases only slightly (0.632 to 0.728), while
the model is reduced from four predictors to two, yielding a much
simpler and more interpretable model.

The variables excluded by Lasso are not necessarily unrelated
to prostate cancer. Rather, after accounting for `lcavol` and `svi`,
their remaining predictive contribution is too small to justify the
additional model complexity. Some of their information is likely shared
with the retained predictors because several clinical measurements are
moderately correlated.

## 2D. Training-based comparison and locked core model

This section compares candidate models using only the training data
and cross-validation results. The holdout test set is not used.

### Training-based comparison

```{r core-comparison, eval=TRUE}
# OLS
cv_ols <- mean(cv_mse)
l2_ols <- get_l2_norm(ols_model)
l1_ols <- get_l1_norm(ols_model)
nz_ols <- count_nonzero(ols_model)

# Ridge
id_min <- which(ridge_model$lambda == ridge_model$lambda.min)
id_1se <- which(ridge_model$lambda == ridge_model$lambda.1se)

cv_ridge_min <- ridge_model$cvm[id_min]
cv_ridge_1se <- ridge_model$cvm[id_1se]

l2_ridge_min <- get_l2_norm(ridge_model, s = "lambda.min")
l2_ridge_1se <- get_l2_norm(ridge_model, s = "lambda.1se")
l1_ridge_min <- get_l1_norm(ridge_model, s = "lambda.min")
l1_ridge_1se <- get_l1_norm(ridge_model, s = "lambda.1se")

nz_ridge_min <- count_nonzero(ridge_model, s = "lambda.min")
nz_ridge_1se <- count_nonzero(ridge_model, s = "lambda.1se")

# Lasso
id_min <- which(lasso_model$lambda == lasso_model$lambda.min)
id_1se <- which(lasso_model$lambda == lasso_model$lambda.1se)

cv_lasso_min <- lasso_model$cvm[id_min]
cv_lasso_1se <- lasso_model$cvm[id_1se]

l2_lasso_min <- get_l2_norm(lasso_model, s = "lambda.min")
l2_lasso_1se <- get_l2_norm(lasso_model, s = "lambda.1se")
l1_lasso_min <- get_l1_norm(lasso_model, s = "lambda.min")
l1_lasso_1se <- get_l1_norm(lasso_model, s = "lambda.1se")

nz_lasso_min <- count_nonzero(lasso_model, s = "lambda.min")
nz_lasso_1se <- count_nonzero(lasso_model, s = "lambda.1se")

comparison_table <- data.frame(
  Model = c("OLS", "Ridge", "Ridge", "Lasso", "Lasso"),
  Tuning = c("None", "lambda.min", "lambda.1se", "lambda.min", "lambda.1se"),
  CV_MSE = c(cv_ols, cv_ridge_min, cv_ridge_1se, cv_lasso_min, cv_lasso_1se),
  L2_Norm = c(l2_ols, l2_ridge_min, l2_ridge_1se, l2_lasso_min, l2_lasso_1se),
  L1_Norm = c(l1_ols, l1_ridge_min, l1_ridge_1se, l1_lasso_min, l1_lasso_1se),
  Nonzero_Slopes = c(nz_ols, nz_ridge_min, nz_ridge_1se, nz_lasso_min, nz_lasso_1se)
)

knitr::kable(comparison_table, digits = 4)
```

Regularization improves predictive performance over OLS, with both Ridge
(`lambda.min`) and Lasso (`lambda.min`) achieving lower cross-validation
errors. Ridge shrinks the coefficients but retains all eight predictors,
whereas Lasso also performs variable selection. Lasso (`lambda.min`)
achieves the lowest cross-validation error while retaining only four
predictors, providing the best balance between prediction accuracy and
model simplicity.

Note on coefficient norms: Both the $L_2$ and $L_1$ norms of the coefficient vectors (excluding the intercept) are reported. While Ridge explicitly penalizes the $L_2$ norm and Lasso penalizes the $L_1$ norm, tracking both metrics across all models provides a complete picture of how each regularization method shrinks the overall coefficient magnitude versus enforcing sparsity.

### Core model nominated before holdout evaluation

Based on the training results, **Lasso with `lambda.min`** is selected
as the core candidate. It achieves the lowest cross-validation error
while reducing the model from eight predictors to four. Although Ridge has similar predictive performance, it retains all predictors and is therefore less interpretable.


### Why those lambdas answer different decision questions

`lambda.min` prioritizes predictive performance by minimizing the
cross-validation error. In contrast, `lambda.1se` selects a larger
penalty to obtain a simpler and more stable model while accepting a
small increase in prediction error.

Predictive performance is assessed by the cross-validation error, whereas interpretability depends on model simplicity. In this analysis, Lasso (`lambda.min`) achieves the lowest prediction error while retaining only four predictors, making it the preferred model before the
final holdout evaluation.

## 2E. Perturbation or stability check

This investigates how Ridge and Lasso respond to severe
multicollinearity by duplicating a predictor with a small amount of
random noise. The experiment uses only the training data.

### Prediction before running the check

Before running the perturbation, we expect Ridge to distribute the
coefficient between the original and duplicated predictors because the
Ridge penalty encourages correlated predictors to share their effects.
In contrast, Lasso is expected to retain only one of the two predictors
while shrinking the other coefficient to zero.

```{r perturbation, eval=TRUE}
set.seed(seeds$perturbation)

# duplicate a strong predictor with small random noise
target_idx <- 1
target_name <- colnames(x_train)[target_idx]

duplicate_var <- x_train[, target_idx] + rnorm(nrow(x_train), sd = 0.01)
x_train_pert <- cbind(x_train, duplicate_var = duplicate_var)

# refit Ridge and Lasso using the perturbed data
ridge_pert <- fit_cv_glmnet(x_train_pert, y_train, alpha = 0, foldid = foldid)
lasso_pert <- fit_cv_glmnet(x_train_pert, y_train, alpha = 1, foldid = foldid)

ridge_coef <- coef(ridge_pert, s = "lambda.1se")
lasso_coef <- coef(lasso_pert, s = "lambda.1se")

cat("Ridge:\n")
print(ridge_coef[c(target_name, "duplicate_var"), , drop = FALSE])
cat("\nLasso:\n")
print(lasso_coef[c(target_name, "duplicate_var"), , drop = FALSE])
```

### Observed result and explanation

The perturbation confirms the expected behaviour of the two
regularization methods.

Ridge assigns almost identical coefficients to the original predictor
(`lcavol` = 0.1560) and its duplicate (0.1561), indicating that the
effect is shared between highly correlated predictors.

In contrast, Lasso retains the original predictor (`lcavol` = 0.4918)
while shrinking the duplicated predictor exactly to zero. This
demonstrates that Lasso performs variable selection by choosing one
representative predictor and discarding redundant correlated variables.

Overall, the perturbation illustrates the different mechanisms of Ridge
and Lasso. Ridge stabilizes coefficient estimates by sharing information
across correlated predictors, whereas Lasso produces a sparse and more
interpretable model by selecting only one predictor from a correlated
group.

# Problem 3. Mathematical mechanisms

## 3A. Ridge and conditioning

Let
$X \in \mathbb{R}^{n \times p}, y \in \mathbb{R}^n, b \in \mathbb{R}^p$
and define $$G = \frac{1}{n} X^\top X, \quad
z = \frac{1}{n} X^\top y$$ consequently, $G \in \mathbb{R}^{p \times p}$
and $z \in \mathbb{R}^p$.

From the objective function
$$Q_\lambda(b) = \frac{1}{2n}\|y - Xb\|_2^2 + \frac{\lambda}{2}\|b\|_2^2$$

We expand Ridge objective function
$$
\begin{align}
  Q_\lambda (b) &= \frac{1}{2n} \|y - X b\|_2^2 + \frac{\lambda}{2} \|b\|_2^2, \quad (\lambda > 0) \\
  &= \frac{1}{2n} (y - X b)^\top (y - X b) + \frac{\lambda}{2} b^\top b \\
  &= \frac{1}{2n} (y^\top y - y^\top X b - b^\top X^\top y + b^\top X^\top X b) + \frac{\lambda}{2} b^\top b \\
  &= \frac{1}{2n} (y^\top y - 2 y^\top X b + b^\top X^\top X b) + \frac{\lambda}{2} b^\top b \\
\end{align}
$$

We take the gradient with respect to $b$ and substitute
$$
\begin{align}
  \nabla_b Q_\lambda (b) &= 0 - \frac{1}{n} X^\top y + \frac{1}{n} X^\top X b + \lambda b \\
  &= - z + G b + \lambda b
\end{align}
$$

Set $\nabla_b Q_\lambda (b) = 0 \Leftrightarrow (G + \lambda I_p) b = z$
which is the $\textbf{normal equation}$.

Because $G = \frac{1}{n}X^\top X$ is a positive semi-definite Gram
matrix, all its eigenvalues are $\ge 0$. Adding a penalty $\lambda > 0$
along the diagonal (i.e., $\lambda I_p$) shifts all eigenvalues up by
$\lambda$. Thus, the matrix $(G + \lambda I_p)$ is strictly positive
definite, making it full rank and invertible. This guarantees a unique minimizer

$$\hat{b}_\lambda^{\text{ridge}} = (G + \lambda I_p)^{-1} z$$

```{r ridge-conditioning, eval=TRUE}
condition_numbers <- safe_condition_numbers(x_train, ridge_model$lambda.min)
kappa_G <- condition_numbers["gram"]
kappa_ridge <- condition_numbers["ridge"]

cat("Condition number of G:", kappa_G, "\n")
cat("Condition number of Ridge:", kappa_ridge, "\n")
```

### Numerical interpretation

The OLS estimator is $\hat b=(X^\top X)^{-1}X^\top y$ so its numerical stability depends on the Gram matrix $G= \frac{1}{n} X^\top X$.

When predictors are highly correlated, the columns of $X$ become nearly
linearly dependent, making $\det(G)\approx 0$. Equivalently,

$$
G=Q\Lambda Q^\top,
\qquad
\Lambda=\operatorname{diag}(\mu_1,\ldots,\mu_p)
$$

contains one or more very small eigenvalues. Since

$$
G^{-1}
=
Q\Lambda^{-1}Q^\top,
\qquad
\Lambda^{-1}
=
\operatorname{diag}
\left(
\frac1{\mu_1},
\ldots,
\frac1{\mu_p}
\right)
$$

A small eigenvalue $\mu_{\min}\approx 0$ produces a very large reciprocal $\frac{1}{\mu_{\min}} \rightarrow \infty$. Therefore, the inverse $(X^\top X)^{-1}$ amplifies small perturbations in the data, making the coefficient estimate $\hat b$ highly sensitive to sampling variation and numerical error. This instability is summarized by the spectral condition number

$$
\kappa_2(G)
=
\frac{\mu_{\max}}
{\mu_{\min}}
$$

where a large value indicates an ill-conditioned matrix.

Ridge replaces $G$ by $G+\lambda I$ shifts every eigenvalue by $\mu_i \longrightarrow \mu_i+\lambda$. Hence,

$$
\kappa_2(G+\lambda I)
=
\frac{\mu_{\max}+\lambda}
{\mu_{\min}+\lambda}
$$

Since the smallest eigenvalue increases the most in relative terms, the
condition number decreases, making the matrix inversion and coefficient
estimates substantially more stable.

### Geometric interpretation

The least-squares objective is

$$RSS(b)=\|y-Xb\|^2$$

whose Hessian is $\nabla^2RSS(b)=2X^\top X=2G$.

Thus, the eigenvalues of $G$ determine the curvature of the loss
surface. When one eigenvalue is much smaller than the others, the
curvature in that direction is very weak, producing long, narrow
elliptical contours. Along this nearly flat valley, many coefficient
vectors produce almost identical values of $RSS(b)$, so a small
perturbation in the data can move the optimum a large distance, leading
to unstable coefficients.

Ridge adds the penalty

$$\lambda\|b\|_2^2$$

whose Hessian is $2\lambda I$. Therefore, the penalized Hessian becomes

$$2(G+\lambda I)$$

increasing the curvature in every direction, especially those
corresponding to the smallest eigenvalues. The elongated ellipses become
more circular, the optimum is better localized, and the coefficient
estimates become more stable while retaining all predictors.

## 3B. Lasso optimality and soft thresholding

Let
$X \in \mathbb{R}^{n \times p}, y \in \mathbb{R}^n, b \in \mathbb{R}^p$
and define $$G = \frac{1}{n} X^\top X = I_p, \quad
z = \frac{1}{n} X^\top y$$ consequently, $G \in \mathbb{R}^{p \times p}$
and $z \in \mathbb{R}^p$.

The objective function
$$R_\lambda (b) = \frac{1}{2n} \|y - X b\|_2^2 + \lambda \|b\|_1, \quad \|b\|_1 = \sum_{j=1}^p |b_j|, \lambda > 0$$

We expand Lasso objective function $$\begin{aligned}
R_\lambda (b) &= \frac{1}{2n} \|y - X b\|_2^2 + \lambda \|b\|_1 \\
              &= \frac{1}{2n} (y - X b)^\top (y - X b) + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} (y^\top y - y^\top X b - b^\top X^\top y + b^\top X^\top X b) + \lambda \sum_{j=1}^p |b_j|
\end{aligned}$$

The term $y^\top X b$ is is a scalar, therefore
$(y^\top X b)^\top = b^\top X^\top (y^\top)^\top = b^\top X^\top y$.

Then the objective becomes $$\begin{aligned}
R_\lambda (b) &= \frac{1}{2n} y^\top y - \frac{1}{n} b^\top X^\top y + \frac{1}{2n} b^\top X^\top X b + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} y^\top y - b^\top z + \frac{1}{2} b^\top b + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} y^\top y + \sum_{j=1}^p \left( - b_j z_j + \frac{1}{2} b_j^2 + \lambda |b_j| \right)
\end{aligned}$$

The term $\frac{1}{2n} y^\top y$ is a constant with respect to the
variable $b$, leaving a sum of $p$ independent 1-dimensional
optimization problems.

Consider a fixed $j$, and let
$l_j (t) = \frac{1}{2} t^2 - t z_j + \lambda |t|$, for
$t \in \mathbb{R}$. We analyze this in three cases because $|t|$ is
*non-differentiable* at $t=0$.

- **Case 1:** If $t > 0$, then $|t| = t$ and
  $l_j (t) = \frac{1}{2} t^2 - t z_j + \lambda t$. Thus,
  $l_j' (t) = t - z_j + \lambda$.
  $$ l_j' (t) = 0 \iff t = z_j - \lambda \quad (1) $$ Moreover, since we
  assumed $t > 0$, equation $(1)$ is a valid minimum if and only if
  $z_j > \lambda$.

- **Case 2:** If $t < 0$, then $|t| = - t$ and
  $l_j (t) = \frac{1}{2} t^2 - t z_j - \lambda t$. Thus,
  $l_j' (t) = t - z_j - \lambda$.
  $$ l_j' (t) = 0 \iff t = z_j + \lambda \quad (2) $$ Moreover, since we
  assumed $t < 0$, equation $(2)$ is a valid minimum if and only if
  $z_j < -\lambda$.

- **Case 3:** If $|z_j| \le \lambda$, there is no valid solution where
  the derivative equals zero (neither Case 1 nor Case 2). Because the
  function is strictly convex, the global minimum must exist, which
  means it is trapped exactly at the non-differentiable corner: $t = 0$.
  We verify that $t = 0$ minimizes $l_j (t)$ (3).

For any $t \in \mathbb{R}$
$$\begin{aligned}
l_j (t) - l_j (0) &= \frac{1}{2} t^2 - t z_j + \lambda |t| \\
                  &\ge \frac{1}{2} t^2 - |t| |z_j| + \lambda |t| \quad (-t z_j \ge -|t| |z_j|) \\
                  &\ge \frac{1}{2} t^2 - |t| (|z_j| - \lambda) \\
                  &\ge 0 \quad (|z_j| - \lambda \le 0)
\end{aligned}$$
Therefore, $l_j (t) \ge l_j (0)$ for all
$t \in \mathbb{R}$, proving $t=0$ is the global minimum for this range.

Combining these cases yields the soft-thresholding estimator for
$\hat{b}_j$: $$\hat{b}_j = \begin{cases}
  z_j - \lambda & \text{if } z_j > \lambda, \\
  0             & \text{if } |z_j| \le \lambda, \\
  z_j + \lambda & \text{if } z_j < -\lambda
\end{cases}$$

The uniqueness is guaranteed because the objective function is strictly
convex. Under the orthogonal design, the variables are perfectly
independent, forming a strictly convex "bowl" shape. Even with the sharp
point added by the $L_1$ penalty, the strict convexity mathematically
ensures there is exactly one absolute global minimum.

**Soft-thresholding operator under orthonormal design** Assume that
$\frac{1}{n} X^T X = I_p$. Let $z = \frac{1}{n} X^T y$. Then the Lasso
minimizer of $R_\lambda (b)$ is *unique* and has coordinates
$$\hat{b}_j = S_\lambda (z_j), \quad \text{where the soft-thresholding function is } S_\lambda (z) = \text{sgn}(z) (|z| - \lambda)_+$$
where $(u)_+ = \max(u, 0)$.

**Why Lasso can produce exact zeros while Ridge generally does not.**
Under $\frac{1}{n} X^T X = I_p$, we can compare the closed-form
estimators

- **Ridge:** $\hat{b}_j^{\text{Ridge}} = (1 +  \lambda)^{-1} z_j$
  (proportional shrinkage). It divides the coefficient by a constant.
  The coefficients get extremely small, but they never become zero. The
  optimality condition means $b_j$ only equals exactly zero if $z_j$ is
  exactly zero (which is highly improbable with real data).
- **Lasso:**
  $\hat{b}_j^{\text{Lasso}} = \text{sgn}(z_j) (|z_j| - \lambda)_+$
  (absolute shrinkage). It subtracts a constant from the magnitude. If a
  coefficient is originally small, the penalty forces it to exactly
  zero. Case 3 shows that Lasso maps a whole range of values to exactly
  zero, creating sparsity. This gives Lasso its **feature selection**
  capability.

## 3C. Connect algebra to evidence

The Ridge results in Section 2B provide direct evidence for the
conditioning analysis in Section 3A. At the selected tuning parameter,

$$\lambda_{\min}=0.28468$$

the condition number decreased from $\kappa_2(G)=16.97$ to $\kappa_2(G+\lambda I)=7.49$
representing a reduction of more than 50%. This confirms that adding
$\lambda I$ increased the smallest eigenvalues of the Gram matrix,
making the linear system substantially better conditioned.

The improved conditioning is reflected in the fitted Ridge coefficients.
Compared with OLS, the coefficient of `lcavol` was shrunk from
**0.6740** to **0.4587**, while the coefficient of the correlated
predictor `lcp` changed from **−0.0924** to **0.0824**. Rather than
allowing these correlated predictors to compete for unstable
coefficients, Ridge shrinks them simultaneously and distributes their
shared contribution more evenly. This behaviour is also visible in the
smooth coefficient paths, where all coefficients move continuously
towards zero as the penalty increases, but none becomes exactly zero.

Overall, the numerical evidence supports the theoretical result of
Section 3A: shifting the eigenvalues by $\lambda$ improves the
conditioning of the Gram matrix, leading to more stable coefficient
estimates while retaining all predictors.

# Problem 4. Elastic net and a neural-feature bridge

## 4A. Research question and source map

**Research Question.**

How does $L_2$ regularization reduce model complexity in neural
networks, and under what conditions is it mathematically equivalent to
weight decay during training?

The evidence in the source map answers the research question by showing
that integrating an $L_2$ penalty into a neural network's objective
function restricts model capacity by discouraging large parameter values
and favoring lower-complexity, minimum-norm solutions (**C5-C7**). This
reduces overfitting and generally improves generalization by preventing
the model from relying excessively on any individual weight.

The source
map also clarifies how this theoretical penalty translates into the
algorithmic process of weight decay during optimization. For standard
stochastic gradient descent, weight decay is mathematically equivalent
to minimizing an objective with an $L_2$ penalty after appropriate
learning-rate rescaling (**C1**). However, this equivalence breaks down
for adaptive optimizers such as Adam because parameter-specific gradient
scaling changes the effective regularization applied to each weight
(**C2**). Although an equivalent weighted $L_2$ interpretation exists
under a fixed preconditioner (**C3**), practical implementations instead
use decoupled weight decay (AdamW), which empirically improves
generalization and simplifies hyperparameter tuning (**C4**).

Overall, the literature shows that $L_2$ regularization provides the theoretical
mechanism for controlling model complexity, while weight decay is the
optimization procedure whose relationship to that penalty depends on the
choice of optimization algorithm.

### Source Map

[@loshchilov2019]

[@krogh1991]

```{r source-map-table, echo=FALSE}
source_map <- data.frame(
  Claim = c(
    "For SGD, weight decay is equivalent to $L_2$ regularization after learning-rate rescaling.",
    "For adaptive optimizers, weight decay is not equivalent to $L_2$ regularization.",
    "With a fixed preconditioner, weight decay corresponds to a weighted $L_2$ regularizer.",
    "AdamW improves generalization and simplifies hyperparameter tuning.",
    "Weight decay selects the minimum-norm solution, improving generalization in linear networks.",
    "Weight decay reduces overfitting to noisy targets.",
    "Weight decay can be extended to non-linear networks via local linearization.",
    "Weight decay lowers generalization and prediction error on NetTalk."
  ),
  `Exact Source and Locator` = c(
    "Loshchilov & Hutter (2019), Section 2, Proposition 1",
    "Loshchilov & Hutter (2019), Section 2, Proposition 2",
    "Loshchilov & Hutter (2019), Section 2, Proposition 3",
    "Loshchilov & Hutter (2019), Sections 4.2–4.4, Figures 2–4",
    "Krogh & Hertz (1991), Section 3, Eqs. 6–10",
    "Krogh & Hertz (1991), Section 4, Eqs. 12–17",
    "Krogh & Hertz (1991), Section 5",
    "Krogh & Hertz (1991), Section 6"
  ),
  `What the Source Establishes` = c(
    "Proves the mathematical equivalence between SGD weight decay and $L_2$ regularization.",
    "Proves adaptive gradient scaling changes the effect of $L_2$ regularization, motivating AdamW.",
    "Derives the equivalent weighted regularizer under fixed preconditioning.",
    "Shows AdamW achieves lower test error and easier hyperparameter tuning than Adam.",
    "Proves weight decay removes irrelevant weight components and favors the minimum-norm solution.",
    "Derives an optimal weight decay for noisy linear models and shows reduced generalization error.",
    "Argues that weight decay favors lower-complexity solutions in non-linear networks.",
    "Demonstrates improved performance with weight decay on the NetTalk benchmark."
  ),
  `What it Does Not Establish` = c(
    "Does not extend this equivalence to adaptive optimizers.",
    "Does not prove AdamW always performs better.",
    "Does not apply directly to practical Adam with a changing preconditioner.",
    "Does not guarantee improvements across all models or datasets.",
    "Does not prove the result for general non-linear networks.",
    "Does not show the same optimum applies to deep neural networks.",
    "Does not provide a rigorous proof for arbitrary non-linear networks.",
    "Does not establish the same gains for modern architectures such as CNNs or Transformers."
  ),
  check.names = FALSE
)

knitr::kable(source_map)
```

## 4B. Elastic net on original predictors

Elastic Net combines the $L_1$ penalty of Lasso and the $L_2$ penalty of
Ridge. The objective is to determine whether a mixture of the two
penalties can achieve a better balance than either method alone.

```{r original-elastic-net, eval=TRUE}
alpha_grid <- c(0.10, 0.25, 0.50, 0.75, 0.90)
enet_models <- vector("list", length(alpha_grid))

enet_summary <- data.frame(
  alpha = alpha_grid,
  lambda.min = NA_real_,
  lambda.1se = NA_real_,
  cvm.min = NA_real_,
  cvm.1se = NA_real_,
  nonzero.min = NA_integer_,
  nonzero.1se = NA_integer_
)

for (i in seq_along(alpha_grid)) {
  fit <- fit_cv_glmnet(
    x = x_train, y = y_train,
    alpha = alpha_grid[i], foldid = foldid
  )
  enet_models[[i]] <- fit

  enet_summary$lambda.min[i] <- fit$lambda.min
  enet_summary$lambda.1se[i] <- fit$lambda.1se
  enet_summary$cvm.min[i] <- fit$cvm[which(fit$lambda == fit$lambda.min)]
  enet_summary$cvm.1se[i] <- fit$cvm[which(fit$lambda == fit$lambda.1se)]
  enet_summary$nonzero.min[i] <- count_nonzero(fit, s = "lambda.min")
  enet_summary$nonzero.1se[i] <- count_nonzero(fit, s = "lambda.1se")
}
knitr::kable(enet_summary, digits = 4)

best_id <- which.min(enet_summary$cvm.min)
best_alpha <- enet_summary$alpha[best_id]
best_model <- enet_models[[best_id]]

cat("Selected alpha :", best_alpha, "\n")
cat("Selected lambda:", best_model$lambda.min, "\n")
cat("Minimum CV MSE :", min(best_model$cvm), "\n")

coef(best_model, s = "lambda.min")

plot(best_model$glmnet.fit, xvar = "lambda", label = TRUE)
title(paste("Elastic Net Paths (alpha =", best_alpha, ")"), line = 2.5)
abline(v = -log(best_model$lambda.min), col = "green", lty = 2, lwd = 1.5)
abline(v = -log(best_model$lambda.1se), col = "blue", lty = 2, lwd = 1.5)
legend("topleft",
  legend = c("lambda.min", "lambda.1se"),
  col = c("green", "blue"), lty = 2, lwd = 1.5, bg = "white"
)
```

### Selection rule

For each candidate value $\alpha$, an Elastic Net model was fitted
using the common five-fold cross-validation assignment. For
each model, the tuning parameter $\lambda$ was selected by `lambda.min`.
The final value of $\alpha$ was then chosen as the one producing the
smallest minimum cross-validation error across the candidate grid.

The results are summarized above. As $\alpha$ increases from 0.10 to
0.90, the minimum cross-validation error decreases gradually from
**0.6363** to **0.6323**, while the number of active predictors at
`lambda.min` decreases from **7** to **4**. Consequently, $\alpha=0.90$
was selected because it achieved the smallest cross-validation error
(**CV MSE = 0.6323**) with the sparsest model among the candidate
values.

At the selected tuning parameters $\alpha = 0.90, \lambda_{\min}=0.1111$, the fitted model retained four predictors `lcavol`, `lweight`, `svi`, and `pgg45`. The remaining four predictors (`age`, `lbph`, `lcp`, and `gleason`) were
shrunk exactly to zero.

### Relationship with Ridge and Lasso in Problem 2

Elastic Net combines the $L_1$ penalty of Lasso with the $L_2$ penalty
of Ridge, producing a solution between the two methods.

In Problem 2, Ridge retained all **eight predictors** because the $L_2$
penalty shrinks coefficients continuously without setting them to zero,
whereas Lasso shrinks several coefficients exactly to zero.

For the prostate dataset, the selected Elastic Net model uses
$\alpha=0.90$ with $\lambda_{\min}=0.1111$ and achieves the lowest
cross-validation error (**CV MSE = 0.6323**).

Thus, the fitted model is much closer to Lasso in terms of variable
selection, but the remaining **10%** $L_2$ penalty ($1-\alpha=0.10$)
still provides additional coefficient stabilization for correlated
predictors. Consequently, Elastic Net combines the sparsity of Lasso
with part of the stability and grouping behaviour of Ridge.

## 4C. Fixed ReLU features

Using the standardized training predictors, we constructed $M = 200$
fixed random ReLU features of the form

$$h_m(x)=\max\{a_m^\top x+c_m,0\},
\qquad m=1,\ldots,M$$

where $a_m\sim N\left(0,\frac1pI_p\right)$ and $c_m\sim N(0,0.5^2)$$

The random weights and offsets were generated once using the declared
seed and then applied unchanged to both the training and test sets.
Hidden features with zero variance in the training data were removed
using training information only.

```{r random-relu-features, eval=TRUE}
set.seed(seeds$features)

relu <- function(z) pmax(z, 0)
M <- 200L
p <- ncol(x_train)

# generate random hidden-layer parameters
A_matrix <- matrix(rnorm(p * M, mean = 0, sd = sqrt(1 / p)),
  nrow = p, ncol = M
)
c_bias <- rnorm(M, mean = 0, sd = 0.5)

# construct ReLU features
H_train_raw <- relu(sweep(x_train %*% A_matrix, 2, c_bias, "+"))
H_test_raw <- relu(sweep(x_test %*% A_matrix, 2, c_bias, "+"))

# remove constant hidden features using training only
keep_hidden <- apply(H_train_raw, 2, sd) > 0
H_train_raw <- H_train_raw[, keep_hidden, drop = FALSE]
H_test_raw <- H_test_raw[, keep_hidden, drop = FALSE]

cat("Hidden features retained:", ncol(H_train_raw), "out of", M, "\n")

# standardize hidden features using training statistics
h_prep <- fit_scaler(H_train_raw)
H_train <- apply_scaler(H_train_raw, h_prep)
H_test <- apply_scaler(H_test_raw, h_prep)
```

Ridge, Lasso, and Elastic Net were fitted on the fixed ReLU feature
matrix using the shared five-fold cross-validation assignment. For
Elastic Net, the mixing parameter $\alpha$ was selected from the same
candidate grid as in Part 4B.

```{r relu-models, eval=TRUE}
ridge_relu <- fit_cv_glmnet(H_train, y_train, alpha = 0, foldid = foldid)
lasso_relu <- fit_cv_glmnet(H_train, y_train, alpha = 1, foldid = foldid)

alpha_grid <- c(0.10, 0.25, 0.50, 0.75, 0.90)
enet_models <- lapply(
  alpha_grid,
  function(alpha) {
    fit_cv_glmnet(H_train, y_train, alpha = alpha, foldid = foldid)
  }
)

cv_errors <- sapply(enet_models, function(model) min(model$cvm))
best_id <- which.min(cv_errors)
best_alpha_relu <- alpha_grid[best_id]
enet_relu <- enet_models[[best_id]]
```

The selected tuning values, training cross-validation errors, and
numbers of active hidden features are summarized below.

```{r relu-summary, eval=TRUE}
relu_summary <- data.frame(
  Model = c("Ridge", "Lasso", "Elastic net"),
  Alpha = c(0, 1, best_alpha_relu),
  Lambda = c(
    ridge_relu$lambda.min,
    lasso_relu$lambda.min,
    enet_relu$lambda.min
  ),
  CV_Error = c(min(ridge_relu$cvm), min(lasso_relu$cvm), min(enet_relu$cvm)),
  Active_Hidden = c(
    count_nonzero(ridge_relu),
    count_nonzero(lasso_relu),
    count_nonzero(enet_relu)
  )
)

knitr::kable(relu_summary,
  digits = 4,
  caption = "Cross-validation results on the fixed ReLU features."
)
```

### Fixed and learned parameters

The hidden-layer weight vectors $a_m$ and offsets $c_m$ were fixed after
being generated once from their specified normal distributions. These
random feature parameters were reused unchanged for both the training
and test sets and were not estimated from the data.

The learned quantities were the regression coefficients associated with
the hidden features and the intercept. Cross-validation selected $\lambda$ for
Ridge and Lasso, and selected both $\alpha$ and $\lambda$ for Elastic Net.

Among the three models, Ridge achieved the smallest training
cross-validation error (0.6980) but retained almost every hidden feature
(198), producing the least sparse model. Lasso selected only 11 active
hidden features, giving the sparsest representation but also the largest
cross-validation error (0.7719). Elastic net selected $\alpha=0.10$,
retained 46 hidden features, and achieved an intermediate
cross-validation error (0.7323), providing a compromise between
prediction accuracy and sparsity.

## 4D. Locked analysis and final holdout evaluation

### Locked analysis declaration

Before inspecting the holdout responses (`y_test`), the analysis was
locked as follows.

- Predictors were standardized using the training statistics only.
- The same five-fold cross-validation assignment was used for all
  regularized models.
- Fixed ReLU features were generated once using `r seeds$features` with
  `M = 200` hidden features and reused unchanged.
- Candidate models
  - OLS
  - Ridge (`lambda.min`)
  - Lasso (`lambda.min`)
  - Elastic Net (`alpha = 0.90`, `lambda.min`)
  - Ridge + ReLU (`lambda.min`)
  - Lasso + ReLU (`lambda.min`)
  - Elastic Net + ReLU (`alpha = 0.10`, `lambda.min`)
- The nominated deployment model was **Elastic Net on the original
  predictors**.
- Holdout performance was evaluated using RMSE and MAE.

```{r final-holdout, eval=TRUE}
pred_ols <- predict(ols_model, newdata = as.data.frame(x_test))
pred_ridge <- as.numeric(predict(ridge_model, newx = x_test, s = "lambda.min"))
pred_lasso <- as.numeric(predict(lasso_model, newx = x_test, s = "lambda.min"))
pred_enet <- as.numeric(predict(best_model, newx = x_test, s = "lambda.min"))

pred_ridge_relu <- as.numeric(
  predict(ridge_relu, newx = H_test, s = "lambda.min")
)

pred_lasso_relu <- as.numeric(
  predict(lasso_relu, newx = H_test, s = "lambda.min")
)

pred_enet_relu <- as.numeric(
  predict(enet_relu, newx = H_test, s = "lambda.min")
)

holdout_results <- data.frame(
  Model = c(
    "OLS", "Ridge", "Lasso", "Elastic Net", "Ridge + ReLU",
    "Lasso + ReLU", "Elastic Net + ReLU"
  ),
  RMSE = c(
    score_regression(y_test, pred_ols)["RMSE"],
    score_regression(y_test, pred_ridge)["RMSE"],
    score_regression(y_test, pred_lasso)["RMSE"],
    score_regression(y_test, pred_enet)["RMSE"],
    score_regression(y_test, pred_ridge_relu)["RMSE"],
    score_regression(y_test, pred_lasso_relu)["RMSE"],
    score_regression(y_test, pred_enet_relu)["RMSE"]
  ),
  MAE = c(
    score_regression(y_test, pred_ols)["MAE"],
    score_regression(y_test, pred_ridge)["MAE"],
    score_regression(y_test, pred_lasso)["MAE"],
    score_regression(y_test, pred_enet)["MAE"],
    score_regression(y_test, pred_ridge_relu)["MAE"],
    score_regression(y_test, pred_lasso_relu)["MAE"],
    score_regression(y_test, pred_enet_relu)["MAE"]
  )
)

knitr::kable(holdout_results,
  digits = 4,
  caption = "Holdout performance of all locked candidate models."
)
```

### Holdout evaluation

The holdout evaluation was performed only once after the analysis had
been locked. No hyperparameters were retuned, no predictors were
replaced, and no random features were regenerated after inspecting the
holdout responses.

OLS achieved the smallest holdout RMSE (0.7296), while Elastic Net with
fixed ReLU features achieved the smallest MAE (0.6134). Among the
regularized models, Elastic Net with fixed ReLU features also produced
the smallest RMSE (0.7563), improving on Elastic Net fitted to the
original predictors (0.7676).

The holdout evidence does not fully support the training-based decision.
Elastic Net on the original predictors was nominated using
cross-validation, but OLS performed best on this particular holdout
sample. Since the holdout set contains only 20 observations, this
difference is reasonably attributed to sampling variability. Following
the locked-analysis protocol, the nominated deployment model is not
changed after viewing the holdout results.

## 4E. Deep-learning connection and limitations

### 1. Ridge penalty versus weight decay

In neural networks, weight decay also encourages smaller weights by
shrinking the parameters during optimization. When using standard
gradient descent, weight decay is mathematically equivalent to adding an
$L_2$ penalty to the loss function because every parameter is updated
using the same global learning rate. The shrinkage introduced by weight
decay is therefore identical to the shrinkage produced by the gradient
of the $L_2$ penalty.

However, this equivalence does not always hold for adaptive optimizers
such as Adam. Adam computes an individual adaptive learning rate for
each parameter based on the history of its gradients. As a result, the
gradient contribution from the $L_2$ penalty is also scaled differently
for each weight, so the effective regularization varies across
parameters. In contrast, weight decay shrinks each parameter directly,
independently of the adaptive gradient scaling. Consequently, adding an
$L_2$ penalty to the objective and applying weight decay during
optimization are no longer mathematically identical. Optimizers such as
AdamW decouple weight decay from the gradient update, restoring the
intended behavior of uniform parameter shrinkage.

### 2. Zero output weights versus sparse input-to-hidden weights

Each hidden neuron computes a nonlinear feature from the input, which is
then passed to the output layer. The prediction can be written as

$$\hat{y}=\sum_{m=1}^{M} w_m h_m(x),$$

where $h_m(x)$ is the output of the $m$-th hidden neuron and $w_m$ is
its corresponding output weight.

If Lasso shrinks an output weight $w_m$ to zero, the corresponding
hidden feature contributes nothing to the prediction and is effectively
removed from the scalar predictor.

However, this does **not** create sparse input-to-hidden weights. The
parameters defining the hidden feature, namely its input weights and
biases, remain unchanged. The hidden neuron still computes the same
feature; its output is simply multiplied by zero before reaching the
prediction. Therefore, sparsity is created only in the **output-layer
weights**, not in the **input-to-hidden connections**.

### 3. Dropout versus permanent pruning

Dropout randomly sets a subset of neuron activations to zero during each
training iteration. The neurons that are masked change from one
iteration to the next, so every neuron continues to participate in
learning over the course of training. This randomness discourages
co-adaptation between neurons and generally improves model
generalization.

Permanent pruning is fundamentally different. It removes selected
weights or neurons permanently so that they no longer participate in
either training or prediction. Thus, dropout is a temporary stochastic
regularization technique, whereas pruning permanently reduces the model
architecture.

### 4. Conclusions cannot be generalized to fully trained deep networks

This experiment uses a **frozen hidden layer**, where the random hidden
weights and biases are generated once and remain fixed. Only the output
weights are learned. Consequently, the experiment evaluates
regularization on **fixed random features** rather than on
representations learned jointly with the hidden layers. The results
therefore cannot be directly generalized to fully trained deep neural
networks, where all layers are optimized together.

### Limitations

1.  **Sensitivity to the random-feature seed.** The hidden features are
    generated from a single random initialization. Different seeds may
    produce different feature representations and model performance, so
    the results may depend on the chosen seed.

2.  **Frozen hidden representation.** Since only the output layer is
    trained, the experiment does not evaluate how regularization affects
    feature learning in end-to-end trained neural networks.

3.  **Limited experimental setting.** The conclusions are based on one
    hidden-layer architecture with 200 ReLU features and a single
    dataset. They may not generalize to deeper architectures, different
    activation functions, or other datasets.

# Problem 5. Report quality and reproducibility

## 5A. Reproduction record

To independently reproduce the results in this report, execute the following steps from a clean R session.

### 1. Folder Layout and Setup

Ensure the working directory contains the R Markdown file, the
bibliography file, and the raw CSV datasets inside a local `data/`
subdirectory:

``` text
STAT02-Lab-Project/
├── ASESII_24A02_ProjectExam_Ridge_Lasso_ProblemStatement.pdf
├── data/
    └── prostate.csv
├── README.md
└── templates/
    ├── ASESII_24A02_ProjectExam_SampleSubmission.Rmd
    ├── ASESII_24A02_ProjectExam_SampleSubmission.pdf
    ├── Group06_AI_Log.csv
    └── references.bib
```

### 2. Commands execution

Open a clean R session (RStudio via `Session -> Restart R`), clear
the workspace, and render the document from the console using relative
paths

``` r
# clear workspace and verify clean session
rm(list = ls())
gc()

# Render the report to HTML/PDF using relative paths
rmarkdown::render("templates/ASESII_24A02_ProjectExam_SampleSubmission.Rmd", output_format = "all")
```

### 3. Holdout

- **Where holdout responses first enter:** The holdout response vector
  `y_test` is partitioned during the initial data split in **Section 1B
  (`split-and-scale`)**, but it is strictly quarantined throughout
  all exploratory data analysis (Section 1C), baseline OLS modeling
  (Section 2A), regularized model tuning (Sections 2B–2C), and
  non-linear feature engineering (Section 4C).
- **First computation:** `y_test` enters mathematical
  computation for the first and only time in **Section 4D
  (`final-holdout-evaluation`)**, where it is evaluated against
  fixed candidate predictions to compute out-of-sample Root Mean Squared
  Error (RMSE) and Mean Absolute Error (MAE).

## 5B. Recommendation and limitations

### 1. Final deployment recommendation

Although the OLS model achieved the lowest
holdout RMSE on this particular train/test split, we recommend the
**Elastic Net model** (using the tuning parameters) for practical deployment.

This recommendation reflects the model selection procedure adopted in
this study. Hyperparameters were selected exclusively by
cross-validation on the training data, while the holdout set was
reserved for a single unbiased final evaluation. Because the prostate
cancer predictors exhibit moderate multicollinearity, Elastic Net
provides a good balance between predictive performance, coefficient
stability, and model complexity. Although OLS performed slightly better on this
particular holdout split, Elastic Net is expected to provide more robust
performance across future samples.

### 2. Prediction versus interpretation

The preferred model depends on the objective of the analysis.

- **Prediction.** For predicting future patient outcomes, the selected
  Elastic Net model provides a good compromise between predictive
  accuracy, coefficient stability, and model complexity. Its regression
  coefficients should be interpreted only as conditional associations
  within the fitted model and **not as evidence of causal effects**.

- **Interpretation.** If the primary objective is identifying a compact
  subset of important predictors rather than maximizing predictive
  accuracy, a more strongly regularized Lasso model may be preferred. This produces a simpler and more interpretable model while accepting a small reduction in predictive performance.

### 3. Study limitations

- **Small sample size.** The analysis is based on only 97 observations
  (77 training and 20 holdout samples). Consequently, performance
  estimates are subject to sampling variability, and different data
  splits may produce different model rankings.

- **Limited external validity.** The dataset represents a historical
  clinical cohort from a single institution. External validation on more
  recent and diverse patient populations would be required before
  clinical deployment.

- **Single holdout evaluation.** Model performance was evaluated using
  one fixed holdout partition. Although this preserves an unbiased final
  assessment, it does not quantify the variability of the reported test
  errors across repeated train/test splits.

- **Fixed random feature mapping.** The ReLU feature expansion in
  Section 4 used randomly generated hidden features without
  backpropagation. Therefore, the results evaluate fixed random feature
  representations rather than fully trained neural networks.

# Preflight and session information {.unnumbered}

```{r preflight, eval=TRUE}
# audit: verify every critical pipeline object exists before generating
# final deliverables
required_objects <- c("train_id", "test_id", "x_train", "x_test", "y_train",
  "y_test", "foldid", "ols_model", "ridge_model", "lasso_model")

missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects) > 0) {
  stop("Preflight Check Failed. Missing required pipeline objects: ",
    paste(missing_objects, collapse = ", "))
}

# safeguard: verify absolute zero row overlap between train and test partitions
stopifnot("Data Leakage Alert: Train and test indices overlap!" =
  length(intersect(train_id, test_id)) == 0L)
cat(
  "Preflight audit passed: Folds, objects, and leakage boundaries are
  verified.\n")
```

```{r session-info, eval=TRUE}
# print full R engine, os, and package version details for
# reproducibility auditing
sessionInfo()
```
