---
title: "ASESII 24A02 Project-Based Quiz 1"
subtitle: "Ridge, Lasso, and Regularization for Neural Features"
author: "Group 06: Le Dat Phuc An_24125052-Pham Nguyen Minh Quan_24125041-Tran Canh Anh Tuan_24125048-Vong Chi Van_24125049"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    latex_engine: xelatex
    fig_crop: false
  html_document:
    toc: true
    toc_depth: 3
bibliography: references.bib
link-citations: true
params:
  group_number: 6
fontsize: 11pt
geometry: margin=1in
editor_options: 
  markdown: 
    wrap: 72
---

# Authorship and contribution statement {.unnumbered}

We confirm that every group member can explain the submitted analysis.
AI use is reported in `Group06_AI_Log.csv`, and the group checked every
AI-assisted item used in this report.

| Student | ID | Main contributions | Verification performed |
|---|---:|---|---|
| Le Dat Phuc An | 24125052 | Sections 1, 2A, 2B, 5A, and 5B. | Verified Sections 2D, 3B, and 4C by reproducing results and reviewing the report. |
| Pham Nguyen Minh Quan | 24125041 | Sections 2D, 2E, 3A, 3B, and 4B. | Verified Sections 1, 2A, 5A, and 5B by checking methodology, code, and documentation. |
| Tran Canh Anh Tuan | 24125048 | Sections 2A,2B,3C, 4A, 4C, 4E, 5A. | Verified Sections 2B, 2C, and 4B by reviewing implementation and numerical results. |
| Vong Chi Van | 24125049 | Sections 2A,2B,2C, 4A, 4E. | Verified Sections 2E, 3A, and 3C by checking derivations, code, and final outputs. |

# Abstract {.unnumbered}

*Maximum 150 words: prediction problem, methods, main evidence, and
recommendation.*

# Reproducible setup {.unnumbered}

```{r setup, message=FALSE, warning=FALSE}
required_packages <- c("tidyverse", "glmnet", "broom", "knitr")
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install the required packages before rendering: ",
       paste(missing_packages, collapse = ", "))
}

library(tidyverse)
library(glmnet)
library(broom)
library(knitr)

group_number <- as.integer(params$group_number)
if (length(group_number) != 1L || is.na(group_number) || group_number < 1L) {
  stop("Set params$group_number to your assigned positive group number.")
}

seeds <- list(
  split = 240200L + group_number,
  cv = 240300L + group_number,
  features = 240400L + group_number
)
```





## Shared helper functions {.unnumbered}

```{r helpers}
find_exam_data <- function(filename) {
  input <- knitr::current_input(dir = TRUE)
  input_dir <- if (length(input) && nzchar(input)) dirname(input) else getwd()
  candidates <- c(
    file.path(input_dir, "data", filename),
    file.path(input_dir, "..", "data", filename),
    file.path(getwd(), "data", filename)
  )
  existing <- candidates[file.exists(candidates)]
  if (!length(existing)) stop("Cannot locate ", filename, " by a relative path.")
  hits <- unique(normalizePath(existing, winslash = "/", mustWork = TRUE))
  if (length(hits) > 1L && length(unique(unname(tools::md5sum(hits)))) > 1L) {
    stop("Multiple nonidentical copies of ", filename, " were found.")
  }
  hits[[1L]]
}

split_rows <- function(n, train_prop = 0.80, seed) {
  stopifnot(n >= 10L, train_prop > 0, train_prop < 1)
  set.seed(seed)
  train <- sort(sample.int(n, floor(train_prop * n), replace = FALSE))
  list(train = train, test = setdiff(seq_len(n), train))
}

fit_scaler <- function(x, tol = 1e-12) {
  x <- as.matrix(x)
  center <- colMeans(x)
  centered <- sweep(x, 2L, center, "-")
  scale <- sqrt(colMeans(centered^2))
  keep <- is.finite(scale) & scale > tol
  if (!any(keep)) stop("No nonconstant training predictor remains.")
  list(
    center = center[keep],
    scale = scale[keep],
    keep = keep,
    dropped = colnames(x)[!keep]
  )
}

apply_scaler <- function(x, prep) {
  x <- as.matrix(x)[, prep$keep, drop = FALSE]
  x <- sweep(x, 2L, prep$center, "-")
  sweep(x, 2L, prep$scale, "/")
}

make_foldid <- function(n, k = 5L, seed) {
  if (k < 3L || k > n) stop("Require 3 <= k <= number of training rows.")
  set.seed(seed)
  sample(rep(seq_len(k), length.out = n))
}

fit_cv_glmnet <- function(x, y, alpha, foldid) {
  glmnet::cv.glmnet(
    x = x,
    y = y,
    family = "gaussian",
    alpha = alpha,
    foldid = foldid,
    standardize = FALSE,
    intercept = TRUE,
    type.measure = "mse"
  )
}

score_regression <- function(y, prediction) {
  y <- as.numeric(y)
  prediction <- as.numeric(prediction)
  if (length(y) != length(prediction) || any(!is.finite(c(y, prediction)))) {
    stop("Metrics require equal-length finite vectors.")
  }
  c(RMSE = sqrt(mean((y - prediction)^2)),
    MAE = mean(abs(y - prediction)))
}

count_nonzero <- function(fit, s = "lambda.min", tol = 1e-8) {
  b <- as.matrix(stats::coef(fit, s = s))
  slopes <- b[rownames(b) != "(Intercept)", 1L]
  sum(abs(slopes) > tol)
}

safe_condition_numbers <- function(x, lambda, tol = 1e-10) {
  G <- crossprod(x) / nrow(x)
  eigenvalues <- eigen(G, symmetric = TRUE, only.values = TRUE)$values
  largest <- max(eigenvalues)
  smallest <- max(min(eigenvalues), 0)
  cutoff <- tol * max(1, largest)
  c(
    gram = if (smallest <= cutoff) Inf else largest / smallest,
    ridge = (largest + lambda) / (smallest + lambda)
  )
}
plot_model_result <- function(model, model_name) { 
   model_result <- augment(model)
  ggplot(model_result, aes(x = .fitted, y = .resid)) +
    geom_point(color = "steelblue", size = 2, alpha = 0.8) +
    
    geom_hline(yintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
    
    theme_minimal() +

    labs(
      title = paste("Residual Plot Model:", model_name),
      subtitle = "Checking for non-linear patterns and constant variance",
      x = "Predicted lpsa (.fitted)",
      y = "Residuals (.resid)"
    ) 
}

vif_manual <- function(model) {
  X <- model.matrix(model)[, -1, drop = FALSE]

  sapply(seq_len(ncol(X)), function(i) {
    fit <- lm(X[, i] ~ X[, -i])
    1 / (1 - summary(fit)$r.squared)
  }) |> setNames(colnames(X))
}

coef_norm <- function(coefs) {
  b <- as.matrix(coefs)
  sqrt(sum(b[-1, ]^2))
}

nonzero_slopes <- function(coefs, tol = 1e-8) {
  b <- as.matrix(coefs)
  sum(abs(b[-1, ]) > tol)
}
```





# Problem 1. Prediction design and leakage control

## 1A. Target and predictors

```{r data-configuration, eval=TRUE}
config <- if (group_number %% 2L == 1L) {
  list(
    file = "fat.csv",
    response = "brozek",
    excluded = c("brozek", "siri", "density", "free")
  )
} else {
  list(
    file = "prostate.csv",
    response = "lpsa",
    excluded = "lpsa"
  )
}

data_raw <- read.csv(find_exam_data(config$file), check.names = FALSE)
predictors <- setdiff(names(data_raw), config$excluded)
analysis_data <- data_raw[, c(config$response, predictors), drop = FALSE]

if (anyNA(analysis_data)) stop("Declare a training-only missing-value plan.")
if (!all(vapply(analysis_data, is.numeric, logical(1)))) {
  stop("The assigned response and predictors must be numeric.")
}
```





*Define the prediction task and justify every exclusion.*

### 1. Response = $lpsa$

$lpsa$ measures the natural Logarithm of Prostate-Specific Antigen,
which is a continuous number

### 2. Candidate Predictors

|                      |                                  |                     |
|---------------------|-------------------------------|--------------------|
| Candidate Predictors | Full Name                        | Type                |
| lcavol               | Log Cancer Volume                | Continuous          |
| lweight              | Log Prostate Weight              | Continuous          |
| age                  | Age                              | Continuous          |
| lbph                 | Log Benign Prostatic Hyperplasia | Continuous          |
| svi                  | Seminal Vesicle Invasion         | Binary (0 / 1)      |
| lcp                  | Log Capsular Penetration         | Continuous          |
| gleason              | Gleason Score                    | Ordered Categorical |
| pgg45                | Percentage Gleason Grade 4/5     | Continuous (0–100)  |

### 3. Unit of Observation

Each row of the dataset contains the information of a radical
prostactomy treated patient which comprises of the target and the
predictors values.

### 4. Study Population

The dataset contains 97 patients aged 41 to 79 with clinically localized
adenocarcinoma who went through radical prostactomy at Stanford
University Center by Dr. Thomas A. Stamey and his surgical team during
the late 1980s.

### 5. Prediction Objective

To build a regression model that predicts the log prostate-specific
antigen level (lpsa) using the clinical variables (lcavol, lweight, age,
lbph, svi, lcp, gleason, and pgg45).

### 6. Dataset Source

<https://sci-hub.sg/storage/2024/6429/4cbb75646440af614abced23a70ba059/stamey1989.pdf>

### 7. Required Exclusions

The response variable, `lpsa`, was excluded from the predictor matrix
because it is the outcome to be predicted. The remaining eight clinical
variables were retained as predictors. Since the dataset contains no
variables derived from `lpsa`, no additional exclusions were required.
This prevents target leakage and ensures a fair comparison among the
regression models.

## 1B. Split and train-only preprocessing

```{r split-and-scale, eval=TRUE}
split <- split_rows(nrow(analysis_data), seed = seeds$split)
train_id <- split$train
test_id <- split$test

x_raw <- as.matrix(analysis_data[, predictors, drop = FALSE])
y_raw <- analysis_data[[config$response]]

x_prep <- fit_scaler(x_raw[train_id, , drop = FALSE])
x_train <- apply_scaler(x_raw[train_id, , drop = FALSE], x_prep)
x_test <- apply_scaler(x_raw[test_id, , drop = FALSE], x_prep)
y_train <- y_raw[train_id]
y_test <- y_raw[test_id]

foldid <- make_foldid(nrow(x_train), k = 5L, seed = seeds$cv)
```





*Report seeds, dimensions, scaling, folds, and leakage safeguards.*

### 1. Split seed

- Seed `split` for 80/20 split: 240206
- Seed `cv` for K-fold split: 240306

### 2. Train/Test Split

The dataset of 97 observations was partitioned into:

- **Training Set:** 77 observations
- **Test Set:** 20 observations

### 3. Feature Scaling Rule

Continuous predictor variables were centered and scaled using the helper
functions `fit_scaler()` and `apply_scaler()`. The scaling parameters
were estimated **only from the training data** in `fit_scaler()` and
then applied unchanged to both the training and test sets by
`apply_scaler()`.

Specifically, each predictor was transformed as $$x_{\text{scaled}} =
\frac{x - \bar{x}_{\text{train}}}
{\sqrt{\frac{1}{n_{\text{train}}} \sum_{i=1}^{n_{\text{train}}} (x_i - \bar{x}_{\text{train}})^2}},$$
where $\bar{x}_{\text{train}}$ is the mean of the predictor in the
training set, and the denominator is the population standard deviation
computed from the training observations. Using the same scaling
parameters for the test set prevents information leakage from the
holdout data.

### 4. Five-Fold Assignment Rule

A single, immutable five-fold cross-validation assignment vector
(`foldid`) was generated on the 77 training observations using the
designated `cv` random seed. To ensure fair and controlled comparisons,
this exact same fold assignment was reused across all models evaluated:

- Ridge Regression
- Lasso Regression
- Elastic Net
- All subsequent hyperparameter grid evaluations

### 5. Missing-Value Plan

No missing values were detected. If any had been present, they would
have been handled using a training-only imputation procedure before
model fitting.

## 1C. Training-data exploration

The exploratory analysis in this section uses only the training
observations to avoid information leakage from the holdout test set.

### Response distribution

```{r response-distribution, eval=FALSE}
summary(y_train)
sd(y_train)

hist(
  y_train,
  main = "Histogram of y_train",
  xlab = "y_train"
)

qqnorm(y_train)
qqline(y_train)
```





The response variable (`lpsa`) has a mean of **2.54** and a median of
**2.59**, indicating an approximately symmetric distribution. Most
observations lie between **1.73** (first quartile) and **3.28** (third
quartile), with a standard deviation of **1.15**.

The histogram suggests that most observations are concentrated around
the center, while only a few observations appear in the lower and upper
tails. The normal Q–Q plot shows that most observations follow the
reference line reasonably well, with slight departures in the tails.
Overall, the response distribution is approximately normal, making it
suitable for linear regression.

### Relationship exploration

#### Scatterplot matrix

```{r scatter-matrix, eval=FALSE}
pairs(
  analysis_data[train_id, ],
  pch = 20,
  cex = 0.6
)
```





The scatterplot matrix provides a visual overview of pairwise
relationships among the response and predictors. Several predictors
exhibit approximately linear relationships with the response,
particularly `lcavol`. Strong positive associations are also visible
between several predictor pairs, suggesting that some variables may
contain overlapping information.

#### Correlation matrix

```{r correlation-heatmap, eval=FALSE}
cor_matrix <- cor(analysis_data[train_id, ])

heatmap_df <- expand.grid(
  Var1 = rownames(cor_matrix),
  Var2 = colnames(cor_matrix)
)

heatmap_df$value <- c(cor_matrix)

ggplot(heatmap_df, aes(Var1, Var2, fill = value)) +
  geom_tile() +
  geom_text(aes(label = round(value, 2)), size = 3) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  coord_fixed() +
  theme_minimal()
```





The correlation heatmap shows that **lcavol** has the strongest linear
relationship with the response (*r* = 0.75), indicating that tumor
volume is the single most informative predictor of `lpsa`. Moderate
positive correlations are also observed for **lcp** (*r* = 0.59) and
**svi** (*r* = 0.58), whereas `age`, `lbph`, and `lweight` exhibit
relatively weak marginal correlations with the response.

### Potential multicollinearity

The correlation matrix indicates several moderately to strongly
correlated predictor pairs.

The strongest pairwise correlations include

- `gleason` and `pgg45` (**r = 0.76**),
- `lcavol` and `lcp` (**r = 0.73**),
- `lcp` and `svi` (**r = 0.67**),
- `lcavol` and `svi` (**r = 0.59**),
- `lcavol` and `gleason` (**r = 0.47**).

These relationships suggest that some predictors may contain overlapping
information. However, pairwise correlation alone does not imply severe
multicollinearity, since multicollinearity depends on the joint linear
relationship among all predictors rather than individual pairs.
Therefore, formal diagnostics such as the variance inflation factor
(VIF) will be performed after fitting the OLS model to determine whether
coefficient instability is present.

### One possible anomaly or influential observation

```{r boxplot, eval=FALSE}
boxplot(analysis_data[train_id, ])
```





The boxplots identify several potential outliers, particularly in
**age** and **pgg45**. Among the predictors, `pgg45` exhibits the
largest variability, while `age` contains several observations outside
the whiskers. These observations are retained because there is no
evidence that they result from data entry errors.

### Question for regularized models

Based on the exploratory analysis, the following question will guide the
remaining analysis:

> Can regularization improve predictive performance and coefficient
> stability when predictors contain overlapping information?

*Interpret multicollinearity, one anomaly, and the question
regularization will address.*

# Problem 2. OLS, ridge, and lasso

## 2A. OLS baseline

The ordinary least squares (OLS) model was fitted using only the
standardized training observations. Throughout this section, the holdout
test set remained completely untouched to ensure a fair comparison with
the regularized regression models.

### Fit OLS

```{r ols-fit, eval=TRUE}
x_train_df <- as.data.frame(x_train)
x_train_df$lpsa <- y_train

ols_model <- lm(lpsa ~ ., data = x_train_df)
```





The fitted OLS model serves as the baseline model for all subsequent
comparisons with ridge and lasso regression.

### Coefficient estimates

```{r ols-summary, eval=FALSE}
summary(ols_model)
```





The fitted OLS model is statistically significant overall (**F =
15.78**, p \< 0.001), indicating that at least one predictor contributes
to explaining the variation in the response variable.

The model achieves a **Multiple** $R^2$ of 0.6500 and an **Adjusted**
$R^2$ of 0.6088, suggesting that approximately **65%** of the
variability in `lpsa` is explained by the eight clinical predictors.

Among the predictors,

- **lcavol** ($\beta=0.674$, p \< 0.001),
- **lweight** ($\beta=0.202$, p = 0.0438),
- **age** ($\beta=-0.194$, p = 0.0435), and
- **svi** ($\beta=0.282$, p = 0.0206)

are statistically significant at the 5% significance level.

The remaining predictors (`lbph`, `lcp`, `gleason`, and `pgg45`) are not
individually significant after adjusting for the other predictors in the
model.

The estimated coefficient for **age** is negative ($\beta=-0.194$). This
result suggests that the estimated effect of `age` changes after
accounting for the remaining predictors in the model. The possibility of
correlation among predictors is therefore investigated using variance
inflation factors in the following subsection.

### Training error

```{r ols-training-error, eval=FALSE}
train_pred <- predict(ols_model)

train_mse <- mean((y_train - train_pred)^2)
train_rmse <- sqrt(train_mse)

cat("Training MSE :", round(train_mse, 4), "\n")
cat("Training RMSE:", round(train_rmse, 4), "\n")
```





The fitted OLS model achieved

- **Training MSE = 0.4544**
- **Training RMSE = 0.6741**

These quantities measure the goodness-of-fit on the training
observations used to estimate the regression coefficients. Since the
model is evaluated on the same data used for fitting, the training error
is expected to underestimate the prediction error on unseen
observations.

### Five-fold cross-validation

```{r ols-cv, eval=TRUE}
cv_mse <- numeric(5)
coefs_list <-list()
for(i in 1:5){

  train_fold <- x_train_df[foldid != i, ]
  valid_fold <- x_train_df[foldid == i, ]

  model <- lm(lpsa ~ ., data = train_fold)

  pred <- predict(model, newdata = valid_fold)
  fold_coef <- as.matrix(coef(model, s="lambda.min"))
  coefs_list[[i]] <- as.data.frame(t(fold_coef))
  cv_mse[i] <- mean((valid_fold$lpsa - pred)^2)
}
coef_ols_df <- bind_rows(coefs_list)
cv_table <- data.frame(
  Fold = 1:5,
  Validation_MSE = round(cv_mse, 4)
)

knitr::kable(cv_table)

cat("Mean CV MSE :", round(mean(cv_mse), 4), "\n")
cat("Mean CV RMSE:", round(sqrt(mean(cv_mse)), 4), "\n")
```





The validation errors obtained from the five folds are

| Fold | Validation MSE |
|-----:|---------------:|
|    1 |         0.8259 |
|    2 |         0.8568 |
|    3 |         0.4011 |
|    4 |         0.4893 |
|    5 |         0.9002 |

The average cross-validation error is

- **Mean CV MSE = 0.6947**
- **Mean CV RMSE = 0.8335**

As expected, the cross-validation error is larger than the training
error because each validation fold consists of observations that were
not used to estimate the regression coefficients. Although some
variation exists among the five folds, the validation errors remain
within a similar range, suggesting that the predictive performance of
the OLS model is reasonably stable across different training subsets.

### Residual diagnostics

#### Residual versus fitted values

```{r ols-residual, eval=FALSE}
plot(
  fitted(ols_model),
  resid(ols_model),
  xlab = "Fitted values",
  ylab = "Residuals",
  main = "Residuals vs Fitted"
)

abline(h = 0, col = "red")
```





The residual-versus-fitted plot shows that the residuals are randomly
scattered around zero without noticeable curvature or funnel-shaped
patterns. This suggests that the assumptions of

- linearity, and
- constant error variance (homoscedasticity)

are reasonably satisfied. No obvious systematic pattern indicating model
misspecification is observed.

#### Normal Q-Q plot

```{r ols-qqplot, eval=FALSE}
qqnorm(resid(ols_model))
qqline(resid(ols_model), col = "red")
```





The normal Q-Q plot indicates that most residuals lie close to the
reference line, with only slight departures in the two tails. Therefore,
the residual distribution appears approximately normal.

The normal Q–Q plot indicates that the residuals are approximately
normally distributed, with only slight deviations in the tails. This
behavior is consistent with the mildly skewed response distribution
observed during the exploratory analysis. Since the residuals satisfy

$$\hat{\mathbf e}=(I-H)\mathbf y,$$

where $\hat{\mathbf e}$ is the residual vector, $\mathbf y$ is the
response vector, $I$ is the identity matrix, and
$H=X(X^\top X)^{-1}X^\top$ is the hat matrix, their distribution
naturally reflects the response after removing the fitted linear
component. Overall, there is no strong evidence against the normality
assumption.

#### Influence plot

```{r ols-influence, eval=FALSE}
plot(ols_model, which = 5)
```





Most observations have low leverage and moderate residuals. Although a
few observations (e.g., 23, 37, and 54) stand out slightly, none appears
to have an unusually large influence on the fitted model.

#### Cook's Distance

```{r ols-cook-distance, eval=FALSE}
cook <- cooks.distance(ols_model)

plot(
  cook,
  type = "h",
  lwd = 3,
  col = "red",
  xlab = "Observation",
  ylab = "Cook's Distance",
  main = "Cook's Distance"
)

abline(
  h = 4 / nrow(x_train_df),
  col = "blue",
  lty = 2
)
```





Several observations exceed the reference line $4/n$, but all Cook's
distance values are well below 1. Therefore, there is no strong evidence
of influential observations.

Taken together, the residual diagnostics do not reveal any serious
violation of the standard assumptions underlying ordinary least squares
regression.

### Ill-conditioned design

```{r ols-vif, eval=FALSE}
vif_manual(ols_model)
```





The variance inflation factors (VIFs) were computed to assess
multicollinearity among the predictors.

The VIF values range from **1.33** to **3.24**, with the largest values
observed for

- `lcp` (3.24),
- `pgg45` (2.97),
- `gleason` (2.68), and
- `lcavol` (2.42).

None of the predictors exceeds the commonly used threshold of **5**,
indicating that severe multicollinearity is not present in the training
design matrix.

This finding is consistent with the correlation analysis presented in
Section 1C. Although several predictor pairs, such as `gleason`–`pgg45`
and `lcavol`–`lcp`, exhibit moderate pairwise correlations, pairwise
correlation alone does not necessarily imply problematic
multicollinearity. The VIF measures how well each predictor can be
explained by **all remaining predictors simultaneously**, making it a
more appropriate diagnostic for coefficient instability.

Overall, the VIF results suggest that the correlations among predictors
are not sufficiently strong to produce substantial inflation of the
coefficient variances. Therefore, no clear evidence of an
ill-conditioned design was found, and the OLS estimates provide an
appropriate baseline for comparison with ridge and lasso regression in
the following sections.

## 2B. Ridge

### Fitting Ridge

The Ridge model is fitted using `glmnet` with `alpha = 0` on the
training data.

```{r ridge}
ridge_model <- fit_cv_glmnet(
  x_train,
  y_train,
  alpha = 0,
  foldid = foldid
)

plot(ridge_model)

abline(v = -log(ridge_model$lambda.min),
       col = "green",
       lty = 2,
       lwd = 1.5)

abline(v = -log(ridge_model$lambda.1se),
       col = "blue",
       lty = 2,
       lwd = 1.5)

y_pos <- max(ridge_model$cvup)

text(-log(ridge_model$lambda.min),
     y_pos,
     "lambda.min",
     pos = 2,
     col = "green",
     cex = 0.9)

text(-log(ridge_model$lambda.1se),
     y_pos,
     "lambda.1se",
     pos = 2,
     col = "blue",
     cex = 0.9)
```





The cross-validation curve shows how the prediction error changes with
the penalty parameter. The mean cross-validation error decreases until
reaching its minimum at **`lambda.min`**. Around the minimum, the error
changes only slightly, indicating that a range of penalty values
provides similar predictive performance.

Because Ridge regression uses an L2 penalty, every coefficient is
continuously shrunk toward zero as the penalty increases. However, Ridge
does not apply coordinate-wise thresholding, so no coefficient is set
exactly to zero and all eight predictors remain in the model.

### Investigation on $\lambda_{min}$ and $\lambda_{1se}$

The optimal penalty values selected by cross-validation are reported
below.

```{r}
cat(sprintf("lambda.min: %.5f\n", ridge_model$lambda.min))
cat(sprintf("lambda.1se: %.5f\n\n", ridge_model$lambda.1se))
```





`lambda.min` is the value of $\lambda$ that minimizes the mean
cross-validation error. It is selected when predictive accuracy is the
primary objective.

`lambda.1se` is the largest value of $\lambda$ whose cross-validation
error is within one standard error of the minimum. It applies stronger
regularization, producing a simpler and more stable model while usually
maintaining similar predictive performance.

The coefficient estimates and corresponding cross-validation error at
`lambda.min` are shown below.

```{r}
min_id <- which(ridge_model$lambda == ridge_model$lambda.min)

cat(sprintf("cvm.min: %.5f\n", ridge_model$cvm[min_id]))

coef(ridge_model, s = "lambda.min")
```





At `lambda.min`, the model achieves the lowest cross-validation error.
The coefficients are moderately shrunk compared with the OLS estimates,
while the most important predictors still retain relatively large
coefficients.

The corresponding results for `lambda.1se` are shown below.

```{r}
se1_id <- which(ridge_model$lambda == ridge_model$lambda.1se)

cat(sprintf("cvm.1se: %.5f\n\n", ridge_model$cvm[se1_id]))

coef(ridge_model, s = "lambda.1se")
```





Using the larger penalty `lambda.1se` shrinks every coefficient further
toward zero than `lambda.min`. The cross-validation error increases only
slightly from , indicating that a more regularized model can achieve
nearly the same predictive performance.

Unlike Lasso, Ridge regression continuously shrinks the coefficients but
does not set any coefficient exactly to zero. Therefore, every predictor
remains in the fitted model under both tuning parameter choices.

### Coefficient paths

The coefficient paths show how the Ridge coefficients change as the
penalty varies.

```{r}

plot(ridge_model$glmnet.fit,
     xvar = "lambda",
     label = TRUE)

title("Ridge Coefficient Paths", line = 2.5)

abline(v = -log(ridge_model$lambda.min),
       col = "green",
       lty = 2,
       lwd = 1.5)

abline(v = -log(ridge_model$lambda.1se),
       col = "blue",
       lty = 2,
       lwd = 1.5)

legend("topleft",
       legend = c("lambda.min", "lambda.1se"),
       col = c("green", "blue"),
       lty = 2,
       lwd = 1.5,
       cex = 0.8,
       bg = "white")

predictor_names <- rownames(ridge_model$glmnet.fit$beta)

legend("topleft",
       inset = c(0, 0.21),
       legend = paste(1:length(predictor_names),
                      predictor_names,
                      sep = " - "),
       col = 1:length(predictor_names),
       lty = 1,
       lwd = 1.5,
       cex = 0.8,
       bg = "white",
       ncol = 2)
```





As the penalty decreases (moving from left to right), the coefficients
move smoothly away from near zero toward their OLS estimates. Larger
penalties produce stronger shrinkage, whereas smaller penalties allow
the coefficients to approach the unregularized solution.

Unlike Lasso, Ridge continuously shrinks every coefficient and therefore
does not produce exact zero coefficients.

### General Behavior of Correlated Predictors Under Ridge Regression

Ridge regression is particularly effective when predictors are highly
correlated. In ordinary least squares, correlated predictors compete to
explain the same variation in the response, leading to unstable
coefficient estimates with high variance.

Ridge addresses this issue by shrinking the coefficients of correlated
predictors towards each other rather than allowing one predictor to
dominate the others. This reduces the variability of the coefficient
estimates and improves model stability while retaining all predictors in
the model. This behaviour is reflected in the coefficient paths, where
all coefficients are smoothly shrunk towards zero as the penalty
increases.

Unlike Lasso, Ridge regression does not perform variable selection.
Instead, all coefficients are continuously shrunk towards zero, but none
becomes exactly zero.

#### Examples from the Data

The effect of Ridge can be seen in several pairs of highly correlated
predictors.

- **`lcavol` and `lcp` (correlation = 0.73):** Under OLS, `lcavol`
  receives a much larger positive coefficient while `lcp` is assigned a
  small negative coefficient. Ridge shrinks both coefficients, reducing
  this imbalance and allowing the two correlated predictors to share
  their predictive contribution.

- **`pgg45` and `gleason` (correlation = 0.76):** OLS assigns
  substantially different coefficients to these correlated predictors.
  Ridge shrinks both estimates towards zero, producing more stable
  coefficient estimates while retaining both variables in the model.

## 2C. Lasso

### Fitting Lasso

The lasso model is fitted using `glmnet` by setting `alpha = 1` on the
training data.

```{r lasso, eval=TRUE}
lasso_model <- fit_cv_glmnet(
  x_train,
  y_train,
  alpha = 1,
  foldid = foldid
)

plot(lasso_model)

abline(v = -log(lasso_model$lambda.min),
       col = "green", lty = 2, lwd = 1.5)

abline(v = -log(lasso_model$lambda.1se),
       col = "blue", lty = 2, lwd = 1.5)

y_pos <- max(lasso_model$cvup)

text(-log(lasso_model$lambda.min), y_pos,
     labels = "lambda.min",
     pos = 2, col = "green", cex = 0.9)

text(-log(lasso_model$lambda.1se), y_pos,
     labels = "lambda.1se",
     pos = 2, col = "blue", cex = 0.9)
```





Unlike Ridge regression, Lasso performs both coefficient shrinkage and
variable selection. As the penalty increases, some coefficients are
shrunk exactly to zero, producing a simpler and more interpretable
model. The cross-validation curve identifies the penalty giving the
lowest prediction error (`lambda.min`) together with the largest penalty
within one standard error (`lambda.1se`).

### Investigation on $\lambda_{min}$ and $\lambda_{1se}$

The optimal penalties are

```{r}
cat(sprintf("lambda.min: %.5f\n", lasso_model$lambda.min))
cat(sprintf("lambda.1se: %.5f\n\n", lasso_model$lambda.1se))
```





The corresponding training cross-validation errors are

```{r}
min_id <- which(lasso_model$lambda == lasso_model$lambda.min)
se1_id <- which(lasso_model$lambda == lasso_model$lambda.1se)

cat(sprintf("cvm.min : %.5f\n",
            lasso_model$cvm[min_id]))
cat(sprintf("cvm.1se : %.5f\n\n",
            lasso_model$cvm[se1_id]))
```





The selected predictors under the two tuning rules are

```{r}
cat("--- Nonzero coefficients (lambda.min) ---\n")
coef_min <- coef(lasso_model, s = "lambda.min")
print(coef_min[coef_min[, 1] != 0, ]) 

cat("\n--- Nonzero coefficients (lambda.1se) ---\n")
coef_1se <- coef(lasso_model, s = "lambda.1se")
print(coef_1se[coef_1se[, 1] != 0, ])
```





`lambda.min` prioritizes predictive accuracy and therefore usually keeps
more predictors. `lambda.1se` accepts a small increase in prediction
error in exchange for a sparser and easier-to-interpret model.

### Coefficient paths

The coefficient paths illustrate how predictors enter and leave the
model as the penalty changes.

```{r}
plot(lasso_model$glmnet.fit,
     xvar = "lambda",
     label = TRUE)

title("Lasso Coefficient Paths", line = 2.5)

abline(v = -log(lasso_model$lambda.min),
       col = "green",
       lty = 2,
       lwd = 1.5)

abline(v = -log(lasso_model$lambda.1se),
       col = "blue",
       lty = 2,
       lwd = 1.5)

legend("topleft", 
       legend = c("lambda.min", "lambda.1se"), 
       col = c("green", "blue"), 
       lty = 2, 
       lwd = 1.5, 
       cex = 0.75,
       bg = "white")
```





Unlike Ridge regression, Lasso coefficient paths frequently terminate at
exactly zero. Variables that reach zero are removed from the model,
while only the strongest predictors remain active under larger penalty
values.

### Application-specific interpretation

Lasso identifies the predictors that provide the strongest independent
contribution to explaining `lpsa` after accounting for the remaining
variables.

At `lambda.min`, four predictors remain active: `lcavol`, `lweight`,
`svi`, and `pgg45`. Among these, `lcavol` has by far the largest
coefficient (0.607), indicating that tumor volume is the strongest
predictor of prostate-specific antigen levels. `svi` also remains,
suggesting that seminal vesicle invasion contributes additional
predictive information beyond tumor volume. `lweight` and `pgg45` have
smaller coefficients but still improve predictive performance enough to
be retained.

At `lambda.1se`, the stronger penalty removes both `lweight` and
`pgg45`, leaving only `lcavol` and `svi`. This indicates that these two
variables contain the most robust independent signal in the data. The
cross-validation error increases only slightly (0.632 to 0.728), while
the model is reduced from four predictors to two, yielding a much
simpler and more interpretable model.

The variables excluded by Lasso (`age`, `lbph`, `lcp`, `gleason`,
`lweight`, and `pgg45` under `lambda.1se`) are not necessarily unrelated
to prostate cancer. Rather, after accounting for `lcavol` and `svi`,
their remaining predictive contribution is too small to justify the
additional model complexity. Some of their information is likely shared
with the retained predictors because several clinical measurements are
moderately correlated.

## 2D. Training-based comparison and locked core model

This section compares the candidate models using only the training data
and cross-validation results. The holdout test set is not used so that
model selection remains unbiased.

### Training-based comparison

```{r core-comparison}
# OLS
cv_ols <- mean(cv_mse)
coef_ols <- coef(ols_model)
norm_ols <- coef_norm(coef_ols)
nz_ols <- nonzero_slopes(coef_ols)

# Ridge
coef_ridge_min <- coef(ridge_model, s = "lambda.min")
coef_ridge_1se <- coef(ridge_model, s = "lambda.1se")

id_min <- which.min(abs(ridge_model$lambda - ridge_model$lambda.min))
id_1se <- which.min(abs(ridge_model$lambda - ridge_model$lambda.1se))

cv_ridge_min <- ridge_model$cvm[id_min]
cv_ridge_1se <- ridge_model$cvm[id_1se]

norm_ridge_min <- coef_norm(coef_ridge_min)
norm_ridge_1se <- coef_norm(coef_ridge_1se)

nz_ridge_min <- nonzero_slopes(coef_ridge_min)
nz_ridge_1se <- nonzero_slopes(coef_ridge_1se)

# Lasso
coef_lasso_min <- coef(lasso_model, s = "lambda.min")
coef_lasso_1se <- coef(lasso_model, s = "lambda.1se")

id_min <- which.min(abs(lasso_model$lambda - lasso_model$lambda.min))
id_1se <- which.min(abs(lasso_model$lambda - lasso_model$lambda.1se))

cv_lasso_min <- lasso_model$cvm[id_min]
cv_lasso_1se <- lasso_model$cvm[id_1se]

norm_lasso_min <- coef_norm(coef_lasso_min)
norm_lasso_1se <- coef_norm(coef_lasso_1se)

nz_lasso_min <- nonzero_slopes(coef_lasso_min)
nz_lasso_1se <- nonzero_slopes(coef_lasso_1se)

comparison_table <- data.frame(
  Model = c(
    "OLS",
    "Ridge",
    "Ridge",
    "Lasso",
    "Lasso"
  ),
  Tuning = c(
    "None",
    "lambda.min",
    "lambda.1se",
    "lambda.min",
    "lambda.1se"
  ),
  CV_MSE = c(
    cv_ols,
    cv_ridge_min,
    cv_ridge_1se,
    cv_lasso_min,
    cv_lasso_1se
  ),
  Coefficient_Norm = c(
    norm_ols,
    norm_ridge_min,
    norm_ridge_1se,
    norm_lasso_min,
    norm_lasso_1se
  ),
  Nonzero_Slopes = c(
    nz_ols,
    nz_ridge_min,
    nz_ridge_1se,
    nz_lasso_min,
    nz_lasso_1se
  )
)

knitr::kable(comparison_table, digits = 4)
```





Regularization improves predictive performance over OLS, with both Ridge
(`lambda.min`) and Lasso (`lambda.min`) achieving lower cross-validation
errors. Ridge shrinks the coefficients but retains all eight predictors,
whereas Lasso also performs variable selection. Lasso (`lambda.min`)
achieves the lowest cross-validation error while retaining only four
predictors, providing the best balance between prediction accuracy and
model simplicity.

### Core model nominated before holdout evaluation

Based on the training results, **Lasso with `lambda.min`** is selected
as the core candidate. It achieves the lowest cross-validation error
while reducing the model from eight predictors to four. Although Ridge
(`lambda.min`) has similar predictive performance, it retains all
predictors and is therefore less interpretable.

### Why `lambda.min` and `lambda.1se` answer different decision questions

`lambda.min` prioritizes predictive performance by minimizing the
cross-validation error. In contrast, `lambda.1se` selects a larger
penalty to obtain a simpler and more stable model while accepting a
small increase in prediction error.

### Predictive performance versus interpretability

Predictive performance is assessed by the cross-validation error,
whereas interpretability depends on model simplicity. In this analysis,
Lasso (`lambda.min`) achieves the lowest prediction error while
retaining only four predictors, making it the preferred model before the
final holdout evaluation.

## 2E. Perturbation or stability check

This stability check investigates how Ridge and Lasso respond to severe
multicollinearity by duplicating a predictor with a small amount of
random noise. The experiment uses only the training data.

### Prediction before running the check

Before running the perturbation, we expect Ridge to distribute the
coefficient between the original and duplicated predictors because the
ridge penalty encourages correlated predictors to share their effects.
In contrast, Lasso is expected to retain only one of the two predictors
while shrinking the other coefficient to zero.

```{r perturbation}
# Duplicate a strong predictor with small random noise
target_idx <- 1
target_name <- colnames(x_train)[target_idx]

duplicate_var <- x_train[, target_idx] +
  rnorm(nrow(x_train), sd = 0.01)

x_train_pert <- cbind(
  x_train,
  duplicate_var = duplicate_var
)

# Refit Ridge and Lasso using the perturbed data
ridge_pert <- fit_cv_glmnet(
  x_train_pert,
  y_train,
  alpha = 0,
  foldid = foldid
)

lasso_pert <- fit_cv_glmnet(
  x_train_pert,
  y_train,
  alpha = 1,
  foldid = foldid
)

ridge_coef <- coef(ridge_pert, s = "lambda.1se")
lasso_coef <- coef(lasso_pert, s = "lambda.1se")

cat("Ridge:\n")
print(ridge_coef[c(target_name, "duplicate_var"), , drop = FALSE])

cat("\nLasso:\n")
print(lasso_coef[c(target_name, "duplicate_var"), , drop = FALSE])
```





### Observed result and explanation

The perturbation confirms the expected behaviour of the two
regularization methods.

Ridge assigns almost identical coefficients to the original predictor
(`lcavol` = 0.1560) and its duplicate (0.1561), indicating that the
effect is shared between highly correlated predictors. This behaviour
improves coefficient stability in the presence of multicollinearity.

In contrast, Lasso retains the original predictor (`lcavol` = 0.4918)
while shrinking the duplicated predictor exactly to zero. This
demonstrates that Lasso performs variable selection by choosing one
representative predictor and discarding redundant correlated variables.

Overall, the perturbation illustrates the different mechanisms of Ridge
and Lasso. Ridge stabilizes coefficient estimates by sharing information
across correlated predictors, whereas Lasso produces a sparse and more
interpretable model by selecting only one predictor from a correlated
group.

# Problem 3. Mathematical mechanisms

## 3A. Ridge and conditioning

*Write the ridge derivation. Define dimensions and explain uniqueness.*

Let
$X \in \mathbb{R}^{n \times p}, y \in \mathbb{R}^n, b \in \mathbb{R}^p$
and define $$G = \frac{1}{n} X^\top X, \quad
z = \frac{1}{n} X^\top y$$ consequently, $G \in \mathbb{R}^{p \times p}$
and $z \in \mathbb{R}^p$.

From the objective function
$$Q_\lambda(b) = \frac{1}{2n}\|y - Xb\|_2^2 + \frac{\lambda}{2}\|b\|_2^2$$

We expand Ridge objective function $$\begin{align}
  Q_\lambda (b) &= \frac{1}{2n} \|y - X b\|_2^2 + \frac{\lambda}{2} \|b\|_2^2, \quad (\lambda > 0) \\
  &= \frac{1}{2n} (y - X b)^\top (y - X b) + \frac{\lambda}{2} b^\top b \\
  &= \frac{1}{2n} (y^\top y - y^\top X b - b^\top X^\top y + b^\top X^\top X b) + \frac{\lambda}{2} b^\top b \\
  &= \frac{1}{2n} (y^\top y - 2 y^\top X b + b^\top X^\top X b) + \frac{\lambda}{2} b^\top b \\
\end{align}$$

We take the gradient with respect to $b$ and substitute $$\begin{align}
  \nabla_b Q_\lambda (b) &= 0 - \frac{1}{n} y^\top X + \frac{1}{n} X^\top X b + \lambda b \\
  &= - z + G b + \lambda b
\end{align}$$

Set $\nabla_b Q_\lambda (b) = 0 \Leftrightarrow (G + \lambda I_p) b = z$
which is the $\textbf{normal equation}$.

Because $G = \frac{1}{n}X^\top X$ is a positive semi-definite Gram
matrix, all its eigenvalues are $\ge 0$. Adding a penalty $\lambda > 0$
along the diagonal (i.e., $\lambda I_p$) shifts all eigenvalues up by
$\lambda$. Thus, the matrix $(G + \lambda I_p)$ is strictly positive
definite (all eigenvalues $\ge \lambda > 0$), making it full rank and
invertible. This guarantees a unique minimizer

$$\hat{b}_\lambda^{\text{ridge}} = (G + \lambda I_p)^{-1} z$$

```{r ridge-conditioning, eval=FALSE}
# Use x_train and the fitted ridge lambda.min.
# safe_condition_numbers(x_train, lambda = ...)

condition_numbers <- safe_condition_numbers(x_train, ridge_model$lambda.min)
kappa_G <- condition_numbers["gram"]
kappa_ridge <- condition_numbers["ridge"]

cat("Condition number of G (kappa_G):", kappa_G, "\n")
cat("Condition number of Ridge (kappa_ridge):", kappa_ridge, "\n")
```





### Numerical interpretation

The OLS estimator is

$$\hat b=(X^\top X)^{-1}X^\top y,$$

so its numerical stability depends on the Gram matrix

$$G=X^\top X.$$

When predictors are highly correlated, the columns of $X$ become nearly
linearly dependent, making

$$\det(G)\approx0.$$

Equivalently,

$$G=Q\Lambda Q^\top,
\qquad
\Lambda=\operatorname{diag}(\mu_1,\ldots,\mu_p),$$

contains one or more very small eigenvalues. Since

$$G^{-1}
=
Q\Lambda^{-1}Q^\top,
\qquad
\Lambda^{-1}
=
\operatorname{diag}
\left(
\frac1{\mu_1},
\ldots,
\frac1{\mu_p}
\right),$$

a small eigenvalue

$$\mu_{\min}\approx0$$

produces a very large reciprocal

$$\frac{1}{\mu_{\min}}
\rightarrow
\infty.$$

Therefore, the inverse $(X^\top X)^{-1}$ amplifies small perturbations
in the data, making the coefficient estimate

$$\hat b=(X^\top X)^{-1}X^\top y$$

highly sensitive to sampling variation and numerical error.

This instability is summarized by the spectral condition number

$$\kappa_2(G)
=
\frac{\mu_{\max}}
{\mu_{\min}},$$

where a large value indicates an ill-conditioned matrix.

Ridge replaces $G$ by

$$G+\lambda I,$$

which shifts every eigenvalue by

$$\mu_i
\longrightarrow
\mu_i+\lambda.$$

Hence,

$$\kappa_2(G+\lambda I)
=
\frac{\mu_{\max}+\lambda}
{\mu_{\min}+\lambda}.$$

Since the smallest eigenvalue increases the most in relative terms, the
condition number decreases, making the matrix inversion and coefficient
estimates substantially more stable.

### Geometric interpretation

The least-squares objective is

$$RSS(b)=\|y-Xb\|^2,$$

whose Hessian is

$$\nabla^2RSS(b)=2X^\top X=2G.$$

Thus, the eigenvalues of $G$ determine the curvature of the loss
surface. When one eigenvalue is much smaller than the others, the
curvature in that direction is very weak, producing long, narrow
elliptical contours. Along this nearly flat valley, many coefficient
vectors produce almost identical values of $RSS(b)$, so a small
perturbation in the data can move the optimum a large distance, leading
to unstable coefficients.

Ridge adds the penalty

$$\lambda\|b\|_2^2,$$

whose Hessian is

$$2\lambda I.$$

Therefore, the penalized Hessian becomes

$$2(G+\lambda I),$$

increasing the curvature in every direction, especially those
corresponding to the smallest eigenvalues. The elongated ellipses become
more circular, the optimum is better localized, and the coefficient
estimates become more stable while retaining all predictors.

## 3B. Lasso optimality and soft thresholding

*Derive the zero/nonzero optimality conditions and the
orthonormal-design formula.*

Let
$X \in \mathbb{R}^{n \times p}, y \in \mathbb{R}^n, b \in \mathbb{R}^p$
and define $$G = \frac{1}{n} X^\top X = I_p, \quad
z = \frac{1}{n} X^\top y$$ consequently, $G \in \mathbb{R}^{p \times p}$
and $z \in \mathbb{R}^p$.

The objective function
$$R_\lambda (b) = \frac{1}{2n} \|y - X b\|_2^2 + \lambda \|b\|_1, \quad \|b\|_1 = \sum_{j=1}^p |b_j|, \lambda > 0$$

We expand Lasso objective function $$\begin{aligned}
R_\lambda (b) &= \frac{1}{2n} \|y - X b\|_2^2 + \lambda \|b\|_1 \\
              &= \frac{1}{2n} (y - X b)^\top (y - X b) + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} (y^\top y - y^\top X b - b^\top X^\top y + b^\top X^\top X b) + \lambda \sum_{j=1}^p |b_j|
\end{aligned}$$

The term $y^\top X b$ is is a scalar, therefore
$(y^\top X b)^\top = b^\top X^\top (y^\top)^\top = b^\top X^\top y$.

Then the objective becomes $$\begin{aligned}
R_\lambda (b) &= \frac{1}{2n} y^\top y - \frac{1}{n} b^\top X^\top y + \frac{1}{2n} b^\top X^\top X b + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} y^\top y - b^\top z + \frac{1}{2} b^\top b + \lambda \sum_{j=1}^p |b_j| \\
              &= \frac{1}{2n} y^\top y + \sum_{j=1}^p \left( - b_j z_j + \frac{1}{2} b_j^2 + \lambda |b_j| \right)
\end{aligned}$$

The term $\frac{1}{2n} y^\top y$ is a constant with respect to the
variable $b$. , leaving a sum of $p$ independent 1-dimensional
optimization problems.

Consider a fixed $j$, and let
$l_j (t) = \frac{1}{2} t^2 - t z_j + \lambda |t|$, for
$t \in \mathbb{R}$. We analyze this in three cases because $|t|$ is
*non-differentiable* at $t=0$.

- **Case 1:** If $t > 0$, then $|t| = t$ and
  $l_j (t) = \frac{1}{2} t^2 - t z_j + \lambda t$. Thus,
  $l_j' (t) = t - z_j + \lambda$.
  $$ l_j' (t) = 0 \iff t = z_j - \lambda \quad (1) $$ Moreover, since we
  assumed $t > 0$, equation $(1)$ is a valid minimum if and only if
  $z_j > \lambda$.

- **Case 2:** If $t < 0$, then $|t| = - t$ and
  $l_j (t) = \frac{1}{2} t^2 - t z_j - \lambda t$. Thus,
  $l_j' (t) = t - z_j - \lambda$.
  $$ l_j' (t) = 0 \iff t = z_j + \lambda \quad (2) $$ Moreover, since we
  assumed $t < 0$, equation $(2)$ is a valid minimum if and only if
  $z_j < -\lambda$.

- **Case 3:** If $|z_j| \le \lambda$, there is no valid solution where
  the derivative equals zero (neither Case 1 nor Case 2). Because the
  function is strictly convex, the global minimum must exist, which
  means it is trapped exactly at the non-differentiable corner: $t = 0$.
  We verify that $t = 0$ minimizes $l_j (t)$ (3).

For any $t \in \mathbb{R}$ $$\begin{aligned}
l_j (t) - l_j (0) &= \frac{1}{2} t^2 - t z_j + \lambda |t| \\
                  &\ge \frac{1}{2} t^2 - |t| |z_j| + \lambda |t| \quad (-t z_j \ge -|t| |z_j|) \\
                  &\ge \frac{1}{2} t^2 - |t| (|z_j| - \lambda) \\
                  &\ge 0 \quad (|z_j| - \lambda \le 0)
\end{aligned}$$ Therefore, $l_j (t) \ge l_j (0)$ for all
$t \in \mathbb{R}$, proving $t=0$ is the global minimum for this range.

Combining these cases yields the soft-thresholding estimator for
$\hat{b}_j$: $$\hat{b}_j = \begin{cases}
  z_j - \lambda & \text{if } z_j > \lambda, \\
  0             & \text{if } |z_j| \le \lambda, \\
  z_j + \lambda & \text{if } z_j < -\lambda
\end{cases}$$

The uniqueness is guaranteed because the objective function is strictly
convex. Under the orthogonal design, the variables are perfectly
independent, forming a strictly convex "bowl" shape. Even with the sharp
point added by the $L_1$ penalty, the strict convexity mathematically
ensures there is exactly one absolute global minimum.

**Soft-thresholding operator under orthonormal design** Assume that
$\frac{1}{n} X^T X = I_p$. Let $z = \frac{1}{n} X^T y$. Then the Lasso
minimizer of $R_\lambda (b)$ is *unique* and has coordinates
$$\hat{b}_j = S_\lambda (z_j), \quad \text{where the soft-thresholding function is } S_\lambda (z) = \text{sgn}(z) (|z| - \lambda)_+$$
where $(u)_+ = \max(u, 0)$.

**Why lasso can produce exact zeros while ridge generally does not.**
Under $\frac{1}{n} X^T X = I_p$, we can compare the closed-form
estimators

- **Ridge:** $\hat{b}_j^{\text{Ridge}} = (1 +  \lambda)^{-1} z_j$
  (proportional shrinkage). It divides the coefficient by a constant.
  The coefficients get extremely small, but they never become zero. The
  optimality condition means $b_j$ only equals exactly zero if $z_j$ is
  exactly zero (which is highly improbable with real data).
- **Lasso:**
  $\hat{b}_j^{\text{Lasso}} = \text{sgn}(z_j) (|z_j| - \lambda)_+$
  (absolute shrinkage). It subtracts a constant from the magnitude. If a
  coefficient is originally small, the penalty forces it to exactly
  zero. Case 3 shows that Lasso maps a whole range of values to exactly
  zero, creating sparsity. This gives Lasso its **feature selection**
  capability.

## 3C. Connect algebra to evidence

The ridge results in Section 2B provide direct evidence for the
conditioning analysis in Section 3A. At the selected tuning parameter,

$$\lambda_{\min}=0.28468,$$

the condition number decreased from

$$\kappa_2(G)=16.97$$

to

$$\kappa_2(G+\lambda I)=7.49,$$

representing a reduction of more than 50%. This confirms that adding
$\lambda I$ increased the smallest eigenvalues of the Gram matrix,
making the linear system substantially better conditioned.

The improved conditioning is reflected in the fitted ridge coefficients.
Compared with OLS, the coefficient of `lcavol` was shrunk from
**0.6740** to **0.4587**, while the coefficient of the correlated
predictor `lcp` changed from **−0.0924** to **0.0824**. Rather than
allowing these correlated predictors to compete for unstable
coefficients, Ridge shrinks them simultaneously and distributes their
shared contribution more evenly. This behaviour is also visible in the
smooth coefficient paths, where all coefficients move continuously
towards zero as the penalty increases, but none becomes exactly zero.

Overall, the numerical evidence supports the theoretical result of
Section 3A: shifting the eigenvalues by $\lambda$ improves the
conditioning of the Gram matrix, leading to more stable coefficient
estimates while retaining all predictors.

# Problem 4. Elastic net and a neural-feature bridge

## 4A. Research question and source map

**Research Question:**

*How does the methodological mechanism of* $L_2$ regularization - which
stabilizes models by damping parameter variance along low-eigenvalue
directions - diverge from algorithmic weight decay when training deep
neural networks with adaptive gradient optimizers?

**Context & Methodological Bridge**

In classical linear models and neural networks trained with standard
Stochastic Gradient Descent (SGD), adding an $L_2$ penalty to the loss
function and applying algorithmic weight decay are mathematically
identical operations. When minimizing a loss function
$\mathcal{L}(w) + \frac{\lambda}{2} \Vert{}w\Vert{}_2^2$, the derivative
of the penalty introduces a $+\lambda w$ term to the gradient. Inside
the standard SGD update rule, this factors into a decay multiplier:
$w_{t+1} = (1 - \alpha\lambda)w_t - \alpha \nabla \mathcal{L}(w_t)$.

However, in modern deep neural networks, models are primarily trained
using adaptive optimizers like Adam. Loshchilov and Hutter (2019)
demonstrated that placing the $L_2$ penalty inside the loss function
causes it to be divided by Adam's adaptive gradient scaling factor. This
neutralizes the penalty for weights with large gradients and
hyper-inflates it for sparse gradients. To restore the classical,
eigenvalue-damping properties of $L_2$ regularization in deep learning,
weight decay must be mathematically decoupled from the loss gradient
(AdamW).

**Source Map**

| Claim | Exact Source and Locator | What the Source Establishes | What it does not establish |
|:-----------------|:-----------------|:-----------------|:-----------------|
| **Proposition 1:** For SGD, weight decay is equivalent to $L_2$ regularization after learning-rate rescaling. | Loshchilov & Hutter (2019), Section 2, Proposition 1 | Proves the mathematical equivalence between SGD weight decay and $L_2$ regularization. | Does not extend this equivalence to adaptive optimizers. |
| **Proposition 2:** For adaptive optimizers, weight decay is not equivalent to $L_2$ regularization. | Loshchilov & Hutter (2019), Section 2, Proposition 2 | Proves adaptive gradient scaling changes the effect of $L_2$ regularization, motivating AdamW. | Does not prove AdamW always performs better. |
| **Proposition 3:** With a fixed preconditioner, weight decay corresponds to a weighted $L_2$ regularizer. | Loshchilov & Hutter (2019), Section 2, Proposition 3 | Derives the equivalent weighted regularizer under fixed preconditioning. | Does not apply directly to practical Adam with a changing preconditioner. |
| **Empirical:** AdamW improves generalization and simplifies hyperparameter tuning. | Loshchilov & Hutter (2019), Sections 4.2–4.4, Figures 2–4 | Shows AdamW achieves lower test error and easier hyperparameter tuning than Adam. | Does not guarantee improvements across all models or datasets. |
| **Theoretical:** Weight decay selects the minimum-norm solution, improving generalization in linear networks. | Krogh & Hertz (1991), Section 3, Eqs. (6)–(10) | Proves weight decay removes irrelevant weight components and favors the minimum-norm solution. | Does not prove the result for general non-linear networks. |
| **Theoretical:** Weight decay reduces overfitting to noisy targets. | Krogh & Hertz (1991), Section 4, Eqs. (12)–(17) | Derives an optimal weight decay for noisy linear models and shows reduced generalization error. | Does not show the same optimum applies to deep neural networks. |
| **Theoretical:** Weight decay can be extended to non-linear networks via local linearization. | Krogh & Hertz (1991), Section 5 | Argues that weight decay favors lower-complexity solutions in non-linear networks. | Does not provide a rigorous proof for arbitrary non-linear networks. |
| **Empirical:** Weight decay lowers generalization and prediction error on NetTalk. | Krogh & Hertz (1991), Section 6, Figure 2 | Demonstrates improved performance with weight decay on the NetTalk benchmark. | Does not establish the same gains for modern architectures such as CNNs or Transformers. |

**References**

- Krogh, A., & Hertz, J. A. (1992). A simple weight decay can improve
  generalization. In *Advances in Neural Information Processing Systems*
  (Vol. 4, pp. 950-957).

- Loshchilov, I., & Hutter, F. (2019). Decoupling weight decay from
  gradient-based optimizers. In *International Conference on Learning
  Representations (ICLR)*.

## 4B. Elastic net on original predictors

Elastic Net combines the $L_1$ penalty of Lasso and the $L_2$ penalty of
Ridge. The objective is to determine whether a mixture of the two
penalties can achieve a better balance between prediction accuracy,
coefficient stability, and model sparsity than either method alone.

```{r original-elastic-net, eval=TRUE}
alpha_grid <- c(0.10, 0.25, 0.50, 0.75, 0.90)

enet_models <- vector("list", length(alpha_grid))

enet_summary <- data.frame(
  alpha = alpha_grid,
  lambda.min = NA_real_,
  lambda.1se = NA_real_,
  cvm.min = NA_real_,
  cvm.1se = NA_real_,
  nonzero.min = NA_integer_,
  nonzero.1se = NA_integer_
)

for(i in seq_along(alpha_grid)){

  fit <- fit_cv_glmnet(
    x = x_train,
    y = y_train,
    alpha = alpha_grid[i],
    foldid = foldid
  )

  enet_models[[i]] <- fit

  id_min <- which.min(abs(fit$lambda - fit$lambda.min))
  id_1se <- which.min(abs(fit$lambda - fit$lambda.1se))

  enet_summary$lambda.min[i] <- fit$lambda.min
  enet_summary$lambda.1se[i] <- fit$lambda.1se

  enet_summary$cvm.min[i] <- fit$cvm[id_min]
  enet_summary$cvm.1se[i] <- fit$cvm[id_1se]

  enet_summary$nonzero.min[i] <-
    count_nonzero(fit, s = "lambda.min")

  enet_summary$nonzero.1se[i] <-
    count_nonzero(fit, s = "lambda.1se")
}

knitr::kable(enet_summary, digits = 4)

best_id <- which.min(enet_summary$cvm.min)

best_alpha <- enet_summary$alpha[best_id]
best_model <- enet_models[[best_id]]

cat("Selected alpha :", best_alpha, "\n")
cat("Selected lambda:", best_model$lambda.min, "\n")
cat("Minimum CV MSE :", min(best_model$cvm), "\n")

coef(best_model, s = "lambda.min")

plot(
  best_model$glmnet.fit,
  xvar = "lambda",
  label = TRUE
)

title(
  paste("Elastic Net Paths (alpha =", best_alpha, ")"),
  line = 2.5
)

abline(
  v = -log(best_model$lambda.min),
  col = "green",
  lty = 2,
  lwd = 1.5
)

abline(
  v = -log(best_model$lambda.1se),
  col = "blue",
  lty = 2,
  lwd = 1.5
)

legend(
  "topleft",
  legend = c("lambda.min", "lambda.1se"),
  col = c("green", "blue"),
  lty = 2,
  lwd = 1.5,
  bg = "white"
)
```





### Selection rule

For each candidate value of $\alpha$, an Elastic Net model was fitted
using the common five-fold cross-validation assignment (`foldid`). For
each model, the tuning parameter $\lambda$ was selected by `lambda.min`,
the value that minimized the mean cross-validation mean squared error.
The final value of $\alpha$ was then chosen as the one producing the
smallest minimum cross-validation error across the candidate grid.

The results are summarized above. As $\alpha$ increases from 0.10 to
0.90, the minimum cross-validation error decreases gradually from
**0.6363** to **0.6323**, while the number of active predictors at
`lambda.min` decreases from **7** to **4**. Consequently, $\alpha=0.90$
was selected because it achieved the smallest cross-validation error
(**CV MSE = 0.6323**) with the sparsest model among the candidate
values.

At the selected tuning parameters,

- $\alpha = 0.90$,
- $\lambda_{\min}=0.1111$,

the fitted model retained four predictors:

- `lcavol`,
- `lweight`,
- `svi`,
- `pgg45`.

The remaining four predictors (`age`, `lbph`, `lcp`, and `gleason`) were
shrunk exactly to zero.

### Relationship with Ridge and Lasso in Problem 2

Elastic Net combines the $L_1$ penalty of Lasso with the $L_2$ penalty
of Ridge, producing a solution between the two methods.

In Problem 2, Ridge retained all **eight predictors** because the $L_2$
penalty shrinks coefficients continuously without setting them to zero,
whereas Lasso selected a sparse model by shrinking several coefficients
exactly to zero.

For the prostate dataset, the selected Elastic Net model uses
$\alpha=0.90$ with $\lambda_{\min}=0.1111$ and achieves the lowest
cross-validation error (**CV MSE = 0.6323**). At this solution, only
four predictors (`lcavol`, `lweight`, `svi`, and `pgg45`) remain active,
while the remaining four coefficients are shrunk exactly to zero.

Thus, the fitted model is much closer to Lasso in terms of variable
selection, but the remaining **10%** $L_2$ penalty ($1-\alpha=0.10$)
still provides additional coefficient stabilization for correlated
predictors. Consequently, Elastic Net combines the sparsity of Lasso
with part of the stability and grouping behaviour of Ridge.

## 4C. Fixed ReLU features

Using the standardized training predictors, we constructed $M = 200$
fixed random ReLU features of the form

$$h_m(x)=\max\{a_m^\top x+c_m,0\},
\qquad m=1,\ldots,M,$$

where

$$a_m\sim N\left(0,\frac1pI_p\right),$$

and

$$c_m\sim N(0,0.5^2).$$

The random weights and offsets were generated once using the declared
seed and then applied unchanged to both the training and test sets.
Hidden features with zero variance in the training data were removed
using training information only.

```{r random-relu-features}
set.seed(seeds$features)

relu <- function(z) pmax(z, 0)

M <- 200L
p <- ncol(x_train)

# Generate random hidden-layer parameters
A_matrix <- matrix(
  rnorm(
    p * M,
    mean = 0,
    sd = sqrt(1 / p)
  ),
  nrow = p,
  ncol = M
)

c_bias <- rnorm(
  M,
  mean = 0,
  sd = 0.5
)

# Construct ReLU features
H_train_raw <- relu(
  sweep(
    x_train %*% A_matrix,
    2,
    c_bias,
    "+"
  )
)

H_test_raw <- relu(
  sweep(
    x_test %*% A_matrix,
    2,
    c_bias,
    "+"
  )
)

# Remove constant hidden features using training only
keep_hidden <- apply(
  H_train_raw,
  2,
  sd
) > 0

H_train_raw <- H_train_raw[
  ,
  keep_hidden,
  drop = FALSE
]

H_test_raw <- H_test_raw[
  ,
  keep_hidden,
  drop = FALSE
]

cat(
  "Hidden features retained:",
  ncol(H_train_raw),
  "out of",
  M,
  "\n"
)

# Standardize hidden features using training statistics
h_prep <- fit_scaler(H_train_raw)

H_train <- apply_scaler(
  H_train_raw,
  h_prep
)

H_test <- apply_scaler(
  H_test_raw,
  h_prep
)
```





Ridge, lasso, and elastic net were fitted on the fixed ReLU feature
matrix using the shared five-fold cross-validation assignment. For
elastic net, the mixing parameter $\alpha$ was selected from the same
candidate grid as in Part 4B.

```{r relu-models}
ridge_relu <- fit_cv_glmnet(
  H_train,
  y_train,
  alpha = 0,
  foldid = foldid
)

lasso_relu <- fit_cv_glmnet(
  H_train,
  y_train,
  alpha = 1,
  foldid = foldid
)

alpha_grid <- c(
  0.10,
  0.25,
  0.50,
  0.75,
  0.90
)

enet_models <- lapply(
  alpha_grid,
  function(alpha)
    fit_cv_glmnet(
      H_train,
      y_train,
      alpha = alpha,
      foldid = foldid
    )
)

cv_errors <- sapply(
  enet_models,
  function(model) min(model$cvm)
)

best_id <- which.min(cv_errors)

best_alpha_relu <- alpha_grid[best_id]

enet_relu <- enet_models[[best_id]]
```





The selected tuning values, training cross-validation errors, and
numbers of active hidden features are summarized below.

```{r relu-summary}
relu_summary <- data.frame(

  Model = c(
    "Ridge",
    "Lasso",
    "Elastic net"
  ),

  Alpha = c(
    0,
    1,
    best_alpha_relu
  ),

  Lambda = c(
    ridge_relu$lambda.min,
    lasso_relu$lambda.min,
    enet_relu$lambda.min
  ),

  CV_Error = c(
    min(ridge_relu$cvm),
    min(lasso_relu$cvm),
    min(enet_relu$cvm)
  ),

  Active_Hidden = c(
    count_nonzero(ridge_relu),
    count_nonzero(lasso_relu),
    count_nonzero(enet_relu)
  )
)

knitr::kable(
  relu_summary,
  digits = 4,
  caption = "Cross-validation results on the fixed ReLU features."
)
```





### Fixed and learned parameters

The hidden-layer weight vectors $a_m$ and offsets $c_m$ were fixed after
being generated once from their specified normal distributions. These
random feature parameters were reused unchanged for both the training
and test sets and were not estimated from the data.

The learned quantities were the regression coefficients associated with
the hidden features and the intercept. Cross-validation selected the
regularization parameter $\lambda$ for ridge and lasso, and selected
both the mixing parameter $\alpha$ and the regularization parameter
$\lambda$ for elastic net.

Among the three models, ridge achieved the smallest training
cross-validation error (0.6980) but retained almost every hidden feature
(198), producing the least sparse model. Lasso selected only 11 active
hidden features, giving the sparsest representation but also the largest
cross-validation error (0.7719). Elastic net selected $\alpha=0.10$,
retained 46 hidden features, and achieved an intermediate
cross-validation error (0.7323), providing a compromise between
prediction accuracy and sparsity.
------------------------------------------------------------------ \##
4D. Locked analysis and final holdout evaluation

### Locked analysis declaration

Before inspecting the holdout responses (`y_test`), the analysis was
locked as follows.

- Predictors were standardized using the training statistics only.
- The same five-fold cross-validation assignment was used for all
  regularized models.
- Fixed ReLU features were generated once using `r seeds$features` with
  `M = 200` hidden features and reused unchanged.
- Candidate models:
  - OLS
  - Ridge (`lambda.min`)
  - Lasso (`lambda.min`)
  - Elastic Net (`alpha = 0.90`, `lambda.min`)
  - Ridge + ReLU (`lambda.min`)
  - Lasso + ReLU (`lambda.min`)
  - Elastic Net + ReLU (`alpha = 0.10`, `lambda.min`)
- The nominated deployment model was **Elastic Net on the original
  predictors**.
- Holdout performance was evaluated using RMSE and MAE.

```{r final-holdout}
## ---------- Original-feature models ----------

pred_ols <- predict(
  ols_model,
  newdata = as.data.frame(x_test)
)

pred_ridge <- as.numeric(
  predict(ridge_model, newx = x_test, s = "lambda.min")
)

pred_lasso <- as.numeric(
  predict(lasso_model, newx = x_test, s = "lambda.min")
)

pred_enet <- as.numeric(
  predict(best_model, newx = x_test, s = "lambda.min")
)

## ---------- ReLU-feature models ----------

pred_ridge_relu <- as.numeric(
  predict(ridge_relu, newx = H_test, s = "lambda.min")
)

pred_lasso_relu <- as.numeric(
  predict(lasso_relu, newx = H_test, s = "lambda.min")
)

pred_enet_relu <- as.numeric(
  predict(enet_relu, newx = H_test, s = "lambda.min")
)

holdout_results <- data.frame(
  Model = c(
    "OLS",
    "Ridge",
    "Lasso",
    "Elastic Net",
    "Ridge + ReLU",
    "Lasso + ReLU",
    "Elastic Net + ReLU"
  ),
  RMSE = c(
    score_regression(y_test, pred_ols)["RMSE"],
    score_regression(y_test, pred_ridge)["RMSE"],
    score_regression(y_test, pred_lasso)["RMSE"],
    score_regression(y_test, pred_enet)["RMSE"],
    score_regression(y_test, pred_ridge_relu)["RMSE"],
    score_regression(y_test, pred_lasso_relu)["RMSE"],
    score_regression(y_test, pred_enet_relu)["RMSE"]
  ),
  MAE = c(
    score_regression(y_test, pred_ols)["MAE"],
    score_regression(y_test, pred_ridge)["MAE"],
    score_regression(y_test, pred_lasso)["MAE"],
    score_regression(y_test, pred_enet)["MAE"],
    score_regression(y_test, pred_ridge_relu)["MAE"],
    score_regression(y_test, pred_lasso_relu)["MAE"],
    score_regression(y_test, pred_enet_relu)["MAE"]
  )
)

knitr::kable(
  holdout_results,
  digits = 4,
  caption = "Holdout performance of all locked candidate models."
)
```





### Holdout evaluation

The holdout evaluation was performed only once after the analysis had
been locked. No hyperparameters were retuned, no predictors were
replaced, and no random features were regenerated after inspecting the
holdout responses.

OLS achieved the smallest holdout RMSE (0.7296), while Elastic Net with
fixed ReLU features achieved the smallest MAE (0.6134). Among the
regularized models, Elastic Net with fixed ReLU features also produced
the smallest RMSE (0.7563), improving on Elastic Net fitted to the
original predictors (0.7676).

The holdout evidence does not fully support the training-based decision.
Elastic Net on the original predictors was nominated using
cross-validation, but OLS performed best on this particular holdout
sample. Since the holdout set contains only 20 observations, this
difference is reasonably attributed to sampling variability. Following
the locked-analysis protocol, the nominated deployment model is not
changed after viewing the holdout results.

## 4E. Deep-learning connection and limitations

## 4E. Discussion

### 1. \(L_2\) penalty versus weight decay

In this experiment, only the **output weights** are optimized while the hidden-layer parameters remain fixed. Applying an \(L_2\) penalty to the output weights is therefore equivalent to **ridge regression**, which shrinks the learned coefficients to reduce overfitting. However, in fully trainable neural networks using adaptive optimizers such as Adam, adding an \(L_2\) penalty to the loss is generally **not equivalent** to weight decay because adaptive gradient scaling changes the effect of the regularization term. AdamW resolves this by decoupling weight decay from the gradient update [@loshchilov2019adamw].

### 2. Zero output weights do not create sparse hidden weights

Each hidden feature is generated from fixed random input-to-hidden weights and remains unchanged throughout training. If an output weight becomes zero, the corresponding hidden feature no longer contributes to the prediction. However, the input-to-hidden weights that generate this feature remain unchanged because they are never optimized. Therefore, sparsity occurs only in the **output layer**, not in the hidden layer or the network structure.

### 3. Dropout is not the same as pruning

Dropout randomly masks a subset of hidden activations during each training iteration, with a new mask sampled for every mini-batch. During inference, all neurons remain active. In contrast, pruning permanently removes weights or neurons from the network. Therefore, dropout is a temporary regularization method, whereas pruning permanently changes the model architecture.

### 4. Conclusions cannot be generalized to fully trained deep networks

This experiment uses a **frozen hidden layer**, where the random hidden weights and biases are generated once and remain fixed. Only the output weights are learned. Consequently, the experiment evaluates regularization on **fixed random features** rather than on representations learned jointly with the hidden layers. The results therefore cannot be directly generalized to fully trained deep neural networks, where all layers are optimized together [@krogh1992weightdecay].

### Limitations

1. **Sensitivity to the random-feature seed.** The hidden features are generated from a single random initialization. Different seeds may produce different feature representations and model performance, so the results may depend on the chosen seed.

2. **Frozen hidden representation.** Since only the output layer is trained, the experiment does not evaluate how regularization affects feature learning in end-to-end trained neural networks.

3. **Limited experimental setting.** The conclusions are based on one hidden-layer architecture with 200 ReLU features and a single dataset. They may not generalize to deeper architectures, different activation functions, or other datasets.

# Abstract {.unnumbered}

This study investigates log prostate-specific antigen (`lpsa`)
prediction in 97 prostatectomy patients using clinical covariates. To
mitigate multicollinearity and prevent target leakage, we implement an
immutable 80/20 train/test partition with training-only preprocessing.
We compare Ordinary Least Squares (OLS) against regularized linear
models (Ridge, Lasso, Elastic Net) and a non-linear Random ReLU feature
representation ($M=200$). Mathematical derivations confirm that $L_2$
regularization stabilizes ill-conditioned Gram matrices by upwardly
shifting eigenvalues, reducing the condition number by over 50%
($\kappa$ dropped from 16.97 to 7.49). Empirical five-fold
cross-validation identifies Lasso (`lambda.min`) and Elastic Net as
superior predictors over baseline OLS and static non-linear projections.
For clinical deployment, we nominate the Elastic Net model on original
predictors, which optimizes out-of-sample generalization while
collaboratively shrinking correlated biomarkers. We emphasize that
regularized coefficients represent conditional predictive slopes rather
than causal biological mechanisms.

# Problem 5. Report quality and reproducibility

## 5A. Reproduction record

To independently reproduce the results, figures, and evaluation tables
in this report, execute the following steps from a clean R session:

### 1. Folder Layout and Setup

Ensure the working directory contains the R Markdown file, the
bibliography file, and the raw CSV datasets inside a local `data/`
subdirectory:

``` text
STAT02_Lab_Project/
├── ASESII_24A02_ProjectExam_Ridge_Lasso_ProblemStatement.pdf
├── data/
    └── prostate.csv
├── README.md     
└── templates/
    ├── ASESII_24A02_ProjectExam_SampleSubmission.Rmd        
    ├── ASESII_24A02_ProjectExam_SampleSubmission.pdf
    ├── Group06_AI_Log.csv       
    └── references.bib        
```

### 2. Execution Commands

Open a clean R session (in RStudio via `Session -> Restart R`), clear
the workspace, and render the document from the console using relative
paths:

``` r
# Clear workspace and verify clean session
rm(list = ls())
gc()

# Verify required packages are installed
required_pkgs <- c("tidyverse", "glmnet", "broom", "knitr", "rmarkdown")
invisible(lapply(required_pkgs, function(pkg) {
  if (!requireNamespace(pkg, quietly = TRUE)) install.packages(pkg)
}))

# Render the report to HTML/PDF using relative paths
rmarkdown::render("templates/ASESII_24A02_ProjectExam_SampleSubmission.Rmd", output_format = "all")
```

### 3. Holdout Quarantine Audit

- **Where holdout responses first enter:** The holdout response vector
  `y_test` is partitioned during the initial data split in **Section 1B
  (`split-and-scale` chunk)**, but it is strictly quarantined throughout
  all exploratory data analysis (Section 1C), baseline OLS modeling
  (Section 2A), regularized model tuning (Sections 2B–2C), and
  non-linear feature engineering (Section 4C).
- **Point of first computation:** `y_test` enters mathematical
  computation for the absolute first and only time in **Section 4D
  (`final-holdout-evaluation` chunk)**, where it is evaluated against
  fixed candidate predictions to compute out-of-sample Root Mean Squared
  Error (RMSE) and Mean Absolute Error (MAE).

## 5B. Recommendation and limitations

### 1. Final deployment recommendation

Although the ordinary least squares (OLS) model achieved the lowest
holdout RMSE on this particular train/test split, we recommend the
**Elastic Net model** (using the tuning parameters selected by
cross-validation) for practical deployment.

This recommendation reflects the model selection procedure adopted in
this study. Hyperparameters were selected exclusively by
cross-validation on the training data, while the holdout set was
reserved for a single unbiased final evaluation. Because the prostate
cancer predictors exhibit moderate multicollinearity, Elastic Net
provides a good balance between predictive performance, coefficient
stability, and model complexity by combining Ridge shrinkage with Lasso
variable selection. Although OLS performed slightly better on this
particular holdout split, Elastic Net is expected to provide more robust
performance across future samples.

### 2. Prediction versus interpretation

The preferred model depends on the objective of the analysis.

- **Prediction.** For predicting future patient outcomes, the selected
  Elastic Net model provides a good compromise between predictive
  accuracy, coefficient stability, and model complexity. Its regression
  coefficients should be interpreted only as conditional associations
  within the fitted model and **not as evidence of causal effects**.

- **Interpretation.** If the primary objective is identifying a compact
  subset of important predictors rather than maximizing predictive
  accuracy, a more strongly regularized Lasso model (for example, using
  `lambda.1se`) may be preferred. This produces a simpler and more
  interpretable model while accepting a small reduction in predictive
  performance.

### 3. Study limitations

- **Small sample size.** The analysis is based on only 97 observations
  (77 training and 20 holdout samples). Consequently, performance
  estimates are subject to sampling variability, and different data
  splits may produce different model rankings.

- **Limited external validity.** The dataset represents a historical
  clinical cohort from a single institution. External validation on more
  recent and diverse patient populations would be required before
  clinical deployment.

- **Single holdout evaluation.** Model performance was evaluated using
  one fixed holdout partition. Although this preserves an unbiased final
  assessment, it does not quantify the variability of the reported test
  errors across repeated train/test splits.

- **Fixed random feature mapping.** The ReLU feature expansion in
  Section 4 used randomly generated hidden features without
  backpropagation. Therefore, the results evaluate fixed random feature
  representations rather than fully trained neural networks.

## 5C. AI verification record

| AI-assisted item | Verification procedure | Evidence | Final decision |
|:----------------|:---------------------|:----------------|:----------------|
| **Ridge regression derivation (Section 3A)** | Re-derived the penalized least squares objective using matrix calculus and verified each differentiation step against standard references. | Confirmed the normal equation $(X^\top X + \lambda I_p)\hat{\beta} = X^\top y$ and the resulting closed-form Ridge estimator. | **Accepted with minor revision:** Improved notation and expanded intermediate algebraic steps for clarity. |
| **Lasso soft-thresholding derivation (Section 3B)** | Verified the coordinate-wise optimization using subgradient conditions under an orthonormal design matrix. | Confirmed that the solution is the soft-thresholding operator and that the derivation is valid only under the stated orthonormal assumption. | **Accepted:** Mathematical derivation was consistent after explicitly stating the required assumptions. |
| **Manual VIF helper function** | Compared the custom implementation with the output of `car::vif()` using the baseline OLS model. | Both methods produced identical variance inflation factors up to numerical precision. | **Accepted:** The custom implementation was retained to avoid introducing an additional package dependency. |
| **Discussion of weight decay and adaptive optimization (Section 4A)** | Checked theoretical claims against the original literature on L2 regularization and decoupled weight decay. | Confirmed that adaptive optimizers such as Adam make L2 regularization and weight decay no longer equivalent, consistent with Loshchilov and Hutter (2019). | **Accepted with revision:** Removed unsupported claims and restricted the discussion to statements directly supported by the cited references. |

# Preflight and session information {.unnumbered}

```{r preflight, eval=FALSE}
# Audit: Verify every critical pipeline object exists before generating final deliverables
required_objects <- c(
  "train_id", "test_id", "x_train", "x_test", "y_train", "y_test", 
  "foldid", "ols_model", "ridge_model", "lasso_model"
)

missing_objects <- setdiff(required_objects, ls())
if (length(missing_objects) > 0) {
  stop(
    "Preflight Check Failed. Missing required pipeline objects: ",
    paste(missing_objects, collapse = ", ")
  )
}

# Safeguard: Verify absolute zero row overlap between train and test partitions
stopifnot("Data Leakage Alert: Train and test indices overlap!" = length(intersect(train_id, test_id)) == 0L)
cat("Preflight audit passed: Folds, objects, and leakage boundaries are verified.\n")
```

```{r session-info}
# Print full R engine, OS, and package version details for reproducibility auditing
sessionInfo()
```